### 2014-08-25 manuscript/AppendixS1.tex
Initial commit

\documentclass[12pt,fleqn]{article}
%\documentclass[12pt,a4paper]{article}
\usepackage{natbib}
\usepackage{lineno}
%\usepackage{lscape}
%\usepackage{rotating}
%\usepackage{rotcapt, rotate}
\usepackage{amsmath,epsfig,epsf,psfrag}
\usepackage{setspace}
\usepackage{ulem}
\usepackage{xcolor}
\usepackage[labelfont=bf,labelsep=period]{caption} %for making figure and table numbers bold
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\hypersetup{pdfpagemode=UseNone}

%\usepackage{a4wide,amsmath,epsfig,epsf,psfrag}

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bX{{\ensuremath\mathbf{X}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\renewcommand{\refname}{\hspace{2.3in} \normalfont \normalsize LITERATURE CITED}
%this tells it to put 'Literature Cited' instead of 'References'
\bibpunct{(}{)}{,}{a}{}{;}
\oddsidemargin 0.0in
\evensidemargin 0.0in
\textwidth 6.5in
\headheight 0.0in
\topmargin 0.0in
\textheight=9.0in
%\renewcommand{\tablename}{\textbf{Table}}
%\renewcommand{\figurename}{\textbf{Figure}}
\renewcommand{\em}{\it}
\renewcommand\thefigure{S1.\arabic{figure}}

\begin{document}

\begin{center} \bf {\large Avoiding extrapolation bias when using statistical models to make ecological prediction}

\vspace{0.7cm}
Paul B. Conn$^{1*}$, Devin S. Johnson$^1$, and Peter L. Boveng$^1$
\end{center}
\vspace{0.5cm}

\rm
\small

\it $^1$National Marine Mammal Laboratory, Alaska Fisheries Science Center,
NOAA National Marine Fisheries Service,
Seattle, Washington 98115 U.S.A.\\

\rm \begin{flushleft}

\raggedbottom
\vspace{.5in}

\begin{center}
Appendix S1: Model formulation and Gibbs sampling algorithms for certain classes and extensions of the generalized linear model
\bigskip
\end{center}
\vspace{.3in}

\doublespacing

We write all statistical models for ecological prediction in the form
\begin{linenomath*}
\begin{equation}
  \label{eq:basic_model}
  Y_i \sim f_Y(g^{-1}(\mu_i)),
\end{equation}
\end{linenomath*}
where $\mu_i=\theta_i + \epsilon_i$, $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function,
$\theta_i$ is a linear predictor, and $\epsilon_i \sim {\rm Normal}(0,\tau_\epsilon)$ is iid Gaussian error with precision parameter $\tau_\epsilon$.  The specification in Eq. \ref{eq:basic_model} is thus doubly stochastic, in the sense that we assume error associated with $f_Y$ as well as in the location parameter $\mu_i$.  This setup can be useful computationally, and can also be used to approximate singly stochastic system by setting $\tau_\epsilon$ to a large value.  For instance, models for count data often assume a Poisson error structure with a log link function to ensure that the Poisson intensity parameter is greater than zero.  In this case, we would specify
\begin{linenomath*}
\begin{equation*}
  Y_i \sim \mathrm{Poisson}(\exp(\theta_i + \epsilon_i)),
\end{equation*}
\end{linenomath*}
a configuration known as a log-Gaussian Cox process.
We now describe how different classes of statistical models can be developed depending on how one structures the linear predictor, $\theta_i$.

\section{Models}
\subsection{Generalized linear models (GLM)}

Generalized linear models \citep{McCullaghNelder1989} are one of the simplest (and most often used) statistical models
used by ecologists to make spatial predictions. In GLMs, the linear predictor is a simple linear function of gathered covariates
(including possible quadratic terms of these covariates.  Statisticians often describe this relationship by
$\boldsymbol{\theta}={\bf X} \boldsymbol{\beta}$, which is written in matrix notation.  In particular, $\boldsymbol{\theta}$ gives
a vector of linear predictor values (one for each data point being analyzed), $\boldsymbol{\beta}$ gives a vector of regression coefficients, and ${\bf X}$ is a design matrix which includes all explanatory variables (and often a column vector of ones to represent an intercept).

\subsection{Generalized additive models (GAM)}

Although one can allow nonlinear relationships between response variables and regressors by including polynomial terms in the design matrix of a GLM, these need to be pre-selected by the analyst and it is often unclear how many such terms one should include. Generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} build upon generalized linear models, but instead allow smooth relationships between the dependent and independent variables using flexible functions such as splines.  Such models have
been employed in a number of spatial prediction scenarios, including transect sampling models for animal abundance \citep{HedleyBuckland2004} and SDMs \citep{GuisanEtAl2002}.  For instance, animal density or presence can be modeled as a smooth, unknown function of a habitat covariate.

There are a number of ways smooth relationships can be modeled; in order to formulate a GAM in the notation of Eq. \ref{eq:basic_model} we employ a knot-based kernel smoother with a radial basis function.  The basic notion is to place $k_j$ knots, which we will denote by $\boldsymbol{\omega}_j = \{ \omega_{1j},\omega_{2j},\hdots,\omega_{k_j j} \}$, throughout the range of a given regressor $j$, where the number of knots controls the level of smoothing.  For each regressor of interest, we can then calculate a matrix ${\bf K}_j$ of dimension $(n,k_j)$, with entries $\mathcal{N}(X_{ij};\omega_j,\tau_{\omega,j})$, where $\mathcal{N}(x;\mu,\tau)$ denotes a Gaussian distribution with mean $\mu$ and precision $\tau$.  We can then introduce additional regression coefficients $\boldsymbol{\alpha}_j$ to help model the smooth relationship, and incorporate these into the linear predictor.  In the case of one smooth term, we then have  in Eq. \ref{eq:basic_model}:
\begin{linenomath*}
\begin{equation}
  \label{eq:GAM}
  \boldsymbol{\theta}={\bf X} \boldsymbol{\beta} + {\bf K}_j \boldsymbol{\alpha}_j
\end{equation}
\end{linenomath*}
When fitting such a model, one needs to select the number and location of knots, as well as the precision parameter $\tau_{\omega,j}$ for the Gaussian basis kernels.  This could be done using some notion of optimality or prediction error (as with cross validation).  In practice, a simple recommendation is to set $\tau_{\omega,j}$ less than the range of the modeled covariate, and greater than the typical difference between observed values of the covariate.

Our main point in writing the GAM as in Eq. \ref{eq:GAM} is to emphasize the structural similarity between GLMs and certain classes
of GAMs.  In particular,
we can rewrite Eq. \ref{eq:GAM} as
$\boldsymbol{\theta}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
where ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K}_j \right]$ (i.e., concatenating ${\bf X}$ and ${\bf K}_j$ horizontally), and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha}_j \right] ^\prime$ (the subscript $aug$ denotes augmentation).  We shall exploit this structure when examining
extrapolation bias.  We note that similar formulations for GAMs (e.g. using thin plate splines) can be represented in a similar form \citep[cf.][]{CrainiceanuEtAl2005}.

\subsection{Introducing spatial and/or temporal autocorrelation: spatio-temporal regression models (STRMs)}

The previous two modeling frameworks (GLMs and GAMs) do not acknowledge spatial autocorrelation above and beyond that induced
by modeled covariates.  However, it is common for residuals from GLM and GAM model fits to include spatial autocorrelation, which violates their common assumption of independently distributed error \citep{Legendre1993,LichsteinEtAl2002}.  Parameter estimates from GLMs and GAMs that display residual autocorrelation should be interpreted with caution, as they will tend to have overstated precision and may even be biased.  In such situations, analysts often employ spatial regression models which explicitly account for spatial autocorrelation above and beyond that explained by modeled covariates.

There are a variety of ways spatial autocorrelation can be included in regression models, depending on (i) the spatial support (i.e., continuous vs. discrete), and (ii) the particular mechanism used to impart correlation.  Here, we shall focus on areal models for discrete spatial support, as our impression is that these are more commonly employed in ecological studies.  Such models require that data are aggregated at the level of some sample unit (plots, grid cells, etc.).  Spatio-temporal regression models (STRMs) for areal data are often specified in a similar fashion to GLMs and GAMs:
\begin{linenomath*}
\begin{equation}
  \label{eq:spat_reg}
  \boldsymbol{\theta}={\bf X} \boldsymbol{\beta} + \boldsymbol{\eta},
\end{equation}
\end{linenomath*}
where $\boldsymbol{\eta}$ represent spatially autocorrelated effects.  In this treatment we shall consider two approaches for
inducing spatial autocorrelation in $\boldsymbol{\eta}$: process convolution
\citep[PC;][]{Higdon1998} and restricted spatial regression \citep[RSR;][]{Reich2006,Hodges2010,Hughes2013}.

The PC implementation works by placing $k$ knots throughout the spatial domain being analyzed, and like our GAM formulation, uses distances from the center point of each sample unit to each of these knot locations to induce spatial structure.  Using similar notation to our GAM formulation, we denote these knots as $\boldsymbol{\omega} = \{ \boldsymbol{\omega_{1}},\boldsymbol{\omega}_{2},\hdots,\boldsymbol{\omega}_{k} \}$; the $\boldsymbol{\omega}_m$ are bivariate in this  instance as space is assumed to be two dimensional.  We then construct a matrix ${\bf K}$ of dimension $(n,k)$ with entries $\mathcal{BVN}({\bf s}_i;\boldsymbol{\omega}_m,\tau_{\omega})$, where $\mathcal{BVN}({\bf x};{\bf \mu},\tau)$ denotes a bivariate Gaussian distribution with mean $\boldsymbol{\mu}$ and precision $\tau$.  As with our GAM model we introduce additional regression coefficients $\boldsymbol{\alpha}$ to estimate the level of spatial smoothing conditional on the assumed knot structure and reparameterize the STRM as
\begin{linenomath*}
\begin{equation}
  \label{eq:PC}
  \boldsymbol{\theta}={\bf X} \boldsymbol{\beta} + {\bf K} \boldsymbol{\alpha}
\end{equation}
\end{linenomath*}
This model is structurally very similar to the GAM model (indeed it may be interpreted as a GAM with a bivariate smooth on spatial location).  Similarly, we can rearrange terms to write it similar to the GLM, such that
$\boldsymbol{\theta}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
where in this case ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K} \right]$, and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha} \right] ^\prime$.

The RSR approach to spatial regression uses a reduced-rank version of the popular intrinsic conditionally autoregressive \citep[ICAR;][]{Besag1995,Rue2005} model for spatial random effects, reparameterized so that basis vectors are orthogonal to the main effects of interest.  This approach has generated substantial recent interest, as fixed effects retain primacy in explaining variation in the ecological process of interest and problems with spatial confounding between fixed and random effects are eliminated. As such, spatial random effects are only used to account for residual autocorrelation \citep{Reich2006,Hodges2010} and the decision to incorporate spatial autocorrelation has little effect on the point estimates of fixed effects.  In addition, reduced dimension spatial models such as RSR lighten computational burden while still accounting for course-scale spatial autocorrelation \citep[see e.g.][]{LatimerEtAl2009,Wikle2010,Hughes2013}.  It turns out that RSR models can also be written in the form of Eq. \ref{eq:PC}, using a
different choice of ${\bf K}$ \citep{Hughes2013}, constructed as follows:
\begin{enumerate}
  \item Define the residual projection matrix ${\bf P}^\perp={\bf I}-{\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$
  \item Calculate the Moran operator matrix $\boldsymbol{\Omega}=J{\bf P}^\perp \mathcal{W}{\bf P}^\perp/{\bf 1}^\prime \mathcal{W} {\bf 1}$
  \item Define ${\bf K}$ as an $(n \times m)$ matrix, where the columns of ${\bf K}$ are composed of the eigenvectors  associated with the largest $m$ eigenvalues of $\boldsymbol{\Omega}$. Hughes and Haran (\citeyear{Hughes2013}) used simulation to explore a range of such values and concluded $m=50-100$ should suffice for most applications.
\end{enumerate}
Here, ${\bf I}$ an identity matrix, ${\bf 1}$ is a column vector of ones, and $\mathcal{W}$ represents an association matrix describing the spatial neighborhood structure of sampling units.  For instance, for a first order neighborhood structure, $\mathcal{W}$ would include a 1 for all rows $i$ and columns $j$ where sampling unit $i$ and $j$ are neighbors \citep[see][for alternative association matrices]{Rue2005}.

\renewcommand{\refname}{Literature Cited}
\bibliographystyle{ecology}

\bibliography{master_bib}

\end{flushleft}
\end{document}

### 2014-08-25 manuscript/SpatPred3.tex
Initial commit

\documentclass[12pt,english]{article}

\usepackage{amsmath,amssymb,amsthm,epsfig,lineno,rotfloat,psfrag,natbib,caption,setspace,url,bm,geometry}
\usepackage{ecology}
\usepackage{graphicx}

\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}

%\setlength{\evensidemargin}{0in} \setlength{\oddsidemargin}{0in}
%\setlength{\topmargin}{-0.65in} \setlength{\textwidth}{6.5in}
%\setlength{\textheight}{9.5in} \setlength{\topskip}{0in}
%\setlength{\headheight}{0in}

%\newcommand{\yst}{\ensuremath{y_{st}}}
%\newcommand{\tyst}{\ensuremath{\tilde y_{st}}}
%\newcommand{\tzs}{\ensuremath{\tilde z_{s}}}
%\newcommand{\zs}{\ensuremath{z_{s}}}
%\newcommand{\bg}{\ensuremath{\boldsymbol{\gamma}}}
%\newcommand{\bxs}{\ensuremath{\mathbf{x}_{s}}}
%\newcommand{\bxst}{\ensuremath{\mathbf{x}_{st}}}
%\newcommand{\bK}{\ensuremath{\mathbf{K}}}
%\newcommand{\bn}{\ensuremath{\boldsymbol{\eta}}}
%\newcommand{\bb}{\ensuremath{\boldsymbol{\beta}}}
%\newcommand{\ba}{\ensuremath{\boldsymbol{\alpha}}}
%\newcommand{\by}{\ensuremath{\mathbf{y}}}
%\newcommand{\bz}{\ensuremath{\mathbf{z}}}
%\newcommand{\bX}{\ensuremath{\mathbf{X}}}
%\newcommand{\be}{\ensuremath{\boldsymbol{\epsilon}}}
%\newcommand{\bQ}{\ensuremath{\mathbf{Q}}}
%\newcommand{\bO}{\ensuremath{\boldsymbol{\Omega}}}
%\newcommand{\bm}{\ensuremath{\boldsymbol{\mu}}}

\bibliographystyle{ecology}
\raggedbottom

\captionsetup[table]{margin=0pt,font=small,labelfont={sc},justification=justified,labelsep=period}
\captionsetup[figure]{margin=0pt,font=small,labelfont={sc},justification=justified,labelsep=period,name=Fig}

\begin{document}
\begin{spacing}{1.9}

\begin{center}
Avoiding extrapolation bias when using statistical models to make ecological prediction
\bigskip\\
\normalsize
{\sc Paul B. Conn$^{1,}$\footnotemark[2] and
Devin S. Johnson$^1$, Peter L. Boveng$^1$ }\smallskip\\
$^1${\em National Marine Mammal Laboratory, NOAA, National Marine Fisheries Service,
Alaska Fisheries Science Center, 7600 Sand Point Way NE, Seattle,
WA 98115 USA }\\ \medskip
\end{center}
\footnotetext[2]{Email: paul.conn@noaa.gov}

\raggedright \setlength{\parindent}{0.3in}
%\renewcommand{\baselinestretch}{1.8}\normalsize
\clubpenalty=0
 \linenumbers

{\em Abstract.\ } We'll do this later

{\em Key words: Abundance, Extrapolation, Forecasting, General additive models, Generalized linear models, Independent Variable Hull, Leverage, Occupancy, Spatial regression}

\centerline{\sc Introduction}

In ecology and conservation, a common goal is to make predictions about an unsampled random variable given a limited sample from the target population.  For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_i$, we often desire to predict a new observation $y_{new}$ at $i$.  For instance, we might use a generalized linear model \citep{McCullaghNelder1989} or one of its extensions to predict species density or occurrence in a new location, or to predict the future trend of a population.

Early in their training, ecologists and statisticians are warned against extrapolating statistical relationships
past the range of observed data.  This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new point $y_{new}$ whenever $x_{new} < \min({\bf x})$ or $x_{new} > \max({\bf x})$.  But what about more complicated situations
where there are multiple explanatory variables, or when one uses a spatial regression model to account for the residual spatial autocorrelation that is inevitably present in patchy ecological data \citep{LichsteinEtAl2002}?  How reliable are spatially- or temporally-explicit predictions in sophisticated models for animal abundance and occurrence?

Statisticians have long struggled with the conditions under which fitted regression models are capable of
making robust predictions at new combinations of explanatory variables.  The issue is sometimes considered more of a
philosophical problem than a statistical one, and has even been likened to soothsaying \citep{EhrenbergBound1993}.  To our mind, the reliability of predictions from statistical models is likely a function of several factors, including (i) the intensity of sampling,
(ii) spatial or temporal proximity of the prediction location to locations where there are data, (iii) variability of the ecological process, and (iv) the similarity of explanatory covariates in prediction locations when compared to the ensemble of covariates for observed data locations.

Our aim in this paper is to investigate extrapolation bias in the generalized linear model and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatial, temporal, or spatio-temporal regression models (STRMs).  In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers \citep{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points. Application of the gIVH can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction.  Also, since the gIVH can be constructed solely with knowledge of sampled locations and explanatory covariates (i.e., it does not necessarily require any observed response variables), it can also be used to help guide survey design. We illustrate use of the gIVH on a simulated occupancy dataset, on a species distribution model (SDM) for ribbon seals in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).

\section{The generalized independent variable hull (gIVH)}

Extrapolation is often distinguished from interpolation.  In a prediction context, we might define (admittedly quite imprecisely) that extrapolation consists of making predictions that are ``outside the range of observed data" while interpolation consists of making predictions ``inside the range of observed data."  But what exactly do we mean by ``outside the range of observed data"?  Predictions outside the range of observed covariates?  Predictions for locations that are so far from places where data are gathered that we are skeptical that the estimated statistical relationship still holds? To help guide our choice of an operational definition, we turn to early work on outlier detection in simple linear regression analysis.

In the context of outlier detection, Cook (\citeyear{Cook1979}) defined an independent variable hull (IVH) as the smallest convex set containing all design points of a full-rank linear regression model.  Linear regression models are often written in matrix form, i.e.
\begin{linenomath*}
\begin{equation*}
  {\bf Y} = {\bf X} \boldsymbol{\beta} + \boldsymbol{\epsilon},
\end{equation*}
\end{linenomath*}
where ${\bf Y}$ give observed data, ${\bf X}$ is a so-called design matrix that includes explanatory variables \citep[see e.g.][]{Draper1966}, and $\boldsymbol{\epsilon}$ represent normally distributed residuals (here and throughout the paper, bold symbols will be used to denote vectors and matrices).
Under this formulation, the IVH is defined relative to the hat matrix, ${\bf V}_{\rm LR}={\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$ (where the subscript ``LR" denotes linear regression).  Letting $v$ denote the maximum diagonal element of ${\bf V}_{LR}$, one can examine whether a new design point, ${\bf x}_0$ is within the IVH.  In particular, ${\bf x}_0$ is within the IVH whenever
\begin{linenomath*}
\begin{equation}
  \label{eq:IVH}
  {\bf x}_0^\prime ({\bf X}^\prime{\bf X})^{-1}{\bf x}_0 \le v.
\end{equation}
\end{linenomath*}
Cook (\citeyear{Cook1979}) used this concept to identify influential observations and possible outliers, arguing that design points near the edge of the IVH are deserving of special attention.  Similarly, points outside the IVH should be interpreted with caution.

We simulated two sets of design data to help illustrate application of the IVH (Fig. \ref{fig:IVH}).  In simple linear regression with one predictor variable, predictions on a hypothetical response variable obtained at covariate values below the lowest observed value or above the highest observed value are primarily outside of the IVH.  We suspect this result conforms to most ecologists intuition about what constitutes ``extrapolating past the observed data."  However, fitting a quadratic model exhibits more nuance; if there is a large gap between design points, it is entirely possible that intermediate covariate values will also be outside of the IVH and thus more likely to result in problematic predictions.  Fitting a model with two covariates and both linear and quadratic effects, the shape of the IVH is somewhat more irregular, and even includes a hole in the middle of the surface when interactions are modeled (Fig. \ref{fig:IVH}).  These simple examples highlight the sometimes counterintuitive nature of the predictive inference problem, a problem that can only become worse as models with more dimensions are contemplated (including those with temporal or spatial structure).  Fortunately, the ideas behind the IVH provide a potential way forward.

Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes iid normally distributed error. Thus, it is not directly applicable to generalized models, such as those including spatial random effects.  Further, the hat matrix is not necessarily well defined for models with more general spatial structure. However, since the hat matrix is proportional to prediction variance, Cook (1979) notes that design points with maximum prediction variance will be located on the boundary of the IVH.  Working on the linear predictor scale, we therefore define a generalized independent variable hull (gIVH) as the set of all points ${\bf x}$ (note that ${\bf x}$ can include both observed and unobserved design points) such that
\begin{linenomath*}
\begin{equation}
  \label{eq:gIVH}
  {\rm var}(\boldsymbol{\mu}_x | {\bf x}) \le {\rm max} [{\rm var}(\boldsymbol{\mu}_X | {\bf X})],
\end{equation}
\end{linenomath*}
where $\boldsymbol{\mu}_x$ correspond to predictions at ${\bf x}$, and ${\bf X}$ and $\boldsymbol{\mu}_X$ denote observed design points and predictions at $X$, respectively.

Generalizations of the linear model are often written in the form
\begin{linenomath*}
\begin{equation}
  Y_i \sim f_Y(g^{-1}(\mu_i)),
\end{equation}
\end{linenomath*}
where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function, and
$\mu_i$ is a predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
\begin{linenomath*}
\begin{equation}
  \boldsymbol{\mu} = {\bf X}_{aug}\boldsymbol{\beta}_{aug} + \boldsymbol{\epsilon},
  \label{eq:mu-aug}
\end{equation}
\end{linenomath*}
where the $\boldsymbol{\epsilon}$ represent Gaussian error, ${\bf X}_{aug}$ denotes an augmented design matrix, and $\boldsymbol{\beta}_{aug}$ denote an augmented vector of parameters.  For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see Appendix S1 for examples of how numerous types of models can be written in this form).

When models are specified as in Eq. \ref{eq:mu-aug}, we can write prediction variance generically as
\begin{linenomath*}
\begin{equation}
  \label{eq:var-mu}
  {\rm var}(\boldsymbol{\mu}_x | {\bf x})= {\bf x} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) {\bf x}^\prime,
\end{equation}
\end{linenomath*}
where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; Appendix S1). For GLMs, prediction variance is proportional to the original hat matrix ${\bf V}_{LR}$ (at least on the linear predictor scale), and Eq. \ref{eq:IVH} may be used directly.  For other models, we have
\begin{linenomath*}
\begin{equation}
  \label{eq:var_Beta}
  {\rm var}(\hat{\boldsymbol{\beta}}_{aug})=\left[ \begin{array}{cc}
      \boldsymbol{\Sigma}_\beta & {\bf 0}\\
      {\bf 0} & \boldsymbol{\Sigma}_\alpha
  \end{array} \right],
\end{equation}
\end{linenomath*}
where we simply replace $\boldsymbol{\beta}_{aug}$ with the augmented vector that includes both regression and spatial or smooth parameters (i.e. $\boldsymbol{\beta}$ and $\boldsymbol{\alpha}$; Appendix S1).

The exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ differs depending on the underlying model structure and estimation procedure.  In the following treatment, we shall focus on Bayesian analysis; although it is not necessarily needed to fit GLMs and GAMs, it helps make STRMs more computationally tractable and puts all of the different models into a common analysis framework.  This approach requires specifying priors for model parameters, and prior parameters may appear in expressions for prediction variance.  Judicious choices of priors can at times limit the influence of priors on calculation of the gIVH.  For example, if we specify the prior on the vector of regression coefficients to be $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$ where $\tau_\beta$ is a fixed constant, a Bayesian implementation of a GLM still results in the gIVH specified in Eq. \ref{eq:IVH}.  The ability to use Eq. 5 directly is quite useful, as we only need to know the explanatory variables to be able to diagnose whether predictions lie in or out of the gIVH. However, the other models considered here (GAMs and STRMs) typically require estimation of an additional precision parameter.   In these cases, one solution is to impose prior distributions and integrate over precision parameters (call these $\boldsymbol{\tau}$) to obtain an expectation for ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$.  Using the law of the unconscious statistician, we have
\begin{linenomath*}
\begin{equation}
  \label{eq:E_var_Beta}
  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}))  = \int_{\boldsymbol{\tau}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) [\boldsymbol{\tau}] d \boldsymbol{\tau}.
\end{equation}
\end{linenomath*}
Here, and throughout the paper, we use the bracket notation to indicate a probability density function; e.g., $[\boldsymbol{\tau}] $ denotes the joint prior distribution for precision parameters $\boldsymbol{\tau}$. If observations have already been gathered, one could also consider integrating over the marginal posterior distribution using MCMC:
\begin{linenomath*}
\begin{equation}
  \label{eq:E_var_Beta_post}
  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}) | {\bf Y})  = \int_{\boldsymbol{\tau}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) [\boldsymbol{\tau}| {\bf Y}] d \boldsymbol{\tau}.
\end{equation}
\end{linenomath*}

We propose to use the gIVH in much the same manner as Cook (1979).  In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).  To our knowledge, this is the first time this definition has been used in the context of spatial prediction, but one we shall show can be quite helpful in diagnosing and mitigating extrapolation bias.  The gIVH seems ideally situated to this task as it does not necessarily need to rely on gathered response data. Thus, one can examine whether or not prediction points lie within the IVH without ever collecting response data there.

\section{Examples}

\subsection{Simulation study}

We conducted simulation analyses to investigate whether the gIVH was useful in diagnosing prediction biases in occupancy and abundance analyses. In both cases, we abundance on a $30 \times 30$ grid, and generated as a function of four habitat covariates in addition to spatial process covariance (Appendix S2). The habitat covariates also

\subsection{Ribbon seal SDM}
As part of an international effort, researchers with the U.S. National Marine Fisheries Service conducted aerial surveys over the eastern Bering Sea in 2012 and 2013.  Agency scientists used infrared video to detect seals that were on ice, and simultaneous automated digital photographs provided information on species identity. Here, we use spatially referenced count data from photographed ribbon seals, {\it Phoca fasciata} (Fig. \ref{fig:ribbon}) on a subset of 10 flights flown over the Bering Sea in April 2012.  These flights were limited to a one week period so that both sea ice conditions and seal distributions could be assumed to be static.

Our objective with this dataset will be to model seal counts on transects through 25km by 25km grid cells as a function of habitat covariates and possible spatial autocorrelation. Estimates of apparent abundance can then be obtained by summing predictions across grid cells. Figure \ref{fig:flights} shows the transects flown and the number of ribbon seals encountered in each cell, and Figure \ref{fig:covs} show explanatory covariates gathered to help predict ribbon seal abundance.  These data are described in fuller detail by \citet{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors \citep[see e.g.][]{ConnEtAl2014}.  Since our focus in this paper is on illustrating spatial modeling concepts, we devote our efforts to the comparably easier problem of estimating apparent abundance (i.e., uncorrected for vagaries of the detection process).

Inspection of ribbon seal data (Fig. \ref{fig:flights}) immediately reveals a potential
issue with spatial prediction: abundance of ribbon seals appears to be maximized in the southern and/or southeast quadrant of the surveyed area.  Predicting abundance in areas further south and east may thus prove problematic.  To illustrate, let $Y_i$ denote the ribbon seal count ($Y_i$) obtained in sampled grid cell $i$.  Suppose that counts arise according to a log-Gaussian Cox process, such that
\begin{eqnarray}
  \label{eqn:Cox}
  Y_i \sim {\rm Poisson}(\lambda_i) \mbox{ and } \\
  \log(\lambda_i) = \log(P_i) + {\bf X}_i \boldsymbol{\beta} + \eta_i + \epsilon_i, \nonumber
\end{eqnarray}
where $P_i$ gives the proportion of area photographed in grid cell $i$ (recall also that ${\bf X}_i$ denotes a vector covariates for cell $i$, $\boldsymbol{\beta}$ are regression coefficients, $\eta_i$ represents a spatially autocorrelated random effect, and $\epsilon_i$ is normally distributed $iid$ error).

We could fit any number of predictive models to these data, but we start with a simple
generalized linear model where we ignore the spatial random effect, $\eta_i$, and use the full suite of predictor covariates (Fig. \ref{fig:covs}) to fit Eq. \ref{eqn:Cox} to our data. In particular, we fit a model with linear effects of all predictor variables, and with an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}. To enable comparison with more complicated types of models, we formulated a generalized Bayesian strategy for parameter estimation (see Appendix S1).  For simplicity, we generated posterior predictions of ribbon seal abundance across the landscape as
\begin{equation}
  N_i \sim {\rm Poisson}(A_i \lambda_i),
\end{equation}
where $A_i$ gives the proportion of suitable habitat in cell $i$ (ribbon seals do not inhabit land masses).

Fitting this model to our data,

\subsection{Steller sea lion trends}

\section{Discussion}

A number of authors have explored optimal knot placement in spatial models.  In the context of predictive process modeling \citep[where a covariance function is specified over a group of knots; see][]{BanerjeeEtAl2008}, \citet{FinleyEtAl2009} and \citet{GelfandEtAl2013} considered near-optimal knot placement by minimizing spatially averaged prediction variance.
Gelfand et al. knot selection

Using GAs to select knot placement using

Contrast with posterior loss (e.g., Jay's linex loss function estimator)

Contrast with cross validation

David Miller (the evil one)/Simon Wood stuff on edge effects

Contrast with other approaches-
Gelfand et al. Bayesian analysis - intrinsic CAR in SDM
Chakraborty et al. '10 - spatial abundance modeling - ordinal
Latimer et al. 2009 - spatial predictive process modelling in SPDs

Can't be complacent... still possible to get poor/biased results, e.g. if $\tau_\epsilon \rightarrow 0$.  Can't resolve pathological problems.

Presence-absence data, other link functions (e.g. probit)

extensions to models w/ secondary observation process, measurement error

much attention has been given to collinearity in multiple linear regression - suggest researchers give as much attention to predictive extrapolation bias in predictive models

For predictions with spatial , our experience is that predictions outside the minimum convex polygon where data are obtained can sometimes be more problematic than predictions within the polygon.  Spatial prediction surfaces may have a tendency to bend up or down in these areas, resulting in ``edge effects" that can lead to positive prediction bias when a log link function is employed \citep{VerHoefJansen2007}.

Fieberg PVA

\bibliography{master_bib}

FIGURE 1. A ribbon seal, {\it Phoca fasciata}; the focus of spatial modeling efforts in this paper.

FIGURE 2. Examples IVHs constructed from simulated data.  In panels A and B, the investigator plans to model a hypothetical (unmeasured) response variable using a linear regression model as a function of a single covariate, $x$, obtained at a number of design points (denoted with an ``x").  Using $x$ as a simple linear effect (A), only predictions less than the minimum observed value of $x$ or greater than the maximum value of $x$ are outside the IVH (shaded area), as scaled prediction variance in these areas (solid line) are greater than the maximum scaled prediction variance for observed data (dashed line).  Using both linear and quadratic effects of $x$, some intermediate points are also outside the IVH; predictions at these points should also be viewed with caution.  Panels C $\&$ D show a more complicated IVH when the investigator wishes to relate an unmeasured response variable to linear and quadratic effects of two covariates, $x$ and $y$, either without interactions (C) or with interactions (D).  Any potential predictions in the shaded area are outside of the IVH and should be viewed with caution.

FIGURE 3. Aerial surveys over the Bering Sea in spring of 2012 (blue lines) overlayed on
a tesselated surface composed of 25km by 25km grid cells.  Gray indicates land, and colored pixels indicate ribbon seal encounters (yellow: 1-2 seals; orange: 3-4 seals; magenta: 5-9 seals; red: 10-15 seals).  On average, photographs covered approximately $2.6km^2$ ($0.4\%$) of each surveyed grid cell.

FIGURE 4. Potential covariates gathered to help explain and predict ribbon seal abundance in the eastern Bering Sea.  Covariates include distance from mainland (\texttt{dist\_mainland}), distance from 1000m depth contour (\texttt{dist\_shelf}), average remotely sensed sea ice concentration while surveys were being conducted (\texttt{ice\_conc}), and distance from the southern sea ice edge (\texttt{dist\_edge}).  All covariates except ice concentration were standardized to have a mean of 1.0 prior to plotting and analysis.

FIGURE 5  Posterior median estimates of ribbon seal apparent abundance across the eastern Bering sea for (A) a generalized linear model (GLM), (B) a generalized additive model (GAM), (C) a GLM with known zero data, and (D) a GAM with known zero data.  Highlighted cells indicate those where predictive covariate values are outside of the generalized independent variable hull.

\clearpage
\centerline{\sc Figures}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{ribbon}
\end{center}
\caption{ }
\label{fig:ribbon}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{IVH_simple2.pdf}
\end{center}
\caption{ }
\label{fig:IVH}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{test_flights_optimized.pdf}
\end{center}
\caption{ }
\label{fig:flights}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{covariates.pdf}
\end{center}
\caption{ }
\label{fig:covs}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{Bering_noRE_maps.pdf}
\end{center}
\caption{ }
\label{fig:covs}
\end{figure}

\clearpage

\end{spacing}
\end{document}

### 2014-08-25 SpatPred/R/mcmc_abundance.R

@@ -1,0 +1,204 @@
+#' function to perform Bayesian analysis of count data using a spatial log Gaussian Cox process model
+#' @param model A formula object specifying the linear predictor for abundance intensity
+#' @param Data   A list holding the following objects:
+#'        Adj - Adjacency matrix of ones and zeros describing spatial connectivity (for ICAR modeling)
+#'        Grid - A data frame holding covariate values for all cells one wants to make predictions on (including sampled cells)
+#'        Count.data - A data frame with the following columns: (1) "Cell" - entries this column provides the numerical index (in 1:S) for the cell surveyed
+#'              (3) "AreaSurveyed" - provides
+#'              the proportion of the target cell that was surveyed, (4) "Count" Number of target species counted
+#' @param Area.adjust   A vector allowing for differences in suitable habitat for each cell.  Can be used for different grid cell sizes or different
+#'        proportions of suitable habitat (e.g., 1.0 = 100% of habitat is suitable for the focal species)
+#' @param Control A list giving MCMC controls and additional model options
+#'  "iter": number of MCMC iterations;
+#'  "burnin": number of MCMC burnin iterations;
+#'	"thin": if specified, how many iterations to skip between recorded posterior samples;
+#'  "srr.tol": Threshold eigenvalue level for SRR; only eigenvectors with higher eigenvalues than srr.tol are included in SRR formulation (default is 0.5)
+#'  "predict" If TRUE (default), calculate posterior predictions across the entire grid
+#'  "MH.mu" A vector providing continuous Uniform half-range values for Metropolis-Hastings proposals
[…]

### 2014-08-25 SpatPred/R/sim_data_abundance.R

@@ -1,0 +1,111 @@
+#' function to simulate spatial count data over a simulated landscape
+#' @param S Number of cells (must be a square number)
+#' @param n.covs Number of habitat covariates (default 4)
+#' @param tau.epsilon Precision of exchangeable errors in log space
+#' @return A list object composed of at least two objects, "Data$Grid" is a list vector holding covariate data by day,
+#'         and "Count.data" which holds simulated transect counts 
+#' @export
+#' @keywords spatio-temporal, simulation
+#' @author Paul B. Conn
+sim_data_generic<-function(S,n.covs=4,tau.epsilon=20){
+  require(ggplot2)
+  require(Matrix)
+  require(spatstat)
+  
+  tau.epsilon=20
+  
+  source('./SpatPred/R/util_funcs.R')
+   
+  if(sqrt(S)%%1>0)cat("error in sim_data_generic; S must be a square number \n")
+  
+  x.len=sqrt(S)
+ 
+  #generate Covariates from a Matern distribution
+  Covs=matrix(0,S,n.covs)
+  for(icov in 1:n.covs){
+    kappa=runif(1,5,10)
+    r=runif(1,0.1,.5)
+    mu=runif(1,500,1500)
+    Dat.matern=rMatClust(kappa,r,mu)  
+    X=round((x.len)*Dat.matern$x+0.5)
+    Y=round((x.len)*Dat.matern$y+0.5)
+    X[X<1]=1
+    Y[Y<1]=1
+    X[X>x.len]=x.len
+    Y[Y>x.len]=x.len
+    Grid=matrix(0,x.len,x.len)
+    for(i in 1:length(X))Grid[X[i],Y[i]]=Grid[X[i],Y[i]]+1
+    Covs[,icov]=Grid/max(as.vector(Grid))
+  }  
+
+  col.names="cov1"
+  if(n.covs>1){
+    for(i in 2:n.covs)col.names=c(col.names,paste("cov",i,sep=''))
+  }
+  colnames(Covs)=col.names
+  
+  X=model.matrix(~poly(Covs,degree=2,raw=TRUE))
+  Beta=c(rnorm(1,3,0.5),rnorm(2*n.covs+gamma(n.covs),0,0.8))
+  
[…]

### 2014-08-25 SpatPred/R/spat_pred.R

@@ -1,0 +1,277 @@
+#' function to perform Bayesian analysis of log Gaussian Cox processes and to make spatial predictions 
+#' @param formula A formula object specifying the model for the fixed effects
+#' @param Data   A SpatialPolygongsDataFrame holding covariate values for all cells one wants to make predictions on (including sampled cells)
+#'  The data frame should also include a column, "Count" which gives the response variable (only values associated with surveyed cells will be modeled.)
+#' @param Mapping  A vector giving the habitat cell id number for each surveyed cell
+#' @param Area.adjust   A vector allowing for differences in suitable habitat for each cell.  Can be used for different grid cell sizes or different
+#'        proportions of suitable habitat (e.g., 1.0 = 100% of habitat is suitable for the focal species)
+#' @param Offset	A vector giving the fraction of each grid cell in Mapping that was surveyed
+#' @param spat.mod Controls whether (and what type) of spatial autocorrelation is implemented (spatmod = 0: no spatial autocorrelation; spatmod=1: process convolution; spatmod=2: srr ICAR) 
+#'        Note that if spatmod=1, the argument Knots needs to be specified; if spatmod=2, Adj must be specified
+#' @param Assoc   Association matrix for all grid cells in Data (needed for ICAR models only)
+#' @param K   A (# cells)x(# knots) matrix - the 'K' matrix for process convolution modeling.  
+#' @param Names.gam A character vector giving the column names of Data that will be modeled as smooth effects
+#' @param Knots.gam A list vector, where each entry is a matrix (default NULL)
+#' @param Control A list giving MCMC controls and additional model options
+#'  "iter": number of MCMC iterations;
+#'  "burnin": number of MCMC burnin iterations;
[…]

### 2014-08-25 SpatPred/inst/compile_plot_ribbon_data.R

@@ -1,0 +1,116 @@
+# script to compile count and effort data for ribbon seals from SLR data and make plots
+#
+# Paul B. Conn
+# 
+
+library(sp)
+library(rgdal)
+library(nPacMaps)  #from Josh London
+library(rgeos)
+library(doBy)
+library(ggplot2)
+
+#First, pull in data
+load('c:/users/paul.conn/git/SpatPred/BOSS_Power_photo_data/slr_power_data.RData') #sp data frame including species & area of photo
+load('c:/users/paul.conn/git/BOSS/BOSS/data/AlaskaBeringData.Rdat') #read in "Data" holding grid info
+load('c:/users/paul.conn/git/BOSS/BOSS/data/Effort_points.Rdat')  #read in Sp Points DF "effort_data" for on effort points (use for covariates)
+
+I.inter=gIntersects(slr_power_data,Data$Grid[[1]],byid=TRUE)
+Points=slr_power_data[which(apply(I.inter,2,'sum')>0),]  #only include effort points that are 'on grid'
+n.points=nrow(Points)
+
+#do some plotting
+plot(Data$Grid[[1]])
+points(Points,pch=19,cex=0.01,col='blue')
+
+data(alaska_dcw)
+data(russia_dcw)
+laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
+                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
+alaska_dcw <- spTransform(alaska_dcw, CRS(laea_180_proj))
+russia_dcw <- spTransform(russia_dcw, CRS(laea_180_proj))
+plot(alaska_dcw, col = "gray", add = TRUE)
+plot(russia_dcw, col = "gray", add = TRUE)
+                                                                     
+                                                                     
+
+##create a single grid object for this analysis, averaging time varying covariates over the dates considered
+Cur.grid=Data$Grid[[1]]
+n.cells=dim(Cur.grid)[1]
+S=nrow(Cur.grid)  #number of cells
+#first object corresponds to April 4 so start with 17th SpDF
+Mean.covs=matrix(0,S,3)
[…]

### 2014-08-25 SpatPred/inst/compile_plot_spotted_data.R

@@ -1,0 +1,116 @@
+# script to compile count and effort data for ribbon seals from SLR data and make plots
+#
+# Paul B. Conn
+# 
+
+library(sp)
+library(rgdal)
+library(nPacMaps)  #from Josh London
+library(rgeos)
+library(doBy)
+library(ggplot2)
+
+#First, pull in data
+load('c:/users/paul.conn/git/SpatPred/BOSS_Power_photo_data/slr_power_data.RData') #sp data frame including species & area of photo
+load('c:/users/paul.conn/git/BOSS/BOSS/data/AlaskaBeringData.Rdat') #read in "Data" holding grid info
+load('c:/users/paul.conn/git/BOSS/BOSS/data/Effort_points.Rdat')  #read in Sp Points DF "effort_data" for on effort points (use for covariates)
+
+I.inter=gIntersects(slr_power_data,Data$Grid[[1]],byid=TRUE)
+Points=slr_power_data[which(apply(I.inter,2,'sum')>0),]  #only include effort points that are 'on grid'
+n.points=nrow(Points)
+
+#do some plotting
+plot(Data$Grid[[1]])
+points(Points,pch=19,cex=0.01,col='blue')
+
+data(alaska_dcw)
+data(russia_dcw)
+laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
+                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
+alaska_dcw <- spTransform(alaska_dcw, CRS(laea_180_proj))
+russia_dcw <- spTransform(russia_dcw, CRS(laea_180_proj))
+plot(alaska_dcw, col = "gray", add = TRUE)
+plot(russia_dcw, col = "gray", add = TRUE)
+                                                                     
+                                                                     
+
+##create a single grid object for this analysis, averaging time varying covariates over the dates considered
+Cur.grid=Data$Grid[[1]]
+n.cells=dim(Cur.grid)[1]
+S=nrow(Cur.grid)  #number of cells
+#first object corresponds to April 4 so start with 17th SpDF
+Mean.covs=matrix(0,S,3)
[…]

### 2014-08-25 SpatPred/inst/plot_IVH_simple.R

@@ -1,0 +1,154 @@
+## plot some examples of IVH
+library(grid)
+#linear
+set.seed(1234)
+X=matrix(0,20,2)
+X[,1]=1
+X[,2]=c(runif(10,1,1.5),runif(10,2.5,3))
+IVH = X%*%solve(crossprod(X))%*%t(X)
+max.IVH=max(diag(IVH))
+Point.df=data.frame(x=X[,2],y=rep(0.5,20))
+
+x0=c(0:399)/100
+X0=matrix(1,400,2)
+X0[,2]=x0
+X0.eval=X0%*%solve(crossprod(X))%*%t(X0)
+Extrap=which(diag(X0.eval)>max.IVH)
+Y=rep(1,400)
+Y[Extrap]=0
+plot(x0,Y)
+
+PredVar.df=data.frame(x=x0,pred.var=diag(X0%*%solve(crossprod(X))%*%t(X0)))
+
+library(ggplot2)
+my.theme=theme(plot.title = element_text(hjust = 0),axis.ticks = element_blank(), axis.text.y = element_blank(),axis.title.y=element_blank(),strip.text=element_text(face="bold"),axis.title=element_text(size=16),text=element_text(size=16),legend.text=element_text(size=16),legend.key.height=unit(1.1,"line"))
+
+#note the next line is hardwired with values obtained by examining elements of Extrap
+Rect.df=data.frame(x.min=c(0,2.98),x.max=c(1.01,4),y.min=c(0,0),y.max=c(1,1))
+linear.plot=ggplot(Rect.df)+geom_rect(aes(xmin=x.min,xmax=x.max,ymin=y.min,ymax=y.max),fill='gray')
+linear.plot=linear.plot+my.theme+geom_point(data=Point.df,aes(x=x,y=y),size=3,shape=4,colour='black')
+linear.plot=linear.plot+geom_line(data=PredVar.df,aes(x=x,y=pred.var),size=1)
+linear.plot=linear.plot+geom_hline(yintercept=max.IVH,linetype=2)+ggtitle("A.")
+linear.plot
+
+
+#quadratic example
+X2=matrix(0,20,3)
+X2[,c(1,2)]=X
+X2[,3]=X2[,2]^2
+
+IVH = X2%*%solve(crossprod(X2))%*%t(X2)
[…]

### 2014-08-25 SpatPred/inst/plot_ribbon_IVH_by_tau.R

@@ -1,0 +1,183 @@
+source("./SpatPred/R/spat_pred.R")
+require(maptools)
+require(ggplot2)
+load('Ribbon_data.Rda')
+load('Knot_cell_distances.Rdata')
+K=K.data$K
+set.seed(12345)
+
+#also include quadratic term for ice_conc in dataset
+ice_conc2=(Cur.grid@data[["ice_conc"]])^2
+dist_edge2=(Cur.grid@data[["dist_edge"]])^2
+dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
+dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
+Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
+#attach Ribbon seal counts to data frame
+Count=Ribbon.count
+Cur.grid@data=cbind(Cur.grid@data,Count)
+
+#add in some zero counts in cells with no ice
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
+Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
+Mapping=c(Mapping,Which.no.ice)
+Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
+
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+
+#run glm model
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
+#formula=Count~1
+#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~dist_mainland+ice_conc+ice_conc2
+formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~ice_conc+ice_conc2
+#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
+#formula=Count~dist_shelf+dist_shelf2
[…]

### 2014-08-25 SpatPred/inst/run_abundance_sims.R

@@ -1,0 +1,164 @@
+# run_generic_sims.R
+# script to run generic spatio-temporal count data simulations
+source("./SpatPred/R/sim_data_abundance.R")
+set.seed(12345)
+n.sims=100 #number of simulations at each design point
+S=400
+t.steps=20
+n.transects=5
+line.width=0.1
+
+i.sim=TRUE
+if(i.sim){
+  for(isim in 1:n.sims){
+    #simulate covariates, abundance
+    Grid=sim_data_generic(S,n.transects,n.covs=4,line.width=,tau.epsilon=20)
+    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    save(Grid,file=Cur.file)
+  }
+}
+
+
+#now simulate 
+for(isim in 1:n.sims){
+  Cur.file=paste("./Sim_data/Sim",isim,".Rda",sep='')
+  load(Cur.file)
+  
+}
+
+#s
+
+
+#call estimation routines
+Est.mods=c("CPIF","AST","STPC","OPRS")
+Sim.ind.20= c(1:19)*5
+for(igen in 1:3){  #loop over generating model to generate data sets
+  for(itrans in 1:1){ #loop over number of transects in each cell
+    for(isim in 1:(n.sims/2)){  
+
+      if(igen==3){
+        fname=paste("simdata_gen",Model.list[igen-1],"_trans",N.transects[itrans],"_sim",isim,sep='')
+        load(paste("./sim_generic_data/",fname,sep=''))
+        Which.distances=Sim.data$Data$Which.distances  
+      }
+      
+      fname=paste("simdata_gen",Model.list[igen],"_trans",N.transects[itrans],"_sim",isim,sep='')
+      load(paste("./sim_generic_data/",fname,sep=''))
+      Data=Sim.data$Data
+      if(igen==3)Data$Which.distances=Which.distances
+      
+      n.knots=length(Data$Knot.locations)
+      #calculate kernel densities at grid cell centroids 
+      Cell.centroids=gCentroid(Data$Grid[[1]],byid=TRUE)
+      Distances=gDistance(Data$Knot.locations,Cell.centroids,byid=TRUE)
+      K=matrix(dnorm(Distances,0,5),S,n.knots)  #knot sd=5 
[…]

### 2014-08-25 SpatPred/inst/run_ribbon_models.R

@@ -1,0 +1,415 @@
+source("./SpatPred/R/spat_pred.R")
+load('Ribbon_data.Rda')
+load('Knot_cell_distances.Rdata')
+K=K.data$K
+set.seed(12345)
+
+#also include quadratic term for ice_conc in dataset
+ice_conc2=(Cur.grid@data[["ice_conc"]])^2
+dist_edge2=(Cur.grid@data[["dist_edge"]])^2
+dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
+dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
+Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
+#attach Ribbon seal counts to data frame
+Count=Ribbon.count
+Cur.grid@data=cbind(Cur.grid@data,Count)
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+
+#add in some zero counts in cells with no ice
+Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
+Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
+#Mapping=c(Mapping,Which.no.ice)
+#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
+
+#run glm model
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
+#formula=Count~1
+#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~dist_mainland+ice_conc+ice_conc2
+formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~ice_conc+ice_conc2
+#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
+#formula=Count~dist_shelf+dist_shelf2
+#formula=Count~dist_edge+dist_edge2
[…]

### 2014-08-25 SpatPred/inst/run_ribbon_models_ice0.R

@@ -1,0 +1,415 @@
+source("./SpatPred/R/spat_pred.R")
+load('Ribbon_data.Rda')
+load('Knot_cell_distances.Rdata')
+K=K.data$K
+set.seed(12345)
+
+#also include quadratic term for ice_conc in dataset
+ice_conc2=(Cur.grid@data[["ice_conc"]])^2
+dist_edge2=(Cur.grid@data[["dist_edge"]])^2
+dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
+dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
+Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
+#attach Ribbon seal counts to data frame
+Count=Ribbon.count
+Cur.grid@data=cbind(Cur.grid@data,Count)
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+
+#add in some zero counts in cells with no ice
+Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
+Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
+Mapping=c(Mapping,Which.no.ice)
+Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
+
+#run glm model
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
+#formula=Count~1
+#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~dist_mainland+ice_conc+ice_conc2
+formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~ice_conc+ice_conc2
+#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
+#formula=Count~dist_shelf+dist_shelf2
+#formula=Count~dist_edge+dist_edge2
[…]

### 2014-08-25 SpatPred/inst/run_ribbon_models_naive.R

@@ -1,0 +1,425 @@
+source("./SpatPred/R/spat_pred.R")
+load('Ribbon_data.Rda')
+load('Knot_cell_distances.Rdata')
+K=K.data$K
+set.seed(12345)
+
+#also include quadratic term for ice_conc in dataset
+ice_conc2=(Cur.grid@data[["ice_conc"]])^2
+dist_edge2=(Cur.grid@data[["dist_edge"]])^2
+dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
+dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
+Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
+#attach Ribbon seal counts to data frame
+Count=Ribbon.count
+Cur.grid@data=cbind(Cur.grid@data,Count)
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+
+#add in some zero counts in cells with no ice
+Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
+Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
+#Mapping=c(Mapping,Which.no.ice)
+#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
+
+#run glm model
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
+#formula=Count~1
+#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~dist_mainland+ice_conc+ice_conc2
+formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~ice_conc+ice_conc2
+#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
+#formula=Count~dist_shelf+dist_shelf2
+#formula=Count~dist_edge+dist_edge2
[…]

### 2014-08-25 SpatPred/inst/run_ribbon_models_noRE.R

@@ -1,0 +1,457 @@
+source("./SpatPred/R/spat_pred.R")
+load('Ribbon_data.Rda')
+load('Knot_cell_distances.Rdata')
+K=K.data$K
+set.seed(12345)
+
+#also include quadratic term for ice_conc in dataset
+ice_conc2=(Cur.grid@data[["ice_conc"]])^2
+dist_edge2=(Cur.grid@data[["dist_edge"]])^2
+dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
+dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
+Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
+#attach Ribbon seal counts to data frame
+Count=Ribbon.count
+Cur.grid@data=cbind(Cur.grid@data,Count)
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+
+#add in some zero counts in cells with no ice
+Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
+Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
+#Mapping=c(Mapping,Which.no.ice)
+#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
+
+#run glm model
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
+#formula=Count~1
+#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~dist_mainland+ice_conc+ice_conc2
+formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~ice_conc+ice_conc2
+#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
+#formula=Count~dist_shelf+dist_shelf2
+#formula=Count~dist_edge+dist_edge2
[…]

### 2014-08-25 SpatPred/inst/run_spotted_models.R

@@ -1,0 +1,214 @@
+source("./SpatPred/R/spat_pred.R")
+load('Spotted_data.Rda')
+
+#also include quadratic term for ice_conc in dataset
+ice_conc2=(Cur.grid@data[["ice_conc"]])^2
+Cur.grid@data=cbind(Cur.grid@data,ice_conc2)
+#attach Ribbon seal counts to data frame
+Count=Spotted.count
+Cur.grid@data=cbind(Cur.grid@data,Count)
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+
+#add in some zero counts in cells with no ice
+Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
+Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
+Mapping=c(Mapping,Which.no.ice)
+Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
+
+#run glm model
+Control=list(iter=11000,burnin=1000,thin=2,adapt=TRUE,srr.tol=NULL,fix.tau.epsilon=FALSE,predict=TRUE,MH.nu=rep(0.2,length(Mapping)))
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+formula=Count~1
+
+Data=Cur.grid
+spat.mod=0
+Offset=Area.photo
+Area.adjust=Area.hab
+Assoc=NULL
+Knots=NULL
+Prior.pars=NULL
+#formula=Count~Ecoregion
+formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge+Ecoregion
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+glm.out <- spat_pred(formula=formula,Data=Data,spat.mod=spat.mod,Offset=Offset,Mapping=Mapping,Area.adjust=Area.adjust,Control=Control,Assoc=Assoc,Knots=Knots,Prior.pars=Prior.pars)
+Control=glm.out$Control
+Control$adapt=FALSE
+Control$iter=120000
+Control$burnin=20000
+Control$thin=50
[…]

### 2014-08-27 SSL_time_series/PCtimeMCMC.cpp
Added some code/figs for SSL example

@@ -1,0 +1,202 @@
+#include "RcppArmadillo.h"
+// [[Rcpp::depends(RcppArmadillo)]]
+using namespace Rcpp;
+using namespace arma;
+
+
+arma::vec armaNorm(int n){
+  NumericVector x = rnorm(n,0,1);
+  arma::vec out(x.begin(), x.size(), false);
+  return out;
+}
+
+arma::vec armaU(int n){
+  NumericVector x = runif(n,0,1);
+  arma::vec out(x.begin(), x.size(), false);
+  return out;
+}
+arma::vec GCN(const arma::mat& V, const arma::vec& v){
+  arma::mat out = solve(V,v) + solve(chol(V), armaNorm(v.n_elem));
+  return out;
+}
+
+arma::vec ln_Pois(const arma::vec& k, const arma::vec& ln_lambda){
+  return k%ln_lambda - exp(ln_lambda);
+}
+
+arma::vec ln_norm(const arma::vec& x, const arma::vec& mu, const arma::vec& sigma2_inv){
+  return -0.5*(x-mu)%(x-mu)%sigma2_inv;
+}
+
+
+// [[Rcpp::export]]
+List PCtimeMCMC(const arma::vec& y, const arma::mat& K_lf, const arma::mat& K_hf, const arma::vec& sampleIndex,
+const double& beta_mean, const double& beta_prec, const double& phi_lf_scale, const double& phi_hf_scale, const double& a_sigma,
+const double & b_sigma, const int& block, const int& burn, const int& iter, const bool& sample_sigma,
+const bool& sample_hf) {
+  //matrices for fitting
+  int na_lf = K_lf.n_cols;
+  int na_hf = K_hf.n_cols;
+  int ntot = K_lf.n_rows;
+  int ns = y.n_elem;
+  int nns = ntot-ns;
+  arma::uvec smp = find(sampleIndex==1);
+  arma::uvec not_smp = find(sampleIndex==0);
+  arma::vec X(ntot, fill::ones);
+  
+  // Storage matrices
+  arma::vec beta_store(iter, fill::zeros);
[…]

### 2014-08-27 SSL_time_series/run_ssl_demo.R

@@ -1,0 +1,82 @@
+###
+# Steller sea lion forecast extrapolation
+###
+
+library(agTrend)
+library(ggplot2)
+library(coda)
+library(Rcpp)
+Rcpp::sourceCpp('PCtimeMCMC.cpp')
+
+data(wdpsNonpups)
+
+mar_data = wdpsNonpups[wdpsNonpups$site=="MARMOT" & wdpsNonpups$year>1989,]
+
+###
+pred_window=20
+lf_freq=5
+hf_freq=5
+###
+
+knots_lf = seq(1990-lf_freq,2011+pred_window+lf_freq, by=lf_freq)
+knots_hf = seq(1990-hf_freq,2011+pred_window+hf_freq, by=hf_freq)
+
+K_lf = exp(-0.5*outer(1990:(2011+pred_window),knots_lf,"-")^2/lf_freq^2)
+K_lf=sweep(K_lf, 2, colSums(K_lf), "/")
+
+K_hf = exp(-0.5*outer(1990:(2011+pred_window),knots_hf,"-")^2/hf_freq^2)
+K_hf=sweep(K_hf, 2, colSums(K_hf), "/")
+
+
+post_smp=PCtimeMCMC(
+  y=mar_data$count, 
+  K_lf=K_lf, 
+  K_hf=0.0*K_hf, 
+  sampleIndex=as.numeric(1990:(2011+pred_window)%in%mar_data$year), 
+  beta_mean=0, 
+  beta_prec=0, 
+  phi_lf_scale=100, 
+  phi_hf_scale=100, 
+  a_sigma = 0,
+  b_sigma=0,
+  block=1000,
+  burn=10000, 
+  iter=1000000,
+  sample_sigma=TRUE,
+  sample_hf=FALSE
+  )
+
+# xxx = apply(post_smp$jump_idx, 2, function(x){cumsum(x)/c(1:length(x))})
+# apply(tail(post_smp$jump_idx, 10000), 2, mean)
+# plot(xxx[,2], type='l')
+# abline(h=0.234)
+# plot(post_smp$tune[,1], type='l')
+# 
+# plot(mcmc(post_smp$beta))
+# plot(mcmc(post_smp$alpha_f))
+# 
+# plot(mcmc(post_smp$phi_hf))
+
+
+pred_data = data.frame(year=1990:(2011+pred_window), pred=apply(post_smp$pred, 2, median), HPDinterval(mcmc(post_smp$pred), prob=0.9))
+pred_plot= ggplot() +
[…]

### 2014-11-24 manuscript/AppendixA.tex
Added ms text on simulation; now fully specify Gibbs samplers in Appendix A; started appendix B for simulation study details.
Some modifications to code for simulating data, including plotting code

--- 2014-08-25 manuscript/AppendixS1.tex
@@
 \renewcommand{\em}{\it}
-\renewcommand\thefigure{S1.\arabic{figure}}
+\renewcommand\thefigure{A\arabic{figure}}
+\renewcommand\thetable{A\arabic{table}}
+\renewcommand\theequation{A.\arabic{equation}}
 
@@
 \begin{center}
-Appendix S1: Model formulation and Gibbs sampling algorithms for certain classes and extensions of the generalized linear model
+Appendix A: Model formulation and Gibbs sampling algorithms for certain classes and extensions of the generalized linear model
 \bigskip
@@
 \end{linenomath*}
-where $\mu_i=\theta_i + \epsilon_i$, $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function,
-$\theta_i$ is a linear predictor, and $\epsilon_i \sim {\rm Normal}(0,\tau_\epsilon)$ is iid Gaussian error with precision parameter $\tau_\epsilon$.
+where $\mu_i=\xi_i + \epsilon_i$, $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function,
+$\xi_i$ is a linear predictor, and $\epsilon_i \sim {\rm Normal}(0,\tau_\epsilon)$ is iid Gaussian error with precision parameter $\tau_\epsilon$.
 The specification in Eq.
@@
 \begin{equation*}
-  Y_i \sim \mathrm{Poisson}(\exp(\theta_i + \epsilon_i)),
+  Y_i \sim \mathrm{Poisson}(\exp(\xi_i + \epsilon_i)),
 \end{equation*}
@@
 a configuration known as a log-Gaussian Cox process.
-We now describe how different classes of statistical models can be developed depending on how one structures the linear predictor, $\theta_i$.
+We now describe how different classes of statistical models can be developed depending on how one structures the linear predictor, $\xi_i$.
 
@@
 used by ecologists to make spatial predictions. In GLMs, the linear predictor is a simple linear function of gathered covariates
-(including possible quadratic terms of these covariates.  Statisticians often describe this relationship by
-$\boldsymbol{\theta}={\bf X} \boldsymbol{\beta}$, which is written in matrix notation.  In particular, $\boldsymbol{\theta}$ gives
+(including possible polynomial terms and interactions).  Statisticians often describe this relationship by
+$\boldsymbol{\xi}={\bf X} \boldsymbol{\beta}$, which is written in matrix notation.  In particular, $\boldsymbol{\xi}$ gives
 a vector of linear predictor values (one for each data point being analyzed), $\boldsymbol{\beta}$ gives a vector of regression coefficients, and ${\bf X}$ is a design matrix which includes all explanatory variables (and often a column vector of ones to represent an intercept).
@@
 
-There are a number of ways smooth relationships can be modeled; in order to formulate a GAM in the notation of Eq.
-\ref{eq:basic_model} we employ a knot-based kernel smoother with a radial basis function.
+\hspace{.5in}There are a number of ways smooth relationships can be modeled; in order to formulate a GAM in the notation of Eq.
+\ref{eq:basic_model} we employ a knot-based kernel smoother with a Gaussian radial basis function.
 The basic notion is to place $k_j$ knots, which we will denote by $\boldsymbol{\omega}_j = \{ \omega_{1j},\omega_{2j},\hdots,\omega_{k_j j} \}$, throughout the range of a given regressor $j$, where the number of knots controls the level of smoothing.
@@
   \label{eq:GAM}
-  \boldsymbol{\theta}={\bf X} \boldsymbol{\beta} + {\bf K}_j \boldsymbol{\alpha}_j
+  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + {\bf K}_j \boldsymbol{\alpha}_j
 \end{equation}
@@
 
-Our main point in writing the GAM as in Eq. \ref{eq:GAM} is to emphasize the structural similarity between GLMs and certain classes
+\hspace{.5in}Our main point in writing the GAM as in Eq. \ref{eq:GAM} is to emphasize the structural similarity between GLMs and certain classes
 of GAMs.  In particular,
 we can rewrite Eq. \ref{eq:GAM} as
-$\boldsymbol{\theta}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
+$\boldsymbol{\xi}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
 where ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K}_j \right]$ (i.e., concatenating ${\bf X}$ and ${\bf K}_j$ horizontally), and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha}_j \right] ^\prime$ (the subscript $aug$ denotes augmentation).
@@
   \label{eq:spat_reg}
-  \boldsymbol{\theta}={\bf X} \boldsymbol{\beta} + \boldsymbol{\eta},
+  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + \boldsymbol{\eta},
 \end{equation}
@@
   \label{eq:PC}
-  \boldsymbol{\theta}={\bf X} \boldsymbol{\beta} + {\bf K} \boldsymbol{\alpha}
+  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + {\bf K} \boldsymbol{\alpha}
 \end{equation}
@@
 Similarly, we can rearrange terms to write it similar to the GLM, such that
-$\boldsymbol{\theta}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
+$\boldsymbol{\xi}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
 where in this case ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K} \right]$, and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha} \right] ^\prime$.
@@
 
+\section{Gibbs sampling}
+
+We now describe methods for conducting Bayesian analysis for the models described above.
+We utilize Gibbs sampling, characterized by cyclically sampling groups of parameters according to their so-called full conditional distributions conditionally on data and other parameters \citep{GelmanEtAl2004}.
+In some cases, we are able to solve for full conditional distributions analytically and to sample from known families of probability distributions; in others, we resort to Metropolis-Hastings steps.
+
+\subsection{Gibbs sampling for the GLM model}
+
+Our Gibbs sampler for the GLM model consists of three sets of parameter updates (1) regression parameters ($\boldsymbol{\beta}$), (2) error precision ($\tau_\epsilon$), and (3) latent log-density ($\boldsymbol{\mu}$).
+
+\underline{Sampling from $[\boldsymbol{\beta}|\cdot]$}
+
+Setting the prior for the suite of regression parameters as $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$ results in the full conditional distribution
+\begin{linenomath*}
+\begin{equation*}
+   [\boldsymbol{\beta} | \cdot] \equiv \mathcal{N}(({\bf X}^\prime {\bf X})^{-1}{\bf X}^\prime \boldsymbol{\mu},({\bf X}^\prime {\bf X})^{-1} (\tau_\epsilon + \tau_\beta)^{-1}).
+\end{equation*}
+\end{linenomath*}
+We set $\tau_\beta = 0.01$ for applications described in this paper.
+
+\underline{Sampling from $[\boldsymbol{\tau_\epsilon}|\cdot]$}
+
+Specifying a conjugate $\text{Gamma}(a,b)$ prior for the precision parameter leads to a full conditional that is also Gamma distributed:
+\begin{eqnarray}
+  [\tau_\epsilon | \cdot] & \equiv & \textrm{Gamma}(0.5n + a,0.5 \boldsymbol{\Delta}^\prime \boldsymbol{\Delta} +b),
+  \label{eq:tau_eps}
+\end{eqnarray}
+where $\Delta = \boldsymbol{\mu}-{\bf X}\boldsymbol{\beta}$.
+For applications in this paper we set $a=1.0$ and $b=0.01$.
+This selection admits a large range of permissable parameter values while maintaining flat probability mass near the origin.
+
+\underline{Sampling from $[\boldsymbol{\mu}|\cdot]$}
+
+Although $\boldsymbol{\mu}$ could potentially be integrated out of the likelihood, its inclusion allows for easier computation of full conditionals for $\boldsymbol{\beta}$ and $\tau_\epsilon$ (and for other full conditionals once the model is expanded, e.g.
+for GAMs and STRMs).
+During the main MCMC phase, the $\boldsymbol{\mu}$ only need to be updated for grid cells that are exposed to sampling (see a subsequent section for information on posterior prediction for details on unsampled areas).
+The full conditional for $\mu_s$ in a given grid cell $s$ is given by
+\begin{equation*}
+  [\mu_{s,t} | \cdot] \propto \mathcal{N}(\mu_s;{\bf X}_s \boldsymbol{\beta},\tau_\epsilon^{-0.5}) \times \textrm{Poisson}(C_s,\exp(o_s+\mu_s)),
+\end{equation*}
+where ${\bf X}_s$ gives the $s$th row of the design matrix and $o_s$ is an offset.  For our examples, quadrats are configured to represent $10\%$ of a grid cell, cell $o_s = \log(0.1)$.
+We use a Metropolis-Hastings step to sample from $[\mu_{s,t} | \cdot]$.
+
+\subsection{Gibbs sampling for the GAM model}
+
+We used a Gibbs sampler for the GAM model that looks very familiar to that for the GLM model.
+
+\underline{Sampling from $[\boldsymbol{\tau_\epsilon}|\cdot]$}
+
+Assuming a Gamma($a,b$) prior, the full conditional for $\boldsymbol{\tau_\epsilon}$ is once again given by Eq. \ref{eq:tau_eps}.  In this case,
+$\Delta = \boldsymbol{\mu}-{\bf X}\boldsymbol{\beta}-{\bf K}\boldsymbol{\alpha}$.
+
+\underline{Sampling from $[\boldsymbol{\mu}|\cdot]$}
+
+The full conditional for $\mu_s$ in a given grid cell $s$ is given by
+\begin{equation*}
+  [\mu_{s,t} | \cdot] \propto \mathcal{N}(\mu_s;{\bf X}\boldsymbol{\beta}+{\bf K}_s \boldsymbol{\alpha},\tau_\epsilon^{-0.5}) \times \textrm{Poisson}(C_s,\exp(o_s+\mu_s)),
+\end{equation*}
+where ${\bf K}_s$ gives the $s$th row of ${\bf K}$.
+
+\underline{Sampling from $[\boldsymbol{\alpha}|\cdot]$}
+
+We impose a Gaussian prior on smoothing kernel weights, $\boldsymbol{\alpha} \sim \mathcal{N}(0,\tau_\alpha^{-1})$.
+Like $\boldsymbol{\beta}$, the resulting full conditional is once again in closed form:
+\begin{linenomath*}
+\begin{equation*}
+   [\boldsymbol{\alpha} | \cdot] \equiv \mathcal{MVN}(({\bf K}^\prime {\bf K})^{-1}{\bf K}^\prime (\boldsymbol{\mu}-{\bf X}\boldsymbol{\beta}),({\bf K}^\prime {\bf K}^{-1} (\tau_\epsilon + \tau_\alpha)^{-1}).
+\end{equation*}
+\end{linenomath*}
+We set $\tau_\alpha = 0.01$ in all subsequent applications.
+
+\subsection{Gibbs sampler for the PC STRM model}
+
+The Gibbs sampler for the process convolution version of the STRM model consists of the following:
+
+\underline{Sampling from $[\boldsymbol{\beta}|\cdot]$}
+
+Using the same prior as for the GLM results in the full conditional distribution
+\begin{linenomath*}
+\begin{equation*}
+   [\boldsymbol{\beta} | \cdot] \equiv \mathcal{N}(({\bf X}^\prime {\bf X})^{-1}{\bf X}^\prime (\boldsymbol{\mu}-{\bf K}\boldsymbol{\alpha}),({\bf X}^\prime {\bf X})^{-1} (\tau_\epsilon + \tau_\beta)^{-1}).
+\end{equation*}
+\end{linenomath*}
+
+\underline{Sampling from $[\boldsymbol{\mu}|\cdot]$}
+
+The full conditional for $\mu_s$ in a given grid cell $s$ is given by
+\begin{equation*}
+  [\mu_{s,t} | \cdot] \propto \mathcal{N}(\mu_s;{\bf X}_s \boldsymbol{\beta}+{\bf K}_s \boldsymbol{\alpha},\tau_\epsilon^{-0.5}) \times \textrm{Poisson}(C_s,\exp(o_s+\mu_s)),
+\end{equation*}
+where ${\bf K}_s$ gives the $s$th row of ${\bf K}$.
+
+\underline{Sampling from $[\tau_\epsilon|\cdot]$}
+
+The full conditional is the same as for the GAM model with the replacement $\Delta = \boldsymbol{\mu}-{\bf X}\boldsymbol{\beta}-{\bf K}\boldsymbol{\alpha}$.
+
+\underline{Sampling from $[\boldsymbol{\alpha}|\cdot]$}
+
+We impose the same Gaussian prior on smoothing kernel weights, $\boldsymbol{\alpha} \sim \mathcal{N}(0,\tau_\alpha^{-1})$ as for the GAM.  The resulting full conditional is simply
+\begin{linenomath*}
+\begin{equation*}
+   [\boldsymbol{\alpha} | \cdot] \equiv \mathcal{MVN}(({\bf K}^\prime {\bf K})^{-1}{\bf K}^\prime (\boldsymbol{\mu}-{\bf X}\boldsymbol{\beta}),({\bf K}^\prime {\bf K}^{-1} (\tau_\epsilon + \tau_\alpha)^{-1}).
+\end{equation*}
+\end{linenomath*}
+
+\subsection{Gibbs sampler for the RSR STRM model}
+
+Although ${\bf K}$ is defined differently in the RSR model, the Gibbs sampler remains the same as for the the PC model, with the exception of a modified full conditional for $[\boldsymbol{\alpha}|\cdot]$ owing to a different prior specification for $\boldsymbol{\alpha}$.
+We must also update an additional precision parameter associated with spatial random effects, $\tau_\eta$.
+
+\underline{Sampling from $[\boldsymbol{\alpha}|\cdot]$}
+
+In the RSR model, the implied prior for reduced dimension spatial effects is $\boldsymbol{\alpha} \sim \text{MVN}(0,\tau_\eta^{-1}{\bf K}^\prime {\bf Q} {\bf K})$.  The resulting full conditional is
+\begin{linenomath*}
+\begin{eqnarray*}
+   [\boldsymbol{\alpha} | \cdot] & = & \mathcal{MVN}({\bf M},\boldsymbol{\Sigma}), \text{ where} \\
+   \boldsymbol{\Sigma}^{-1} & = & {\bf K}^\prime {\bf K}\tau_\epsilon+\tau_\eta{\bf K}^\prime {\bf Q} {\bf K} \text{ and} \\
+    {\bf M} & = & \boldsymbol{\Sigma} \tau_\epsilon {\bf K}^\prime (\boldsymbol{\mu} - {\bf X} \boldsymbol{\beta}).
+\end{eqnarray*}
+\end{linenomath*}
+See for example, \citet{ConnEtAl2014}.
+
+\underline{Sampling from $\tau_\eta|\cdot]$}
+
+Using a conjugate Gamma$(a,b)$ prior on $\tau_\eta$, we have
+  \begin{equation*}
+  \tau_{\eta} \sim {\rm Gamma}(0.5m + a,0.5 \boldsymbol{\alpha}^\prime
+  {\bf K}^\prime {\bf Q} {\bf K}\boldsymbol{\alpha}+b).
+  \end{equation*}
+  Here, $m$ is the number of $\boldsymbol{\alpha}$ parameters.\\
+
+\subsection{Generating posterior predictions}
+If all we wanted to do was to estimate regression coefficients, the previous samplers would suffice.
+However, we typically also wish to predict animal abundance across the landscape to come up with a total abundance estimate and produce a density map.
+Fortunately, this is easy to do using posterior prediction, which consists of the following steps:
+\begin{enumerate}
+\item For each of $n$ samples (where $n$ is the desired posterior predictive sample size), choose an iteration of the MCMC sampler randomly with replacement (call this iteration $t$).
+\item Label the $t$th sample of $\boldsymbol{\alpha}$, $\boldsymbol{\beta}$, $\boldsymbol{\tau_\epsilon}$, and $\boldsymbol{\mu}$ as $\boldsymbol{\alpha}^{(t)}$, $\boldsymbol{\beta}^{(t)}$, $\boldsymbol{\tau_\epsilon}^{(t)}$, and $\boldsymbol{\mu}^{(t)}$, respectively.
+Recall that $\boldsymbol{\mu}^{(t)}$ is only available for sampled grid cells.
+\item For unsampled grid cells, generate $\boldsymbol{\mu}^{(t)} \sim \mathcal{N}({\bf X}\boldsymbol{\beta}^{(t)}+{\bf K}\boldsymbol{\alpha}^{(t)},\tau_\epsilon^{(t)})$
+\item Sample ${\bf N}^{(t)} \sim \text{Poisson}(\boldsymbol{\mu}^{(t)})$, where ${\bf N}^{(t)}$ is a vector of posterior predictions across the landscape.
+\item Compute a posterior prediction for total abundance as $N^{(t)}=\sum_s N_s^{(t)}$.
+\end{enumerate}
+
 \renewcommand{\refname}{Literature Cited}

### 2014-11-24 manuscript/AppendixB.tex
Added ms text on simulation; now fully specify Gibbs samplers in Appendix A; started appendix B for simulation study details.
Some modifications to code for simulating data, including plotting code

\documentclass[12pt,fleqn]{article}
%\documentclass[12pt,a4paper]{article}
\usepackage{natbib}
\usepackage{lineno}
%\usepackage{lscape}
%\usepackage{rotating}
%\usepackage{rotcapt, rotate}
\usepackage{amsmath,epsfig,epsf,psfrag}
\usepackage{setspace}
\usepackage{ulem}
\usepackage{xcolor}
\usepackage[labelfont=bf,labelsep=period]{caption} %for making figure and table numbers bold
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\hypersetup{pdfpagemode=UseNone}

%\usepackage{a4wide,amsmath,epsfig,epsf,psfrag}

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bX{{\ensuremath\mathbf{X}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\renewcommand{\refname}{\hspace{2.3in} \normalfont \normalsize LITERATURE CITED}
%this tells it to put 'Literature Cited' instead of 'References'
\bibpunct{(}{)}{,}{a}{}{;}
\oddsidemargin 0.0in
\evensidemargin 0.0in
\textwidth 6.5in
\headheight 0.0in
\topmargin 0.0in
\textheight=9.0in
%\renewcommand{\tablename}{\textbf{Table}}
%\renewcommand{\figurename}{\textbf{Figure}}
\renewcommand{\em}{\it}
\renewcommand\thefigure{B\arabic{figure}}
\renewcommand\thetable{B\arabic{table}}
\renewcommand\theequation{B.\arabic{equation}}

\begin{document}

\begin{center} \bf {\large Avoiding extrapolation bias when using statistical models to make ecological prediction}

\vspace{0.7cm}
Paul B. Conn$^{1*}$, Devin S. Johnson$^1$, and Peter L. Boveng$^1$
\end{center}
\vspace{0.5cm}

\rm
\small

\it $^1$National Marine Mammal Laboratory, Alaska Fisheries Science Center,
NOAA National Marine Fisheries Service,
Seattle, Washington 98115 U.S.A.\\

\rm \begin{flushleft}

\raggedbottom
\vspace{.5in}

\begin{center}
Appendix B: Full details of simulation study examining predictive extrapolation
\bigskip
\end{center}
\vspace{.3in}

\doublespacing

We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where extrapolations from statistical species distribution models may be problematic.  Each simulation consisted of several steps, including
\begin{enumerate}
  \item Simulate four hypothetical environmental covariates over a 30 by 30 grid,
  \item Simulate animal abundance across the landscape as a function of covariates,
  \item Simulate the position of animal count quadrats across the landscape (i.e. locations where sampling is to occur) and animal counts on each quadrat,
  \item Estimate animal abundance as a function of environmental covariates according to each of three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).  For the latter, only a spatial dimension was considered.  All such models were specified hierarchically, with MCMC used for posterior simulation.
  \item Calculation of the gIVH as determined by (1) integrating over prior distributions for unknown parameters ($\text{gIVH}_{int}$), or (2) using realized posterior variance (i.e. after data were collected and analyzed; $\text{gIVH}_{post}$).
\end{enumerate}
We now describe each of these tasks and results of the simulation study in further detail.  All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the supplement.

\underline{1. Simulating environmental covariates} \\
For each simulation, we generated environmental covariates by discretizing the realizations of a Matern Cluster Process (Fig. \ref{fig:Matern}) using the rMatClust function within the R \texttt{spatstat} package \citep{BaddeleyTurner2005}.  The Matern Cluster Process is a point process that includes a total of $\kappa$ cluster centers.  Within a radius $r$ for each cluster center, a total of $n$ points are randomly distributed, where $n \sim \text{Poisson}(\lambda)$ (e.g. Fig. \ref{fig:Matern}).  For each simulated covariate, we drew $\kappa \sim U(5,10)$, $r \sim U(0.2,0.4)$, and $\lambda \sim U(200,500)$ to provide variation in the relative density and clustering for each covariate.  Next, we overlayed a 30 by 30 grid over simulated points.  To generate a hypothetical covariate, we then summed the number of points within each grid cell, and standardized so that the maximum covariate value was 1.0 (e.g. \ref{fig:Matern}).

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{MaternCov.pdf}
\end{center}
\caption{Environmental covariates were generated via a discretized Matern Cluster Process.  First, a Matern Cluster Process (A.) is simulated.  Next, the number of simulated points in each grid cell is counted and standardized to have a maximum of 1.0 (B.).} 
\label{fig:Matern}
\end{figure}

\underline{2. Simulating animal abundance} \\

Given values of the four simulated covariates, we generated a vector of expected log-abundance in each cell ($\boldsymbol{\mu}$) as
\begin{linenomath}
\begin{equation*}
  $\boldsymbol{\mu}$ = \beta_0 + {\bf X}\boldsymbol{\beta} + \boldsymbol{\epsilon},
\end{equation*}
\end{linenomath}
where the intercept, $\beta_0$, was drawn from a $\mathcal{N}(2.5,0.25)$ distribution, the other regression coefficients ($\boldsymbol{\beta}$) were each drawn from a $\mathcal{N}(0,0.8)$ distribution, and residual Gaussian errors ($\boldsymbol{\epsilon}$) were drawn from a $\mathcal{N}(0,0.1)$ distribution.  The design matrix ${\bf X}$ was constructed assuming linear and quadratic effects for each covariate, together with all one-way interactions for a total of 14 regression coefficients in addition to the intercept.
Next, abundance in each cell $i$ was generated as  
\begin{linenomath}
\begin{equation*}
  N_i \sim \text{Poisson}(\exp(\mu_i)).
\end{equation*}
\end{linenomath}

\underline{3. Simulating sample locations and count data} \\

For each simulated landscape, we selected 60 grid cells at random (without replacement) for sampling.  We configured simulations such that sample quadrats covered 10\% of the target grid cell, and generated animal counts for each quadrat as
\begin{linenomath}
\begin{equation*}
  C_i \sim \text{Binomial}(N_i,0.1).
\end{equation*}
\end{linenomath}

\underline{4. Estimating animal abundance} \\

For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models.  The models were built to resemble a GLM, a GAM, and an STRM.

\renewcommand{\refname}{Literature Cited}
\bibliographystyle{ecology}

\bibliography{master_bib}

\end{flushleft}
\end{document}

### 2014-11-24 manuscript/SpatPred3.tex
Added ms text on simulation; now fully specify Gibbs samplers in Appendix A; started appendix B for simulation study details.
Some modifications to code for simulating data, including plotting code

@@
 
-{\em Key words: Abundance, Extrapolation, Forecasting, General additive models, Generalized linear models, Independent Variable Hull, Leverage, Occupancy, Spatial regression}
+{\em Key words: Abundance, Extrapolation, General additive models, Generalized linear models, Independent Variable Hull, Leverage, Prediction variance, Spatial regression, Species distribution model}
 
@@
 In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers \citep{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points.
-Application of the gIVH can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction.
+Application of the gIVH and related criterion (e.g.
+prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction.
 Also, since the gIVH can be constructed solely with knowledge of sampled locations and explanatory covariates (i.e., it does not necessarily require any observed response variables), it can also be used to help guide survey design.
@@
 
-Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes iid normally distributed error.
-Thus, it is not directly applicable to generalized models, such as those including spatial random effects.
+Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes iid Gaussian error.
+Thus, it is not directly applicable to generalized models, such as those including alternative error distributions (e.g., Poisson, binomial) or spatial random effects.
 Further, the hat matrix is not necessarily well defined for models with more general spatial structure.
 However, since the hat matrix is proportional to prediction variance, Cook (1979) notes that design points with maximum prediction variance will be located on the boundary of the IVH.
-Working on the linear predictor scale, we therefore define a generalized independent variable hull (gIVH) as the set of all points ${\bf x}$ (note that ${\bf x}$ can include both observed and unobserved design points) such that
+We therefore define a generalized independent variable hull (gIVH) as the set of all points ${\bf x}$ (note that ${\bf x}$ can include both observed and unobserved design points) such that
 \begin{linenomath*}
@@
   \label{eq:gIVH}
-  {\rm var}(\boldsymbol{\mu}_x | {\bf x}) \le {\rm max} [{\rm var}(\boldsymbol{\mu}_X | {\bf X})],
+  {\rm var}({\bf y}_x | {\bf x}) \le {\rm max} [{\rm var}({\bf y}_X | {\bf X})],
 \end{equation}
 \end{linenomath*}
-where $\boldsymbol{\mu}_x$ correspond to predictions at ${\bf x}$, and ${\bf X}$ and $\boldsymbol{\mu}_X$ denote observed design points and predictions at $X$, respectively.
+where ${\bf Y}_x$ correspond to predictions at ${\bf x}$, and ${\bf X}$ and ${\bf Y}_X$ denote observed design points and predictions at $X$, respectively.
 
@@
 \begin{equation}
-  Y_i \sim f_Y(g^{-1}(\mu_i)),
+  Y_i \sim f_Y(g(\mu_i)),
 \end{equation}
 \end{linenomath*}
-where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function, and
+where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives an inverse link function, and
 $\mu_i$ is a predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
@@
 where the $\boldsymbol{\epsilon}$ represent Gaussian error, ${\bf X}_{aug}$ denotes an augmented design matrix, and $\boldsymbol{\beta}_{aug}$ denote an augmented vector of parameters.
-For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see Appendix S1 for examples of how numerous types of models can be written in this form).
+For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see Appendix A for examples of how numerous types of models can be written in this form).
 
@@
 \end{linenomath*}
-where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; Appendix S1).
-For GLMs, prediction variance is proportional to the original hat matrix ${\bf V}_{LR}$ (at least on the linear predictor scale), and Eq.
+where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; Appendix A).
+This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method \citep{Dorfman1938,VerHoef2012} to approximate prediction variance on the real scale (i.e.
+the scale in which data are measured) as
+\begin{linenomath*}
+\begin{eqnarray*}
+  \label{eq:var-mu}
+  {\rm var}(g(\mu_x) | {\bf x}) & = & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x | {\bf x})\boldsymbol{\Delta}^\prime, 
+%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
+%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
+%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
+%      \vdots & \vdots & \ddots & \vdots \\
+%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
+%  \end{array} \right].
+\end{eqnarray*}
+\end{linenomath*}
+where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).
+Under common univariate link functions (e.g.
+log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g.
+multinomial logit) $\boldsymbol{\Delta}$ will be dense.
+
+%For GLMs, prediction variance is proportional to the original hat matrix ${\bf V}_{LR}$ (at %least on the linear predictor scale), and Eq.
 \ref{eq:IVH} may be used directly.
-For other models, we have
-\begin{linenomath*}
-\begin{equation}
-  \label{eq:var_Beta}
-  {\rm var}(\hat{\boldsymbol{\beta}}_{aug})=\left[ \begin{array}{cc}
-      \boldsymbol{\Sigma}_\beta & {\bf 0}\\
-      {\bf 0} & \boldsymbol{\Sigma}_\alpha
-  \end{array} \right],
-\end{equation}
-\end{linenomath*}
-where we simply replace $\boldsymbol{\beta}_{aug}$ with the augmented vector that includes both regression and spatial or smooth parameters (i.e.
-$\boldsymbol{\beta}$ and $\boldsymbol{\alpha}$; Appendix S1).
+For other %models, we have
+%\begin{linenomath*}
+%\begin{equation}
+%  \label{eq:var_Beta}
+%  {\rm var}(\hat{\boldsymbol{\beta}}_{aug})=\left[ \begin{array}{cc}
+%      \boldsymbol{\Sigma}_\beta & {\bf 0}\\
+%      {\bf 0} & \boldsymbol{\Sigma}_\alpha
+%  \end{array} \right],
+%\end{equation}
+%\end{linenomath*}
+%where we simply replace $\boldsymbol{\beta}_{aug}$ with the augmented vector that includes %both regression and spatial or smooth parameters (i.e.
+$\boldsymbol{\beta}$ and %$\boldsymbol{\alpha}$; Appendix S1).
 
@@
 5 directly is quite useful, as we only need to know the explanatory variables to be able to diagnose whether predictions lie in or out of the gIVH.
-However, the other models considered here (GAMs and STRMs) typically require estimation of an additional precision parameter.
-In these cases, one solution is to impose prior distributions and integrate over precision parameters (call these $\boldsymbol{\tau}$) to obtain an expectation for ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$.
+However, the other models considered here (GAMs and STRMs) typically require estimation of one or more precision or autocorrelation parameters.
+In these cases, one solution is to impose prior distributions and integrate over these additional parameters (call them $\boldsymbol{\theta}$) to obtain an expectation for ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$.
 Using the law of the unconscious statistician, we have
@@
   \label{eq:E_var_Beta}
-  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}))  = \int_{\boldsymbol{\tau}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) [\boldsymbol{\tau}] d \boldsymbol{\tau}.
+  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}))  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
 \end{equation}
 \end{linenomath*}
-Here, and throughout the paper, we use the bracket notation to indicate a probability density function; e.g., $[\boldsymbol{\tau}] $ denotes the joint prior distribution for precision parameters $\boldsymbol{\tau}$.
+Here, and throughout the paper, we use the bracket notation to indicate a probability density function; e.g., $[\boldsymbol{\theta}] $ denotes the joint prior distribution for $\boldsymbol{\theta}$.
 If observations have already been gathered, one could also consider integrating over the marginal posterior distribution using MCMC:
@@
   \label{eq:E_var_Beta_post}
-  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}) | {\bf Y})  = \int_{\boldsymbol{\tau}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) [\boldsymbol{\tau}| {\bf Y}] d \boldsymbol{\tau}.
+  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}) | {\bf Y})  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}| {\bf Y}] d \boldsymbol{\theta}.
 \end{equation}
 \end{linenomath*}
+Some care must be taken with prior distributions if Eq.
+\ref{eq:E_var_Beta} is used for calculation of prediction variance.
+For instance, standard ``non-informative" or ``flat" priors may place substantial mass on implausible values.
+When using the gIVH or prediction variance to help guide sampling design \citep[see][for an example using prediction variance]{DiggleLophaven2006}, we suggest using informative prior distributions.
 
@@
 In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).
-To our knowledge, this is the first time this definition has been used in the context of spatial prediction, but one we shall show can be quite helpful in diagnosing and mitigating extrapolation bias.
-The gIVH seems ideally situated to this task as it does not necessarily need to rely on gathered response data.
-Thus, one can examine whether or not prediction points lie within the IVH without ever collecting response data there.
+The gIVH, together with prior-integrated prediction variance, seem ideally situated to diagnosing potential extrapolation issues as it does not necessarily need to rely on gathered response data.
+Thus, one can examine whether or not prediction points lie within the gIVH without ever collecting response data there.
+Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions are being based on.
 
@@
 
-We conducted simulation analyses to investigate whether the gIVH was useful in diagnosing prediction biases in occupancy and abundance analyses.
-In both cases, we abundance on a $30 \times 30$ grid, and generated as a function of four habitat covariates in addition to spatial process covariance (Appendix S2).
-The habitat covariates also
+We conducted a simulation study to investigate whether the gIVH (and accompanying prediction variance) was useful in diagnosing prediction biases when analyzing animal count data.
+For each of 100 simulations, we generated animal abundance over a $30 \times 30$ grid assuming that animal density was homogeneous in each grid cell.
+Animal abundance was generated as a function of four hypothetical habitat covariates (Appendix B).
+For each simulated landscape, we randomly selected 60 grid cells for sampling, assuming that sampling quadrats encompassed 10\% of the area of each selected grid cell.
+For ease of presentation and analysis, we assumed detection probability was 100\% in each quadrat.
+Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (Appendix B).
+The GLM and STRM expressed log-density as a function of linear and quadratic covariate effects, while the GAM used a kernel smoother with 4 knots for each covariate (Appendix B).
+Each model was provided with three of the four covariates used to generate the data.
+
+For each simulated data set and model structure, we calculated (1) $\text{gIVH}_{int}$, the gIVH resulting from integrating over prior parameter variance (i.e.
+using Eq.
+\ref{eq:E_var_Beta} with informative priors), and (2) $\text{gIVH}_{post}$, the gIVH arising from posterior predictive variance (i.e.
+posterior variance when data are actually analyzed).
+We then calculated posterior predictions of animal abundance within and outside of each gIVH in order to gauge bias as a function of whether or not inference is constrained to the gIVH.
+Specifically, the performance of the $\text{gIVH}_{int}$ criterion in diagnosing extrapolation bias is a test of whether it is useful in designing animal count surveys, while the performance of $\text{gIVH}_{post}$ may help decide its utility in limiting the scope of landscape-based inference once data have been collected and analyzed.
 
@@
 Using $x$ as a simple linear effect (A), only predictions less than the minimum observed value of $x$ or greater than the maximum value of $x$ are outside the IVH (shaded area), as scaled prediction variance in these areas (solid line) are greater than the maximum scaled prediction variance for observed data (dashed line).
-Using both linear and quadratic effects of $x$, some intermediate points are also outside the IVH; predictions at these points should also be viewed with caution.
+Using both linear and quadratic effects of $x$, some intermediate points are also outside the IVH; predictions at these points should also be viewed with some caution.
 Panels C $\&$ D show a more complicated IVH when the investigator wishes to relate an unmeasured response variable to linear and quadratic effects of two covariates, $x$ and $y$, either without interactions (C) or with interactions (D).
-Any potential predictions in the shaded area are outside of the IVH and should be viewed with caution.
+Any potential predictions in the shaded area are outside of the IVH.
 

### 2014-11-24 SpatPred/R/sim_data_abundance.R

@@ -7,12 +7,12 @@
 #' @export
 #' @keywords spatio-temporal, simulation
 #' @author Paul B. Conn
-sim_data_generic<-function(S,n.covs=4,tau.epsilon=20){
+sim_data_generic<-function(S,n.covs=4,tau.epsilon=10){
   require(ggplot2)
   require(Matrix)
   require(spatstat)
   
-  tau.epsilon=20
+  tau.epsilon=10
   
   source('./SpatPred/R/util_funcs.R')
    
@@ -24,8 +24,8 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=20){
   Covs=matrix(0,S,n.covs)
   for(icov in 1:n.covs){
     kappa=runif(1,5,10)
-    r=runif(1,0.1,.5)
-    mu=runif(1,500,1500)
+    r=runif(1,0.2,.4)
+    mu=runif(1,200,500)
     Dat.matern=rMatClust(kappa,r,mu)  
     X=round((x.len)*Dat.matern$x+0.5)
     Y=round((x.len)*Dat.matern$y+0.5)
@@ -36,7 +36,34 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=20){
     Grid=matrix(0,x.len,x.len)
     for(i in 1:length(X))Grid[X[i],Y[i]]=Grid[X[i],Y[i]]+1
     Covs[,icov]=Grid/max(as.vector(Grid))
-  }  
+  } 
+  
+  PLOT=FALSE
+  if(PLOT==TRUE)
+    set.seed(1111)
+    kappa=runif(1,5,10)
+    r=runif(1,0.2,.4)
+    mu=runif(1,200,500)
+    Dat.matern=rMatClust(kappa,r,mu)  
+    X=round((x.len)*Dat.matern$x+0.5)
+    Y=round((x.len)*Dat.matern$y+0.5)
+    X[X<1]=1
+    Y[Y<1]=1
+    X[X>x.len]=x.len
+    Y[Y>x.len]=x.len
+    Grid=matrix(0,x.len,x.len)
+    for(i in 1:length(X))Grid[X[i],Y[i]]=Grid[X[i],Y[i]]+1
+    Cov=Grid/max(as.vector(Grid))
+    XY=data.frame(x=Dat.matern$x,y=Dat.matern$y)
+    Grid=data.frame(y=rep(1:30,each=30),x=rep(c(1:30),30),cov=as.vector(Cov))
+    library(ggplot2)
[…]

### 2014-12-22 manuscript/SpatPred3.tex
Spatial prediction code & paper with gIVH, prediction variance using delta method.  Currently spatial gIVH programmed with RSR-ICAR model which doesn't work well because you can't limit K to just those cells that are sampled since Q is still full dimensional. Need to switch to process convolution for gIVH computation.

@@
 \begin{center}
-Avoiding extrapolation bias when using statistical models to make ecological prediction
+Avoiding extrapolation bias when using statistical models to make ecological predictions
 \bigskip\\
@@
 Also, since the gIVH can be constructed solely with knowledge of sampled locations and explanatory covariates (i.e., it does not necessarily require any observed response variables), it can also be used to help guide survey design.
-We illustrate use of the gIVH on a simulated occupancy dataset, on a species distribution model (SDM) for ribbon seals in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).
+We illustrate use of the gIVH on simulated count data, on a species distribution model (SDM) for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).
 
@@
   \label{eq:gIVH}
-  {\rm var}({\bf y}_x | {\bf x}) \le {\rm max} [{\rm var}({\bf y}_X | {\bf X})],
+  {\rm var}(\boldsymbol{\lambda}_x) \le {\rm max} [{\rm var}(\boldsymbol{\lambda}_X)],
 \end{equation}
 \end{linenomath*}
-where ${\bf Y}_x$ correspond to predictions at ${\bf x}$, and ${\bf X}$ and ${\bf Y}_X$ denote observed design points and predictions at $X$, respectively.
+where $\boldsymbol{\lambda}_x$ correspond to mean predictions at ${\bf x}$, and ${\bf X}$ and $\boldsymbol{\lambda}_X$ denote observed design points and predictions at $X$, respectively.
 
@@
 where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives an inverse link function, and
-$\mu_i$ is a predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
+$\mu_i$ is a linear predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
 \begin{linenomath*}
@@
   \label{eq:var-mu}
-  {\rm var}(\boldsymbol{\mu}_x | {\bf x})= {\bf x} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) {\bf x}^\prime,
+  {\rm var}(\boldsymbol{\mu}_x)= {\bf x} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) {\bf x}^\prime,
 \end{equation}
@@
 \begin{linenomath*}
-\begin{eqnarray*}
-  \label{eq:var-mu}
-  {\rm var}(g(\mu_x) | {\bf x}) & = & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x | {\bf x})\boldsymbol{\Delta}^\prime, 
+\begin{eqnarray}
+  \label{eq:var-lambda}
+  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
 %  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
@@
 %  \end{array} \right].
-\end{eqnarray*}
+\end{eqnarray}
 \end{linenomath*}
@@
 
-%For GLMs, prediction variance is proportional to the original hat matrix ${\bf V}_{LR}$ (at %least on the linear predictor scale), and Eq.
-\ref{eq:IVH} may be used directly.
-For other %models, we have
-%\begin{linenomath*}
-%\begin{equation}
-%  \label{eq:var_Beta}
-%  {\rm var}(\hat{\boldsymbol{\beta}}_{aug})=\left[ \begin{array}{cc}
-%      \boldsymbol{\Sigma}_\beta & {\bf 0}\\
-%      {\bf 0} & \boldsymbol{\Sigma}_\alpha
-%  \end{array} \right],
-%\end{equation}
-%\end{linenomath*}
-%where we simply replace $\boldsymbol{\beta}_{aug}$ with the augmented vector that includes %both regression and spatial or smooth parameters (i.e.
-$\boldsymbol{\beta}$ and %$\boldsymbol{\alpha}$; Appendix S1).
-
-The exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ differs depending on the underlying model structure and estimation procedure.
-In the following treatment, we shall focus on Bayesian analysis; although it is not necessarily needed to fit GLMs and GAMs, it helps make STRMs more computationally tractable and puts all of the different models into a common analysis framework.
-This approach requires specifying priors for model parameters, and prior parameters may appear in expressions for prediction variance.
-Judicious choices of priors can at times limit the influence of priors on calculation of the gIVH.
-For example, if we specify the prior on the vector of regression coefficients to be $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$ where $\tau_\beta$ is a fixed constant, a Bayesian implementation of a GLM still results in the gIVH specified in Eq.
-\ref{eq:IVH}.
-The ability to use Eq.
-5 directly is quite useful, as we only need to know the explanatory variables to be able to diagnose whether predictions lie in or out of the gIVH.
-However, the other models considered here (GAMs and STRMs) typically require estimation of one or more precision or autocorrelation parameters.
-In these cases, one solution is to impose prior distributions and integrate over these additional parameters (call them $\boldsymbol{\theta}$) to obtain an expectation for ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$.
-Using the law of the unconscious statistician, we have
+Unlike Cook's original formulation, the gIVH for nonidentity link functions often involve unknown parameters.
+For example, using Eq.
+\ref{eq:var-lambda} with a log link function means that diagonal entries of  $\boldsymbol{\Delta}$, $\exp{\boldsymbol{\mu}}$, include $\boldsymbol{\beta}$.
+Further, the exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ differs depending on the underlying model structure and estimation procedure.
+In the following treatment, we shall focus on Bayesian analysis; although it is not necessarily needed to fit GLMs and GAMs, it puts different models into a common analysis framework and allows us to specify informative probability distributions for unknown parameters.
+Specifically, our solution is to impose a probability distribution for unknown parameters, $[\boldsymbol{\theta}$, and to compute prediction variance as
 \begin{linenomath*}
 \begin{equation}
-  \label{eq:E_var_Beta}
-  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}))  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
+  \label{eq:E_var_lambda}
+  var(\boldsymbol{\lambda}_x) = \int_{\boldsymbol{\theta}} {\rm var}(\boldsymbol{\lambda}_x | \boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
 \end{equation}
 \end{linenomath*}
-Here, and throughout the paper, we use the bracket notation to indicate a probability density function; e.g., $[\boldsymbol{\theta}] $ denotes the joint prior distribution for $\boldsymbol{\theta}$.
-If observations have already been gathered, one could also consider integrating over the marginal posterior distribution using MCMC:
-\begin{linenomath*}
-\begin{equation}
-  \label{eq:E_var_Beta_post}
-  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}) | {\bf Y})  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}| {\bf Y}] d \boldsymbol{\theta}.
-\end{equation}
-\end{linenomath*}
+
+If data have already been collected or a pilot study has been conducted, a posterior distribution can be substituted for $[\boldsymbol{\theta}]$; otherwise, informative priors that encompass the likely range of model parameters can be considered.
+Judicious choices of priors can at times limit the influence of priors on calculation of the gIVH.
+For example, if we specify a Zellner's g-prior \citep{Zellner1986} on the vector of regression coefficients, $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$, then priors are automatically scaled to the level of the gathered covariates, and the only remaining task is to investigate sensitivity to $\tau_\beta$.
 Some care must be taken with prior distributions if Eq.
-\ref{eq:E_var_Beta} is used for calculation of prediction variance.
+\ref{eq:E_var_lambda} is used for calculation of prediction variance.
 For instance, standard ``non-informative" or ``flat" priors may place substantial mass on implausible values.
@@
 In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).
-The gIVH, together with prior-integrated prediction variance, seem ideally situated to diagnosing potential extrapolation issues as it does not necessarily need to rely on gathered response data.
-Thus, one can examine whether or not prediction points lie within the gIVH without ever collecting response data there.
+The gIVH, together with prior or posterior prediction variance, seem ideally situated to diagnosing potential extrapolation issues as it does not necessarily need to rely on gathered response data.
+Thus, it may prove useful when formulating survey designs since one can examine whether or not prediction points lie within the gIVH without ever collecting response data there.
 Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions are being based on.
@@
 
-\subsection{Simulation study}
+We used simulation to investigate the usefulness of the gIVH when (i) designing animal count surveys, and (ii) limiting the scope of inference when fitting models to count data that have already been collected.
+We also applied the gIVH to examining the robustness of a species distribution model (SDM) of ribbon seals in the eastern Bering Sea, and to time series projections of Steller sea lion population trends.
+
+\subsection{Simulation study I: study design}
+
+To investigate the utility of the gIVH in comparing alternative animal survey designs, we first generated animal abundance over a  $30 \times 30$ grid assuming that animal density was homogeneous in each grid cell.
+Animal abundance was generated as a function of four hypothetical spatially autocorrelated habitat variables (Appendix B).
+We systematically examined designs where abundance is modeled as a function of 1-3 of these covariates (representing a low to a high degree of understanding about animal habitat utilization and preference).
+Three possible sampling configurations were initially compared: (1) a spatially balanced sample of $n=50$ \citep[using a random tesselation design][]{StevensOlsen2004} (2) a convenience sample of $n=60$ observations where the probability of sampling was greater for cells closer to a ``base of operations" (reflecting the principle that greater sample sizes are usually possible if logistics are easier); and (3) a $n=50$ sample obtained by sampling proportionally to prediction variance associated with fixed effects (Fig.
+\ref{fig:designs}, Appendix B).
+
+\subsection{Simulation study II: Limiting scope of inference}
 
@@
 
-Fieberg PVA
+In the context of polynomial response surface design, \citet{BoxDraper1959} showed that when one wishes to reduce predictive variance, but also has uncertainty about the functional form of the response surface to be estimated, that the ``optimal" strategy in terms of bias and variance combined is often to concentrate on minimizing bias alone.
+For this purpose, a first order orthogonal design is required; in spatial applications without only easting and northing covariates, this would imply some type of systematic sample.
+``.
+.
+.
+conclusions obtained by minimizing variance error only and assuming the postulated model to be correct are likely to be wrong in many practical design situations." (Draper and Smith 1966, pp 18-19).
 

### 2014-12-22 SpatPred/R/gIVH.R

@@ -1,0 +1,69 @@
+### calculate prior prediction variance/gIVH via Monte Carlo
+
+gIVH <- function(Data,Effort,Assoc=NULL,my.formula,Tau.beta,Tau.eta,Tau.epsilon,srr.tol=0.8){
+  n.reps=1000
+  X.pred=model.matrix(my.formula,data=Data@data)
+  X.obs=model.matrix(my.formula,data=Data@data[Effort$Mapping,]) 
+  n.obs=length(Effort$Mapping)
+  XpXinv=solve(crossprod(X.obs))
+  n.beta=ncol(X.pred)
+  I.beta=diag(nrow(XpXinv))
+  
+  if(is.null(Assoc)==FALSE){
+    Q=-Assoc
+    diag(Q)=apply(Assoc,2,'sum')
+    Q=Matrix(Q)  
+    L.t=XpXinv
+    L=L.t
+    Qt=L.t
+    cross.L=L.t
+    Theta=L.t
+    P.c=diag(S)-X.pred%*%solve(crossprod(X.pred),t(X.pred))
+    Omega=Matrix((P.c%*%Assoc%*%P.c)*(S/sum(Assoc)))
+    Eigen=eigen(Omega)
+    if(srr.tol>1){ #in this case, the number of eigenvalues is preselected
+      L.t=Eigen$vectors[,1:srr.tol]      
+    }
+    else{ #in this case, compute the number of eigenvalues > the threshold
+      if(max(Eigen$values)<srr.tol)cat(paste("\n Error: maximum eigenvalue (",max(Eigen$values),") < Meta$srr.tol; decrease srr.tol"))
+      Ind=which(Eigen$values>srr.tol)
+      L.t=Eigen$vectors[,Ind]
+      cat(paste("\n",length(Ind)," eigenvectors selected for spatially restricted regression \n"))
+    }
+    L=t(L.t)
+    Qt=L%*%Q%*%L.t
+    #Qt=L[,Effort$Mapping]%*%Q[Effort$Mapping,Effort$Mapping]%*%L.t[Effort$Mapping,]  #NOTE: limiting rows of K to those locations observed here
+    cross.L=L%*%L.t
+    n.theta=nrow(Qt)
[…]

### 2014-12-22 SpatPred/R/sim_data_abundance.R

@@ -2,17 +2,19 @@
 #' @param S Number of cells (must be a square number)
 #' @param n.covs Number of habitat covariates (default 4)
 #' @param tau.epsilon Precision of exchangeable errors in log space
-#' @return A list object composed of at least two objects, "Data$Grid" is a list vector holding covariate data by day,
-#'         and "Count.data" which holds simulated transect counts 
+#' @return A SpatialPolygonsDataFrame holding covariates, abundance, and lambda (expected abundance)
 #' @export
 #' @keywords spatio-temporal, simulation
 #' @author Paul B. Conn
-sim_data_generic<-function(S,n.covs=4,tau.epsilon=10){
+sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
   require(ggplot2)
   require(Matrix)
   require(spatstat)
+  require(sp)
   
-  tau.epsilon=10
+  DEBUG=FALSE   
+  tau.epsilon=100
+  #if(DEBUG)tau.epsilon=100
   
   source('./SpatPred/R/util_funcs.R')
    
@@ -21,7 +23,7 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=10){
   x.len=sqrt(S)
  
   #generate Covariates from a Matern distribution
-  Covs=matrix(0,S,n.covs)
+  Covs=data.frame(matrix(0,S,n.covs))
   for(icov in 1:n.covs){
     kappa=runif(1,5,10)
     r=runif(1,0.2,.4)
@@ -35,11 +37,14 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=10){
     Y[Y>x.len]=x.len
     Grid=matrix(0,x.len,x.len)
     for(i in 1:length(X))Grid[X[i],Y[i]]=Grid[X[i],Y[i]]+1
-    Covs[,icov]=Grid/max(as.vector(Grid))
+    Covs[,icov]=as.vector(Grid)/max(as.vector(Grid))
+    #Covs[,icov]=expit(logit(Covs[,icov])+rnorm(length(Grid),0,0.05)) # add a bit of noise
[…]

### 2014-12-22 SpatPred/R/spat_pred.R

@@ -1,8 +1,7 @@
 #' function to perform Bayesian analysis of log Gaussian Cox processes and to make spatial predictions 
 #' @param formula A formula object specifying the model for the fixed effects
-#' @param Data   A SpatialPolygongsDataFrame holding covariate values for all cells one wants to make predictions on (including sampled cells)
-#'  The data frame should also include a column, "Count" which gives the response variable (only values associated with surveyed cells will be modeled.)
-#' @param Mapping  A vector giving the habitat cell id number for each surveyed cell
+#' @param Data   A SpatialPolygonsDataFrame holding covariate values for all cells one wants to make predictions on (including sampled cells)
+#' @param Effort A list including "Counts" which gives the response variable for surveyed cells, and "Mapping" which is a vector indicating which sample unit is associated with which count
 #' @param Area.adjust   A vector allowing for differences in suitable habitat for each cell.  Can be used for different grid cell sizes or different
 #'        proportions of suitable habitat (e.g., 1.0 = 100% of habitat is suitable for the focal species)
 #' @param Offset	A vector giving the fraction of each grid cell in Mapping that was surveyed
@@ -11,7 +10,7 @@
 #' @param Assoc   Association matrix for all grid cells in Data (needed for ICAR models only)
 #' @param K   A (# cells)x(# knots) matrix - the 'K' matrix for process convolution modeling.  
 #' @param Names.gam A character vector giving the column names of Data that will be modeled as smooth effects
-#' @param Knots.gam A list vector, where each entry is a matrix (default NULL)
+#' @param Knots.gam A list vector (length is # of smooth params), where each list element is a vector providing locations of knots (one location for each knot)
[…]

### 2014-12-22 SpatPred/R/util_funcs.R

@@ -123,6 +123,10 @@ d_biv_normal<-function(Tmp.vec,XY,Sigma){
 #' @keywords abundance map, plot
 #' @author Paul Conn \email{paul.conn@@noaa.gov}
 plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance"){
+  library(sp)
+  library(RColorBrewer)
+  myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
+  
   Tmp=Grid[[1]]
   if(is.null(highlight)==FALSE){
     midpoints=data.frame(gCentroid(Tmp[highlight,],byid=TRUE))
@@ -134,7 +138,7 @@ plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundan
   new.colnames[1:2]=c("Easting","Northing")
   colnames(Cur.df)=new.colnames
   tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank())
-  p1=ggplot(Cur.df)+aes(Easting,Northing,fill=Abundance)+geom_raster()+tmp.theme+scale_fill_continuous(name=leg.title)
+  p1=ggplot(Cur.df)+aes(Easting,Northing,fill=Abundance)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=myPalette(100))
   if(is.null(highlight)==FALSE){
     #p1=p1+geom_rect(data=midpoints,size=0.5,fill=NA,colour="yellow",aes(xmin=Easting-25067,xmax=Easting,ymin=Northing,ymax=Northing+25067))
     p1=p1+geom_rect(data=midpoints,size=0.5,fill=NA,colour="yellow",aes(xmin=Easting-1/2,xmax=Easting+1/2,ymin=Northing-1/2,ymax=Northing+1/2))

### 2014-12-22 SpatPred/inst/plot_survey_designs.R

@@ -1,0 +1,84 @@
+#plot designs
+
+S=900
+XY=expand.grid(y=c(sqrt(S):1),x=c(1:sqrt(S)))
+ids.reg=factor(c(1:50))
+ids.clust=factor(c(1:60))
+ids.IVH=ids.reg
+isim=1
+
+Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+load(Cur.file) #load abundance,covariate grid
+
+Cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
+load(Cur.file) #load Effort
+Effort.reg=Effort
+
+Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
+load(Cur.file) #load Effort
+Effort.clust=Effort
+
+Cur.file=paste("./Sim_data/Effort_IVH1",isim,".Rda",sep='')
+load(Cur.file) #load Effort
+Effort.IVH=Effort
+
+
+Data=Grid
+
+#compute gIVH for different designs
+
+
+
+df<-data.frame(id=ids.reg,x=XY[Effort.reg$Mapping,"x"],y=XY[Effort.reg$Mapping,"y"])
+Abund=Data@data[,"cov1"]*100 
+Abund.df=data.frame(x=XY[,"x"],y=XY[,"y"],Abundance=round(as.vector(Abund)))
+colnames(Abund.df)=c("x","y","Covariate")
+require(ggplot2)
+plot1<-ggplot(Abund.df,aes(x,y))+geom_tile(aes(fill=Covariate))+scale_x_continuous(expand=c(0,0))+scale_y_continuous(expand=c(0,0))+scale_fill_gradient(low="white",high="red")+xlab("")+ylab("")+ggtitle("A.")
+plot1<-plot1+geom_point(data=df,aes(x=x,y=y),shape=3,size=2)+theme(plot.title = element_text(hjust = 0,size = rel(1.5)),legend.position="none",axis.ticks = element_blank(), axis.text = element_blank())
+plot1   
+
+df<-data.frame(id=ids.clust,x=XY[Effort.clust$Mapping,"x"],y=XY[Effort.clust$Mapping,"y"])
+Abund=Data@data[,"cov1"]*100 
+Abund.df=data.frame(x=XY[,"x"],y=XY[,"y"],Abundance=round(as.vector(Abund)))
+colnames(Abund.df)=c("x","y","Covariate")
+require(ggplot2)
[…]

### 2014-12-22 SpatPred/inst/run_abundance_sims.R

@@ -1,132 +1,119 @@
 # run_generic_sims.R
 # script to run generic spatio-temporal count data simulations
 source("./SpatPred/R/sim_data_abundance.R")
-set.seed(12345)
-n.sims=100 #number of simulations at each design point
-S=400
-t.steps=20
-n.transects=5
-line.width=0.1
+source("./SpatPred/R/util_funcs.R")
+
+set.seed(123456)
+n.sims=1#number of simulations at each design point
+S=900
+prop.sampled=0.1
+n.transects=60
 
 i.sim=TRUE
 if(i.sim){
   for(isim in 1:n.sims){
     #simulate covariates, abundance
-    Grid=sim_data_generic(S,n.transects,n.covs=4,line.width=,tau.epsilon=20)
+    Grid=sim_data_generic(S=S,n.covs=4,tau.epsilon=10)
     Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
     save(Grid,file=Cur.file)
+  }
+  #now simulate count datasets (could possibly use clustered "convenience" sampling, but just using random right now)
+  for(isim in 1:n.sims){
+    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    load(Cur.file)
+    Effort=sim_effort(Data=Grid,S=S,formula=NULL,type="random",prop.sampled=prop.sampled,n.points=n.transects)
+    Cur.file=paste("./Sim_data/Effort_random",isim,".Rda",sep='')
+    save(Effort,file=Cur.file)
   }
 }
 
 
-#now simulate 
-for(isim in 1:n.sims){
-  Cur.file=paste("./Sim_data/Sim",isim,".Rda",sep='')
-  load(Cur.file)
-  
-}
+
 
 #s
 
 
 #call estimation routines
-Est.mods=c("CPIF","AST","STPC","OPRS")
-Sim.ind.20= c(1:19)*5
-for(igen in 1:3){  #loop over generating model to generate data sets
-  for(itrans in 1:1){ #loop over number of transects in each cell
-    for(isim in 1:(n.sims/2)){  
+Adj=square_adj(sqrt(S))
+for(igen in 1:1){  #loop over generating model to generate data sets
+  for(isim in 1:n.sims){  
[…]

### 2014-12-22 SpatPred/inst/run_design_sims.R

@@ -1,0 +1,167 @@
+# run_generic_sims.R
+# script to run generic spatio-temporal count data simulations
+source("./SpatPred/R/sim_data_abundance.R")
+source("./SpatPred/R/util_funcs.R")
+
+set.seed(123456)
+n.sims=1#number of simulations at each design point
+S=900
+prop.sampled=0.1
+n.transects=60
+
+i.sim=TRUE
+if(i.sim){
+  for(isim in 1:n.sims){
+    #simulate covariates, abundance
+    Grid=sim_data_generic(S=S,n.covs=4,tau.epsilon=10)
+    Grid$cov1.quad=Grid$cov1^2
+    Grid$cov2.quad=Grid$cov2^2
+    Grid$cov3.quad=Grid$cov3^2
+    Grid$cov12=Grid$cov1*Grid$cov2
+    Grid$cov13=Grid$cov1*Grid$cov3
+    Grid$cov23=Grid$cov2*Grid$cov3
+    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    save(Grid,file=Cur.file)
+  }
+  #now simulate count datasets 
+  for(isim in 1:n.sims){
+    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    load(Cur.file)
+    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="balanced",prop.sampled=prop.sampled,n.points=50)
+    Cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
+    save(Effort,file=Cur.file)
+    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="clustered",prop.sampled=prop.sampled,n.points=60)    
+    Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
+    save(Effort,file=Cur.file)
+    my.formula=~cov1+cov1.quad
+    Effort=sim_effort(Data=Grid,S=S,my.formula=my.formula,type="IVH",prop.sampled=prop.sampled,n.points=50)    
+    Cur.file=paste("./Sim_data/Effort_IVH1",isim,".Rda",sep='')
+    save(Effort,file=Cur.file)
+    my.formula=~(cov1+cov2)^2+cov1.quad+cov2.quad
[…]

### 2014-12-22 SpatPred/inst/test_GAM.R

@@ -1,0 +1,39 @@
+#test GAM 
+library(mvtnorm)
+X=c(0:100)/100
+n=length(X)
+knot.loc=c(0,0.33,0.66,1)
+Y=X+X^3+rnorm(n,0,0.1)
+sd=0.5
+mcmc.iter=10000
+
+K=matrix(0,101,4)
+for(icol in 1:4){
+  K[,icol]=dnorm(X,knot.loc[icol],sd)
+}
+K=K/apply(K,1,'sum')
+
+Alpha.mc=matrix(0.5,4,mcmc.iter)
+Beta=0
+Beta.mc=rep(0,mcmc.iter)
+tau.epsilon=100
+
+KpKinv=solve(crossprod(K))
+for(iiter in 2:mcmc.iter){
+  Delta=Y-K%*%Alpha.mc[,iiter]
+  #update intercept
+  Beta.mc[iiter]=rnorm(1,sum(Delta)/n,1/(n*tau.epsilon+0.01*n))
+
+  #update tau.epsilon
+  Delta=Y-K%*%Alpha.mc[,iiter]-Beta.mc[iiter]
+  tau.epsilon=rgamma(1,0.5*n+1,0.5*crossprod(Delta)+0.01)
+  
+  #update Alpha
+  Alpha.mc[,iiter]=rmvnorm(1,KpKinv%*%crossprod(K,Y-Beta.mc[iiter]),KpKinv/(tau.epsilon+0.01))
+  
+}
+
+Alpha.mean=apply(Alpha.mc,1,'mean')
+plot(X,Y)
+lines(X,K%*%Alpha.mean+mean(Beta.mc))
+#SWEET!

### 2015-01-20 manuscript/AppendixB.tex
Spatial prediction code & paper with gIVH, prediction variance using delta method.  Currently spatial gIVH programmed with RSR-ICAR model which doesn't work well because you can't limit K to just those cells that are sampled since Q is still full dimensional. Need to switch to process convolution for gIVH computation.
SpatPred draft to Peter and Devin for first time - doesn't have Steller example integrated into text yet

@@
 
-\begin{center} \bf {\large Avoiding extrapolation bias when using statistical models to make ecological prediction}
+\begin{center} \bf {\large On extrapolating past the range of observed data when making statistical predictions in ecology}
 
@@
 
-We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where extrapolations from statistical species distribution models may be problematic.
-Each simulation consisted of several steps, including
+Spatially explicit statistical models are increasingly used to estimate animal abundance and predict species distributions from count data.
+We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where predictions of animal abundance using these models may be problematic, and to determine whether statistical inferences are more robust when restricted to sampling units within the gIVH.
+Each simulation replicate consisted of several steps, including
 \begin{enumerate}
-  \item Simulate four hypothetical environmental covariates over a 30 by 30 grid,
-  \item Simulate animal abundance across the landscape as a function of covariates,
-  \item Simulate the position of animal count quadrats across the landscape (i.e. locations where sampling is to occur) and animal counts on each quadrat,
-  \item Estimate animal abundance as a function of environmental covariates according to each of three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).
-For the latter, only a spatial dimension was considered.
+  \item Simulate three hypothetical, statistically dependent, spatially autocorrelated environmental covariates over a 30 by 30 grid,
+  \item Simulate animal abundance across the landscape as a function of environmental covariates,
+  \item Simulate animal population surveys across the landscape, including the position of count quadrats and the resulting animal counts,
+  \item Estimate animal abundance as a function of two of the environmental covariates according three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).
+For the latter, only a spatial dimension was modeled.
 All such models were specified hierarchically, with MCMC used for posterior simulation.
-  \item Calculation of the gIVH as determined by (1) integrating over prior distributions for unknown parameters ($\text{gIVH}_{int}$), or (2) using realized posterior variance (i.e.
-after data were collected and analyzed; $\text{gIVH}_{post}$).
+  \item Calculate the gIVH using realized posterior variance (i.e. after data were collected and analyzed).
 \end{enumerate}
 We now describe each of these tasks and results of the simulation study in further detail.
-All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the supplement.
+All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the R package \texttt{SpatPred} that accompanies this article.
 
 \underline{1. Simulating environmental covariates} \\
-For each simulation, we generated environmental covariates by discretizing the realizations of a Matern Cluster Process (Fig.
-\ref{fig:Matern}) using the rMatClust function within the R \texttt{spatstat} package \citep{BaddeleyTurner2005}.
-The Matern Cluster Process is a point process that includes a total of $\kappa$ cluster centers.
-Within a radius $r$ for each cluster center, a total of $n$ points are randomly distributed, where $n \sim \text{Poisson}(\lambda)$ (e.g.
-Fig.
-\ref{fig:Matern}).
-For each simulated covariate, we drew $\kappa \sim U(5,10)$, $r \sim U(0.2,0.4)$, and $\lambda \sim U(200,500)$ to provide variation in the relative density and clustering for each covariate.
-Next, we overlayed a 30 by 30 grid over simulated points.
-To generate a hypothetical covariate, we then summed the number of points within each grid cell, and standardized so that the maximum covariate value was 1.0 (e.g.
-\ref{fig:Matern}).
+In the real world, different habitat covariates are often correlated with each other (e.g., altitude and precipitation), and are often patchily distributed across the landscape (i.e., are spatially autocorrelated).
+We thus desired a procedure for generating covariates that would allow some level of statistical dependence among covariates, together with spatial autocorrelation.
+For each simulation, we generated three spatially autocorrelated environmental covariates using a procedure motivated by linear coregionalization models in multivariate spatial statistics \citep[e.g.][]{GoulardVoltz1992} to impart desired behavior.
+To start, we used the R package \texttt{RandomFields} to simulate 10 realizations ${\bf y}_i$ ($i \in 1,2,\hdots,10$) of independent, mean-zero random fields over a $30 \times 30$ grid (where the lower case bold type denotes a vector).
+Each random field had a stationary, isotropic, exponential covariance structure, where the covariance $C$ between two survey units (i.e.
+grid cells) was a function of the distance $r$ between grid cell centroids, $C(r)=\exp(r/v)$.
+Note that the distance between horizontally and vertically adjacent grid cell centroids was standardized to 1.0; the scale parameter $v$ of the exponential covariance function for each random field was drawn from a Uniform(5,100) distribution to induce heterogeneity in the spatial scale of each process.
 
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=6in]{MaternCov.pdf}
-\end{center}
-\caption{Environmental covariates were generated via a discretized Matern Cluster Process.
-First, a Matern Cluster Process (A.) is simulated.
-Next, the number of simulated points in each grid cell is counted and standardized to have a maximum of 1.0 (B.).}
-\label{fig:Matern}
-\end{figure}
+Next, we determined the values of three spatially autocorrelated habitat covariates, ${\bf z}_j$ by writing them as linear functions of the ${\bf y}_i$:
+\begin{linenomath}
+\begin{equation*}
+  {\bf z}_j = \sum_i \omega_{ij} {\bf y}_i.
+\end{equation*}
+\end{linenomath}
+Evidently, the covariance between each induced habitat covariate is a function of the weights $\omega_{ij}$.  We used the following strategy to set $\omega_{ij}$:
+\begin{enumerate}
+  \item Set all $\omega_{ij}=0$
+  \item For each desired covariate, $j$, randomly sample four values ${\bf u}_j$ from the set $\{ 1,2,\hdots,10\}$ without replacement.
+  \item For each value $u \in {\bf u}_j$, set $\omega_{uj} \sim \mathcal{N}(0,1)$.
+\end{enumerate} 
+This procedure led to ``patchy" habitat covariates with realistic levels of covariation (Fig.
+\ref{fig:covs}).
+To prevent redundancy and collinearity, we rejected and resampled covariate values (using the same procedure) whenever maximum absolute correlation among covariates was greater than 0.75.
 
@@
 
-Given values of the four simulated covariates, we generated a vector of expected log-abundance in each cell ($\boldsymbol{\mu}$) as
+Given values of the three simulated covariates, we generated a vector of expected log-abundance in each cell ($\boldsymbol{\mu}$) as
 \begin{linenomath}
 \begin{equation*}
-  $\boldsymbol{\mu}$ = \beta_0 + {\bf X}\boldsymbol{\beta} + \boldsymbol{\epsilon},
+  \boldsymbol{\mu} = \beta_0 + {\bf X}\boldsymbol{\beta} + \boldsymbol{\epsilon},
 \end{equation*}
 \end{linenomath}
-where the intercept, $\beta_0$, was drawn from a $\mathcal{N}(2.5,0.25)$ distribution, the other regression coefficients ($\boldsymbol{\beta}$) were each drawn from a $\mathcal{N}(0,0.8)$ distribution, and residual Gaussian errors ($\boldsymbol{\epsilon}$) were drawn from a $\mathcal{N}(0,0.1)$ distribution.
-The design matrix ${\bf X}$ was constructed assuming linear and quadratic effects for each covariate, together with all one-way interactions for a total of 14 regression coefficients in addition to the intercept.
-Next, abundance in each cell $i$ was generated as  
+where the regression coefficients $\boldsymbol{\beta}$ were each drawn from a $\mathcal{N}(0,\tau)$ distribution, and the intercept, $\beta_0$, was drawn from a $\mathcal{N}(2.5,0.25)$ distribution.
+The design matrix ${\bf X}$ was constructed assuming linear and quadratic effects for each covariate, together with all one-way interactions for a total of 9 coefficients in addition to the intercept.
+The precision, $\tau$, was set to 2.5 for linear fixed effects, and to 5.0 for quadratic effects and one-way interaction terms.
+Residual Gaussian errors ($\boldsymbol{\epsilon}$) were drawn from a $\mathcal{N}(0,10)$ distribution.
+Abundance in each cell $i$ was generated as
 \begin{linenomath}
@@
 \end{linenomath}
+Regression coefficients were redrawn whenever $\sum_i N_i > 100,000$ or if the 20 most populous grid cells included $>90\%$ of total abundance to prevent unreasonably high or constricted distributions of abundance.
 
@@
 
-For each simulated landscape, we selected 60 grid cells at random (without replacement) for sampling.
-We configured simulations such that sample quadrats covered 10\% of the target grid cell, and generated animal counts for each quadrat as
+For each simulated landscape, we selected 45 grid cells ($5\%$) for sampling.
+We employed two possible survey designs: i) spatially balanced sampling using a random tessellation design \citep{StevensOlsen2004}, and ii) a convenience sample where the inclusion probability of each grid cell $i$ in the sampling frame was set proportional to $\exp(-0.2*r_i)$, where $r_i$ is the Euclidean distance between the centroid of grid cell $i$ and the center of the survey grid.
+The latter survey design was meant to approximate the case where there is a ``base of operations" in the middle of the survey grid and more effort is expended close to the center due to simpler sampling logistics.
+We configured simulations such that sample quadrats covered 10\% of each targeted grid cell, and generated animal counts for each quadrat $j$ as
 \begin{linenomath}
 \begin{equation*}
-  C_i \sim \text{Binomial}(N_i,0.1).
+  C_j \sim \text{Binomial}(N_j,0.1).
 \end{equation*}
@@
 
-For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models.  The models were built to resemble a GLM, a GAM, and an STRM.
+For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models, corresponding to a generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a generalized additive model (GAM; smooth effects of covariates on the log scale), and a spatio-temporal regression model (with both linear, fixed effects and spatially autocorrelated random effects).
+Further details on these models, together with the procedure used for posterior simulation, are presented in Appendix A.
+For this study, we used the RSR implementation of the STRM outlined in Appendix A.
+Prior distributions for each precision parameter, $\tau$, were set to $\text{Gamma}(1.0,0.01)$, which is diffuse while maintaining a flat shape near the origin.
+Regression parameters ($\beta$) were given vague $\mathcal{N}(0,\tau=0.01)$ priors.
+
+\underline{5. Calculating the gIVH} \\
 
@@
 \end{flushleft}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=6in]{Covs_coregion.pdf}
+\end{center}
+\caption{For each simulation, three spatially autocorrelated environmental covariates were generated via a linear coregionalization model.
+Panels A-C show a single realization from this procedure, while panel D shows the distribution of sample correlations between two randomly selected covariates over 1000 simulations.}
+\label{fig:covs}
+\end{figure}
+
 \end{document}

### 2015-01-20 manuscript/SpatPred3.tex
SpatPred draft to Peter and Devin for first time - doesn't have Steller example integrated into text yet

--- 2014-11-24 manuscript/SpatPred3.tex
@@
 \begin{center}
-Avoiding extrapolation bias when using statistical models to make ecological prediction
+On extrapolating past the range of observed data when making statistical predictions in ecology
 \bigskip\\
@@
 In ecology and conservation, a common goal is to make predictions about an unsampled random variable given a limited sample from the target population.
-For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_i$, we often desire to predict a new observation $y_{new}$ at $i$.
+For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_s$, we often desire to predict a new observation $y_s$ at $s$.
 For instance, we might use a generalized linear model \citep{McCullaghNelder1989} or one of its extensions to predict species density or occurrence in a new location, or to predict the future trend of a population.
@@
 past the range of observed data.
-This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new point $y_{new}$ whenever $x_{new} < \min({\bf x})$ or $x_{new} > \max({\bf x})$.
+This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new point $y_s$ whenever $x_s < \min({\bf x})$ or $x_s > \max({\bf x})$.
 But what about more complicated situations
@@
 
-Our aim in this paper is to investigate extrapolation bias in the generalized linear model and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatial, temporal, or spatio-temporal regression models (STRMs).
+Our aim in this paper is to investigate one possibility for defining extrapolation in the generalized linear model and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatial, temporal, or spatio-temporal regression models (STRMs).
 In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers \citep{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points.
 Application of the gIVH and related criterion (e.g.
-prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction.
-Also, since the gIVH can be constructed solely with knowledge of sampled locations and explanatory covariates (i.e., it does not necessarily require any observed response variables), it can also be used to help guide survey design.
-We illustrate use of the gIVH on a simulated occupancy dataset, on a species distribution model (SDM) for ribbon seals in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).
+prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction and survey design.
+We illustrate use of the gIVH on simulated count data, on a species distribution model (SDM) for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).
 
@@
 \ref{fig:IVH}).
-These simple examples highlight the sometimes counterintuitive nature of the predictive inference problem, a problem that can only become worse as models with more dimensions are contemplated (including those with temporal or spatial structure).
+These simple examples highlight the sometimes counterintuitive nature of predictive inference, a problem that can only become worse as models with more dimensions are contemplated (including those with temporal or spatial structure).
 Fortunately, the ideas behind the IVH provide a potential way forward.
@@
 However, since the hat matrix is proportional to prediction variance, Cook (1979) notes that design points with maximum prediction variance will be located on the boundary of the IVH.
-We therefore define a generalized independent variable hull (gIVH) as the set of all points ${\bf x}$ (note that ${\bf x}$ can include both observed and unobserved design points) such that
+We therefore define a generalized independent variable hull (gIVH) as the set of all potential design points $\mathcal{S}_0$ (note that $\mathcal{S}_0$ can include both observed and unobserved design points) such that
 \begin{linenomath*}
@@
   \label{eq:gIVH}
-  {\rm var}({\bf y}_x | {\bf x}) \le {\rm max} [{\rm var}({\bf y}_X | {\bf X})],
+  {\rm var}(\lambda_s) \le {\rm max} [{\rm var}(\boldsymbol{\lambda}_{\mathcal{S}})],
 \end{equation}
 \end{linenomath*}
-where ${\bf Y}_x$ correspond to predictions at ${\bf x}$, and ${\bf X}$ and ${\bf Y}_X$ denote observed design points and predictions at $X$, respectively.
+where $s \in \mathcal{S}_0$, $\lambda_s$ corresponds to the mean prediction at $s$, $\mathcal{S}$ denotes the set of observed design points, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.
 
@@
 \begin{equation}
-  Y_i \sim f_Y(g(\mu_i)),
+  Y_s \sim f_Y(g(\mu_s)),
 \end{equation}
@@
 where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives an inverse link function, and
-$\mu_i$ is a predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
+$\mu_s$ is a linear predictor.  For many such generalizations, it is possible to specify the $\mu_s$ as
 \begin{linenomath*}
@@
 \begin{linenomath*}
-\begin{eqnarray*}
-  \label{eq:var-mu}
-  {\rm var}(g(\mu_x) | {\bf x}) & = & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x | {\bf x})\boldsymbol{\Delta}^\prime, 
+\begin{eqnarray}
+  \label{eq:var-lambda}
+  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
+  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
 %  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
@@
 
-%For GLMs, prediction variance is proportional to the original hat matrix ${\bf V}_{LR}$ (at %least on the linear predictor scale), and Eq.
-\ref{eq:IVH} may be used directly.
-For other %models, we have
-%\begin{linenomath*}
-%\begin{equation}
-%  \label{eq:var_Beta}
-%  {\rm var}(\hat{\boldsymbol{\beta}}_{aug})=\left[ \begin{array}{cc}
-%      \boldsymbol{\Sigma}_\beta & {\bf 0}\\
-%      {\bf 0} & \boldsymbol{\Sigma}_\alpha
-%  \end{array} \right],
-%\end{equation}
-%\end{linenomath*}
-%where we simply replace $\boldsymbol{\beta}_{aug}$ with the augmented vector that includes %both regression and spatial or smooth parameters (i.e.
-$\boldsymbol{\beta}$ and %$\boldsymbol{\alpha}$; Appendix S1).
-
-The exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ differs depending on the underlying model structure and estimation procedure.
-In the following treatment, we shall focus on Bayesian analysis; although it is not necessarily needed to fit GLMs and GAMs, it helps make STRMs more computationally tractable and puts all of the different models into a common analysis framework.
-This approach requires specifying priors for model parameters, and prior parameters may appear in expressions for prediction variance.
-Judicious choices of priors can at times limit the influence of priors on calculation of the gIVH.
-For example, if we specify the prior on the vector of regression coefficients to be $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$ where $\tau_\beta$ is a fixed constant, a Bayesian implementation of a GLM still results in the gIVH specified in Eq.
-\ref{eq:IVH}.
-The ability to use Eq.
-5 directly is quite useful, as we only need to know the explanatory variables to be able to diagnose whether predictions lie in or out of the gIVH.
-However, the other models considered here (GAMs and STRMs) typically require estimation of one or more precision or autocorrelation parameters.
-In these cases, one solution is to impose prior distributions and integrate over these additional parameters (call them $\boldsymbol{\theta}$) to obtain an expectation for ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$.
-Using the law of the unconscious statistician, we have
+Unlike Cook's original formulation, the gIVH will often involve unknown parameters.
+Further, the exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the underlying model structure and estimation procedure.
+For most of the following treatment, we shall assume that data have already been collected (see Discussion for comments on the potential use of the gIVH in survey planning), and that informative probability distributions can be specified for these parameters.
+Specifically, we impose a probability distribution $[\boldsymbol{\theta}]$ for unknown parameters, and compute prediction variance as
 \begin{linenomath*}
 \begin{equation}
-  \label{eq:E_var_Beta}
-  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}))  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
+  \label{eq:E_var_lambda}
+  {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}) = \int_{\boldsymbol{\theta}} {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}, \boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
 \end{equation}
 \end{linenomath*}
-Here, and throughout the paper, we use the bracket notation to indicate a probability density function; e.g., $[\boldsymbol{\theta}] $ denotes the joint prior distribution for $\boldsymbol{\theta}$.
-If observations have already been gathered, one could also consider integrating over the marginal posterior distribution using MCMC:
-\begin{linenomath*}
-\begin{equation}
-  \label{eq:E_var_Beta_post}
-  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}) | {\bf Y})  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}| {\bf Y}] d \boldsymbol{\theta}.
-\end{equation}
-\end{linenomath*}
-Some care must be taken with prior distributions if Eq.
-\ref{eq:E_var_Beta} is used for calculation of prediction variance.
-For instance, standard ``non-informative" or ``flat" priors may place substantial mass on implausible values.
-When using the gIVH or prediction variance to help guide sampling design \citep[see][for an example using prediction variance]{DiggleLophaven2006}, we suggest using informative prior distributions.
+Our approach in this paper is to conduct a Bayesian analysis and substitute the posterior distribution $[\boldsymbol{\theta}|{\bf Y}]$ for $[\boldsymbol{\theta}]$ in Eq.
+\ref{eq:E_var_lambda}.
+However, one could also consider substituting parametric approximations based on output from standard likelihood-based analyses.
 
@@
 In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).
-The gIVH, together with prior-integrated prediction variance, seem ideally situated to diagnosing potential extrapolation issues as it does not necessarily need to rely on gathered response data.
-Thus, one can examine whether or not prediction points lie within the gIVH without ever collecting response data there.
-Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions are being based on.
+The gIVH, together with posterior prediction variance, seem ideally situated to diagnosing potential extrapolation issues.
+Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions will be based on.
+
+\subsection{Computing}
+
+We developed a package \texttt{SpatPred} in the R statistical programming environment \citep{RTeam2012} to simulate data and conduct all analyses.
+The seal and sea lion datasets are included as part of this package, and are available at \url{https://github.com/pconn/SpatPred}.
+The R package has also been archived via figshare [WE PLAN TO DO THIS WHEN THE MANUSCRIPT IS ACCEPTED FOR PUBLICATION].
 
 \section{Examples}
+
+We used simulation to investigate the usefulness of the gIVH for limiting the scope of inference when fitting models to count data that have already been collected.
+We also applied the gIVH to examining the robustness of a species distribution model (SDM) of ribbon seals in the eastern Bering Sea, and to time series projections of Steller sea lion population trends.
 
@@
 For each of 100 simulations, we generated animal abundance over a $30 \times 30$ grid assuming that animal density was homogeneous in each grid cell.
-Animal abundance was generated as a function of four hypothetical habitat covariates (Appendix B).
-For each simulated landscape, we randomly selected 60 grid cells for sampling, assuming that sampling quadrats encompassed 10\% of the area of each selected grid cell.
+Animal abundance was generated as a function of three hypothetical spatially autocorrelated habitat covariates (Appendix B).
+For each simulated landscape, we conducted virtual surveys of $n=45$ survey units using two different designs: (1) a spatially balanced sample using a random tesselation design \citep{StevensOlsen2004}, and (2) a convenience sample where the probability of sampling was greater for cells closer to a ``base of operations" located in the middle of the survey grid.
+The former approach preserves randomness while seeking a degree or regularity when distributing sampling locations across the landscape, while the latter may be easier to implement.
+
+We configured virtual sampling quadrats such that they encompassed 10\% of the area of each selected grid cell.
 For ease of presentation and analysis, we assumed detection probability was 100\% in each quadrat.
-Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (Appendix B).
-The GLM and STRM expressed log-density as a function of linear and quadratic covariate effects, while the GAM used a kernel smoother with 4 knots for each covariate (Appendix B).
-Each model was provided with three of the four covariates used to generate the data.
+Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (Appendix A).
+The GLM and STRM expressed log-density as a function of linear and quadratic covariate effects (as well as first order interactions), while the GAM used a kernel smoother with 4 knots for each covariate (Appendix A).
+Each model was provided with two of the three covariates used to generate the data.
 
-For each simulated data set and model structure, we calculated (1) $\text{gIVH}_{int}$, the gIVH resulting from integrating over prior parameter variance (i.e.
-using Eq.
-\ref{eq:E_var_Beta} with informative priors), and (2) $\text{gIVH}_{post}$, the gIVH arising from posterior predictive variance (i.e.
-posterior variance when data are actually analyzed).
-We then calculated posterior predictions of animal abundance within and outside of each gIVH in order to gauge bias as a function of whether or not inference is constrained to the gIVH.
-Specifically, the performance of the $\text{gIVH}_{int}$ criterion in diagnosing extrapolation bias is a test of whether it is useful in designing animal count surveys, while the performance of $\text{gIVH}_{post}$ may help decide its utility in limiting the scope of landscape-based inference once data have been collected and analyzed.
+For each simulated data set and model structure, we calculated the posterior predictive variance and resulting $\text{gIVH}$ as in Eq.
+\ref{eq:gIVH}.
+We then calculated posterior predictions of animal abundance within and outside of each gIVH in order to gauge bias as a function of this restriction.
+Specifically, the performance of $\text{gIVH}$ may help decide its utility in limiting the scope of inference once data have been collected and analyzed, and perhaps point out areas worthy of additional sampling.
+A fuller, technical description of the simulation study design is provided in Appendix B; for illustration of a single simulation replicate see Fig.
+\ref{fig:sim_maps}.
+
+Posterior predictions from simulations indicated that the distribution for proportional error in total abundance was right skewed when statistical inference was made with regard to the entire survey area (Fig.
+\ref{fig:sim_boxplots}).
+This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.
+The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.
+Interestingly, the convolution kernel GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.
+For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude (e.g.
+$\approx 0.3$; Fig.
+\ref{fig:sim_boxplots}) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
 
@@
 National Marine Fisheries Service conducted aerial surveys over the eastern Bering Sea in 2012 and 2013.
-Agency scientists used infrared video to detect seals that were on ice, and simultaneous automated digital photographs provided information on species identity.
+Agency scientists used infrared video to detect seals that were on ice, and collected simultaneous digital photographs to provide information on species identity.
 Here, we use spatially referenced count data from photographed ribbon seals, {\it Phoca fasciata} (Fig.
-\ref{fig:ribbon}) on a subset of 10 flights flown over the Bering Sea in April 2012.
-These flights were limited to a one week period so that both sea ice conditions and seal distributions could be assumed to be static.
+\ref{fig:ribbon}) on a subset of 10 flights flown over the Bering Sea from April 20-27, 2012.
+We limited flights to a one week period because sea ice melts rapidly in the Bering Sea in the spring, and modeling counts over a longer duration would likely require addressing how sea ice and seal abundance changes over both time and space \citep[see][]{ConnEtAl2015}.
+However, limiting analysis to a one week period makes the assumption of static sea ice and seal densities tenable \citep{ConnEtAl2014}.
 
@@
 Figure \ref{fig:flights} shows the transects flown and the number of ribbon seals encountered in each cell, and Figure \ref{fig:covs} show explanatory covariates gathered to help predict ribbon seal abundance.
-These data are described in fuller detail by \citet{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors \citep[see e.g.][]{ConnEtAl2014}.
+These data are described in fuller detail by \citet{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors.
 Since our focus in this paper is on illustrating spatial modeling concepts, we devote our efforts to the comparably easier problem of estimating apparent abundance (i.e., uncorrected for vagaries of the detection process).
@@
 issue with spatial prediction: abundance of ribbon seals appears to be maximized in the southern and/or southeast quadrant of the surveyed area.
-Predicting abundance in areas further south and east may thus prove problematic.
+Predicting abundance in areas further south and east may thus prove problematic, as the values of several explanatory covariates (Fig.
+\ref{fig:covs}) are also maximized in these regions.
 To illustrate, let $Y_i$ denote the ribbon seal count ($Y_i$) obtained in sampled grid cell $i$.
@@
   Y_i \sim {\rm Poisson}(\lambda_i) \mbox{ and } \\
-  \log(\lambda_i) = \log(P_i) + {\bf X}_i \boldsymbol{\beta} + \eta_i + \epsilon_i, \nonumber
+  \log(\lambda_i) = \log(P_i) + \log(A_i) + \theta_i + \epsilon_i, \nonumber
 \end{eqnarray}
-where $P_i$ gives the proportion of area photographed in grid cell $i$ (recall also that ${\bf X}_i$ denotes a vector covariates for cell $i$, $\boldsymbol{\beta}$ are regression coefficients, $\eta_i$ represents a spatially autocorrelated random effect, and $\epsilon_i$ is normally distributed $iid$ error).
+where $P_i$ gives the proportion of area photographed in grid cell $i$, $A_i$ gives the proportion of suitable habitat in cell $i$ (ribbon seals do not often utilize land), $\theta_i$ is a linear predictor, and $\epsilon_i$ is normally distributed $iid$ error.
+By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs, GAMs, and STRMs (see Appendix S1).
 
-We could fit any number of predictive models to these data, but we start with a simple
-generalized linear model where we ignore the spatial random effect, $\eta_i$, and use the full suite of predictor covariates (Fig.
-\ref{fig:covs}) to fit Eq.
-\ref{eqn:Cox} to our data.
-In particular, we fit a model with linear effects of all predictor variables, and with an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}.
-To enable comparison with more complicated types of models, we formulated a generalized Bayesian strategy for parameter estimation (see Appendix S1).
-For simplicity, we generated posterior predictions of ribbon seal abundance across the landscape as
+We start by fitting a hierarchical GLM, GAM, and STRM to the full suite of predictor covariates (Fig.
+\ref{fig:covs}).
+For the GLM and STRM, we fit a model with linear effects of all predictor variables, and with an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}.
+For the STRM, we imposed a restricted spatial regression (RSR) formulation for spatially autocorrelated random effects, where dimension reduction was accomplished by only selecting eigenvectors of the spectral decomposition associated with eigenvalues that were greater than 0.5 (see Appendix S1 for additional information on model structure).
+For the GAM, we employed smooth terms with 8 knots for each covariate.
+This included 4 knots at the minimum, 1/3 and 2/3 quantiles, and the maximum observed covariate values, in addition to 4 knots above and below the range of observed data to reduce the magnitude of biased endpoint behavior.
+
+As with simulated data examples, we adopted a Bayesian perspective and conducted estimation using MCMC (see Appendix S1 for algorithm details) with 60,000 iterations where the first 10,000 iterations were discarded as a burn-in.
+We generated posterior predictions of ribbon seal abundance across the landscape as
 \begin{equation}
@@
 \end{equation}
-where $A_i$ gives the proportion of suitable habitat in cell $i$ (ribbon seals do not inhabit land masses).
+and calculated the gIVH as in Eqn. \ref{eq:gIVH}, with delta method modifications as specified in Eqns. \ref{eq:var-mu}-\ref{eq:E_var_lambda}.
+Initial spatial predictions using the three models (GLM, GAM, STRM) all produced extremely high, unbelievable predictions along the southern boundary of the study area \ref{fig:ribbon_plot}.
+Predictions in this region were also largely out of the gIVH, indicating the potential utility for the gIVH in revealing problematic extrapolations.
 
-Fitting this model to our data,
+We considered several possible alternatives for trying to obtain more robust abundance estimates before settling on a preferred alternative.
+First, one could refine the study area to eliminate predictions outside of the gIVH (as in the simulation study).
+However, this is not ideal in that one does not get an abundance estimate for the whole study area, and it may be difficult to compare abundance from one year to the next using this approach.
+Second, one could consider a simpler estimation model, such as one post-stratifying by ecoregion as was used in a previous analysis of ribbon seal data to try to avoid this problem \citep{ConnEtAl2014}.
+Finally, one could build in a priori knowledge of habitat preferences into the model structure.
+We adopted the latter solution, incorporating ``pseudo-absences" (i.e., zero counts where sampling was not conducted) in locations where it would have been (nearly) impossible to detect seals.
+Specifically, we inserted pseudo-absences in cells where ice concentrations were $<0.1\%$.
+This solution seemed the most logical, as many of the large, anomalous predictions were over open water along the southern edge of the study area, where we would have obtained zero counts had they been surveyed.
+
+Fitting our three models to the augmented dataset with pseudo-absences, most predictions occurred within the gIVH (Fig.
+\ref{fig:ribbon_plot}).
+The largest difference among the three models was in the southwest corner of the study area, coinciding with the few cells that were outside the gIVH in the three models.
+Posterior summaries of abundance across the study area were very comparable (albeit with a longer right hand tail for the STRM), with $5\%$, $50\%$, and $95\%$ posterior prediction quantiles as follows: GLM (48686,64836,93927); GAM(46120,61214,90873); STRM(41039,63717,194095).
+We note that these estimates are for example illustration only, as they are uncorrected for imperfect detection \citep[e.g., incomplete detection of thermal cameras, animals that were unavailable for sampling because they were in the water, species misidentification; see][]{ConnEtAl2014}.
 
@@
 
-A number of authors have explored optimal knot placement in spatial models.
-In the context of predictive process modeling \citep[where a covariance function is specified over a group of knots; see][]{BanerjeeEtAl2008}, \citet{FinleyEtAl2009} and \citet{GelfandEtAl2013} considered near-optimal knot placement by minimizing spatially averaged prediction variance.
-Gelfand et al. knot selection
+We have demonstrated the capacity for certain classes of statistical models to produce biased predictions of animal abundance when extrapolating past the range of observed data.
+In simulations, commonly used models (GLMs, STRMs) exhibited positive bias when predictions were required for the entire study area, particularly when convenience sampling was employed; in the ribbon seal example, naive extrapolation of fitted statistical relationships produced high positive bias along the southern boundary of the study area; and in the Steller sea lion example, projected abundance trends were increasingly implausible the farther the predictions were from the realm of observed data.
+However, the gIVH appeared useful in diagnosing places where extrapolations from the fitted statistical model were problematic.
+For instance, in the simulation study, limiting the scope of inference to locations falling within the gIVH reduced or eliminated bias.
+For ribbon seal relative abundance, it was useful for confirming that the naive models needed to be reformulated, and that reformulated models (with pseudo-absence data) produced predictions in unsurveyed areas that were largely within the realm of observed data.
 
-Using GAs to select knot placement using
+Use of the gIVH in seal and sea lion applications to identify extrapolation issues was not particularly profound, as experts could easily use ecological knowledge of these populations to identify biologically implausible predictions and come to the conclusion that models either needed to be reformulated or inference limited to a smaller set of design points.
+Likewise, one could argue that the high positive bias in the simulation study could likely have been reduced had an expert meticulously examined diagnostics from each simulation replicate and reconfigured estimation models to better match the range of covariates observed.
+However, such determinations are likely to be quite subjective, and may prove insufficient when there are large number of regression coefficients and interaction terms.
+Even relatively simple regression models may exhibit non-intuitive patterns (e.g.
+Fig.
+\ref{fig:IVH}).
+Further, relying on expert opinion alone in successive rounds of model formulation and fitting may lead to investigators choosing models based on how much they like the results, which is clearly not ideal scientific practice.
 
-Contrast with posterior loss (e.g., Jay's linex loss function estimator)
+Our intent in writing this paper is to raise awareness of potential problems with extrapolation bias in statistical models, and to provide an additional tool (the gIVH) to help diagnose its presence.
+Other approaches for selecting models to reduce overfitting, such as cross validation \citep[e.g.][]{PicardCook1984}, are also useful for this purpose, but may not entirely eliminate the problem (particularly for sparse datasets).
+In fields such as species distribution modeling and model-based abundance estimation, the goal for analysts is often to build predictive maps of species abundance and/or occurrence using a limited number of sample locations.
+In such applications, the ultimate aim of analysts should be to build models that have low bias and maximal precision.
+However, traditional approaches to quantifying bias (e.g., goodness-of-fit statistics) only work with observed data points.
+When inference is extended to unsampled locations, the gIVH appears to be a useful diagnostic for whether bias for predictions in unsampled locations can be expected.
 
-Contrast with cross validation
+The analyses in this paper focused on abundance estimation, which is necessarily non-negative.
+As such, counts are usually analyzed with a log link function, and there is a much greater potential for positive bias than negative.
+However, one could also apply the gIVH when predicting species occurrence from presence/absence data.
+In this case, common link functions (e.g.
+probit or logit) are symmetric, and potential for positive and negative bias in predictive maps seem equally likely.
+Future research should be directed to examine conditions under which the gIVH is a useful diagnostic in such applications.
 
-David Miller (the evil one)/Simon Wood stuff on edge effects
+One area that gIVH ideas may also prove useful is in formulating survey designs.
+The topic of optimal (or near-optimal) spatial design has received considerable attention in the statistical literature, often in the context of designing environmental monitoring systems \citep{DiggleLophaven2006,XiaEtAl2006,Muller2007}.
+Optimal designs can differ depend on whether one wishes to minimize a criterion such as integrated prediction variance, or a summary of the information matrix (i.e.
+if one wishes to optimize regression parameter performance).
+Optimal designs can also be sensitive to the structure of the estimation model that is used, so that tailoring a survey design to a particular model can be somewhat dangerous if there is uncertainty about the ultimate ``best" structure for the model used to relate animal abundance and occurrence to landscape or habitat covariates.
+Nevertheless, one could still think about augmenting a spatially balanced sampling design with a number of locations which are known or thought to have high leverage as a function of available covariates.
+We are excited about this prospect, and it is a subject of current research.
 
-Contrast with other approaches-
-Gelfand et al. Bayesian analysis - intrinsic CAR in SDM
-Chakraborty et al. '10 - spatial abundance modeling - ordinal
-Latimer et al. 2009 - spatial predictive process modelling in SPDs
-
-Can't be complacent... still possible to get poor/biased results, e.g. if $\tau_\epsilon \rightarrow 0$.  Can't resolve pathological problems.
-
-Presence-absence data, other link functions (e.g. probit)
-
-extensions to models w/ secondary observation process, measurement error
-
-much attention has been given to collinearity in multiple linear regression - suggest researchers give as much attention to predictive extrapolation bias in predictive models
-
-For predictions with spatial , our experience is that predictions outside the minimum convex polygon where data are obtained can sometimes be more problematic than predictions within the polygon.
-Spatial prediction surfaces may have a tendency to bend up or down in these areas, resulting in ``edge effects" that can lead to positive prediction bias when a log link function is employed \citep{VerHoefJansen2007}.
-
-Fieberg PVA
+With regards to ribbon seal abundance, we once again caution that the estimates presented here are for illustration only, as numerous factors that influence the detectability of seals was not included in count models used in this paper \citep[but see][]{ConnEtAl2014}.
+Our approach here was to examine extrapolation and prediction error using relatively simple models, with the understanding that such effects are also likely to occur in complex models with more realistic observation components.
+Indeed, our experience is that the magnitude of extrapolation bias can increase when additional data on imperfect detection or species misclassification are included.
 
 \bibliography{master_bib}
+
+\clearpage
 
@@
 FIGURE 2.
-Examples IVHs constructed from simulated data.
+Example IVHs constructed from simulated data.
 In panels A and B, the investigator plans to model a hypothetical (unmeasured) response variable using a linear regression model as a function of a single covariate, $x$, obtained at a number of design points (denoted with an ``x").
@@
 
-FIGURE 3. Aerial surveys over the Bering Sea in spring of 2012 (blue lines) overlayed on
-a tesselated surface composed of 25km by 25km grid cells.
-Gray indicates land, and colored pixels indicate ribbon seal encounters (yellow: 1-2 seals; orange: 3-4 seals; magenta: 5-9 seals; red: 10-15 seals).
-On average, photographs covered approximately $2.6km^2$ ($0.4\%$) of each surveyed grid cell.
+FIGURE 3.
+Depiction of a single simulation scenario.
+Panels (A-C) give simulated covariate values, panel D gives true animal abundance, (E) gives estimated abundance from a GLM run on count data from a spatially balanced survey design, and (F) gives abundance from a GLM applied to count data from a convenience survey.
+In (E-F), predictions outside the gIVH are represented by black boxes, and sampling locations are represented with an x.
+For the convenience sample, the median posterior abundance prediction for the entire survey area is $57\%$ greater than true abundance when inference is made to the whole study area.
+When inference is limited to the gIVH, median posterior abundance was just $16\%$ greater than true abundance.
 
 FIGURE 4.
+Boxplots summarizing distribution of proportional error in the posterior predictive median of abundance for the simulation study as a function of estimation model (x-axis), survey design (columns) and whether or not inference was restricted to the gIVH (rows).
+The lower and upper limits of each box correspond to first and third quartiles, while whiskers extend to the lowest and highest observed bias within 1.5 interquartile range units from the box.
+Outliers outside of this range are denoted with points.
+Horizontal lines within boxes denote median bias.
+The two numbers located below each boxplot indicate mean bias (upper number) and the number of additional outliers for which proportional bias was greater than 2.0 (lower number).
+
+FIGURE 5.
+Aerial survey tracks over the Bering Sea, April 22-29, 2012 (blue lines) overlayed on a tesselated study area consisting of 25km by 25km grid cells (gray lines).
+Dark gray indicates land, while the brown line indicates a 1000m depth contour, and the orange line shows the U.S Exclusive Economic Zone (EEZ) boundary (which was used to construct the easterly edge of the survey area).
+Colored pixels indicate ribbon seal counts along aerial transects.
+The average effective area surveyed in each grid cell was approximately $2.6km^2$ ($0.4\%$).
+Note that surveys were designed to target multiple seal species, several of which had high densities further north (results not shown).
+
+FIGURE 6.
 Potential covariates gathered to help explain and predict ribbon seal abundance in the eastern Bering Sea.
@@
 
-FIGURE 5  Posterior median estimates of ribbon seal apparent abundance across the eastern Bering sea for (A) a generalized linear model (GLM), (B) a generalized additive model (GAM), (C) a GLM with known zero data, and (D) a GAM with known zero data.
-Highlighted cells indicate those where predictive covariate values are outside of the generalized independent variable hull.
+FIGURE 7  Posterior mean predictions of ribbon seal apparent abundance across the eastern Bering sea from models fit to survey data.
+Each row gives result for different model types (GLM, GAM, or STRM, respectively); left column plots give results for naive runs without pseudo-absences, while plots in the right column give predictions for runs where psuedo-absence data (i.e., 0 counts in cells with less than $0.1\%$ ice) were included.
+Cells highlighted in black indicate those where predictions were outside the generalized independent variable hull (gIVH).
 
@@
 \begin{center}
-\includegraphics[width=6in]{test_flights_optimized.pdf}
+\includegraphics[width=6in]{sim_maps.pdf}
+\end{center}
+\caption{ }
+\label{fig:sim_maps}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=6in]{bias_generic.pdf}
+\end{center}
+\caption{ }
+\label{fig:sim_boxplots}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=6in]{BOSS_survey_ribbon.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=6in]{Bering_noRE_maps.pdf}
+\includegraphics[width=6in]{ribbon_maps.pdf}
 \end{center}
 \caption{ }
-\label{fig:covs}
+\label{fig:ribbon_plot}
 \end{figure}

### 2015-01-20 SpatPred/R/gIVH.R

@@ -1,6 +1,6 @@
 ### calculate prior prediction variance/gIVH via Monte Carlo
 
-gIVH <- function(Data,Effort,Assoc=NULL,my.formula,Tau.beta,Tau.eta,Tau.epsilon,srr.tol=0.8){
+gIVH <- function(Data,Effort,spat.mod=1,K=NULL,Assoc=NULL,my.formula,Tau.beta,Tau.eta,Tau.epsilon,srr.tol=0.8){
   n.reps=1000
   X.pred=model.matrix(my.formula,data=Data@data)
   X.obs=model.matrix(my.formula,data=Data@data[Effort$Mapping,]) 
@@ -9,7 +9,18 @@ gIVH <- function(Data,Effort,Assoc=NULL,my.formula,Tau.beta,Tau.eta,Tau.epsilon,
   n.beta=ncol(X.pred)
   I.beta=diag(nrow(XpXinv))
   
-  if(is.null(Assoc)==FALSE){
+  if(spat.mod==1 & is.null(K)==TRUE)cat("ERROR: User must provide K if spat.mod=1")
+  
+  if(spat.mod==1){ #define some matrices, etc. needed for PC model
+    n.knots=ncol(K)
+    K.t=t(K)
+    cross.K=crossprod(K)
+    Theta=rep(0,n.knots)
+    KpKinv=solve(cross.K) # add a little noise to make matrix non-singular
+    KpKinvKp=XpXinv%*%t(X.obs)
+  }
+  
+  if(spat.mod==2){
     Q=-Assoc
     diag(Q)=apply(Assoc,2,'sum')
     Q=Matrix(Q)  
@@ -41,7 +52,9 @@ gIVH <- function(Data,Effort,Assoc=NULL,my.formula,Tau.beta,Tau.eta,Tau.epsilon,
   Lambda.mc=matrix(0,S,n.reps)
   Var.lambda=matrix(0,S,n.reps)
   #Var.obs=matrix(0,n.obs,n.reps)
-  X.aug=cbind(X.pred,L.t)
+  X.aug=X.pred
+  if(spat.mod==1)X.aug=cbind(X.aug,K)
+  if(spat.mod==2)X.aug=cbind(X.aug,L.t)
   
   for(irep in 1:n.reps){
     tau.beta<-runif(1,Tau.beta[1],Tau.beta[2])
[…]

### 2015-01-20 SpatPred/R/sim_data_abundance.R

@@ -11,8 +11,11 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
   require(Matrix)
   require(spatstat)
   require(sp)
+  require(RandomFields)
   
   DEBUG=FALSE   
+  PLOT=FALSE
+  if(PLOT)set.seed(11111)
   tau.epsilon=100
   #if(DEBUG)tau.epsilon=100
   
@@ -22,52 +25,71 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
   
   x.len=sqrt(S)
  
-  #generate Covariates from a Matern distribution
-  Covs=data.frame(matrix(0,S,n.covs))
-  for(icov in 1:n.covs){
-    kappa=runif(1,5,10)
-    r=runif(1,0.2,.4)
-    mu=runif(1,200,500)
-    Dat.matern=rMatClust(kappa,r,mu)  
-    X=round((x.len)*Dat.matern$x+0.5)
-    Y=round((x.len)*Dat.matern$y+0.5)
-    X[X<1]=1
-    Y[Y<1]=1
-    X[X>x.len]=x.len
-    Y[Y>x.len]=x.len
-    Grid=matrix(0,x.len,x.len)
-    for(i in 1:length(X))Grid[X[i],Y[i]]=Grid[X[i],Y[i]]+1
-    Covs[,icov]=as.vector(Grid)/max(as.vector(Grid))
-    #Covs[,icov]=expit(logit(Covs[,icov])+rnorm(length(Grid),0,0.05)) # add a bit of noise
-    Which.0=which(Covs[,icov]==0)
-    if(length(Which.0)>0)Covs[,icov][Which.0]=abs(rnorm(length(Which.0),0,0.01))
-  } 
+  n.basis=10
+  n.zero=6
+  Omega=matrix(0,n.basis,S)
+  Cov=matrix(0,n.covs,S)
+  flag=1
+  while(flag==1){  #reject if any |cor|>0.75
+    for(i in 1:n.basis){
+      my.mod=RMexp(var=1,scale=runif(1,5,100))
+      Omega[i,]=RFsimulate(model=my.mod,x=c(1:x.len),y=c(1:x.len))$variable1
+    }
+    for(i in 1:n.covs){
+      Wts=matrix(rnorm(n.basis),1,n.basis)
+      Wts[sample(c(1:n.basis),n.zero)]=0
+      #cat(Wts)
+      #cat('\n')
+      Cov[i,]=Wts%*%Omega
[…]

### 2015-01-20 SpatPred/R/spat_pred.R

@@ -102,7 +102,7 @@ spat_pred<-function(formula,Data,Effort,spat.mod=0,Offset,Area.adjust,Control,As
     Alpha=rep(0,n.alpha)
     cross.K.gam=crossprod(K.gam[Mapping,])
     Alpha.mc=matrix(0,n.alpha,mcmc.length)
-    KpKinv.gam=solve(cross.K.gam)
+    #KpKinv.gam=solve(cross.K.gam)
   }
   Gamma=rep(0,S)  #combined smooth effects by site
   
@@ -137,6 +137,7 @@ spat_pred<-function(formula,Data,Effort,spat.mod=0,Offset,Area.adjust,Control,As
     Theta=rep(0,n.knots)
     KpKinv=solve(cross.K) # add a little noise to make matrix non-singular
     KpKinvKp=XpXinv%*%t(X.obs)
+    n.theta=n.knots
   }
   
   if(spat.mod==2){ #define some matrices, etc. needed for spatial updates under srr model
@@ -167,6 +168,8 @@ spat_pred<-function(formula,Data,Effort,spat.mod=0,Offset,Area.adjust,Control,As
     Theta=rnorm(n.theta,0,sqrt(1/tau.eta))
   }
   
+  if(spat.mod>0)Theta.mc=matrix(0,n.theta,mcmc.length)
+     
   if(is.null(Names.gam)==FALSE & n.beta==1){
     Beta[1]=0
     n.beta=0
@@ -254,7 +257,10 @@ spat_pred<-function(formula,Data,Effort,spat.mod=0,Offset,Area.adjust,Control,As
         Tau.alpha.mc[(iiter-Control$burnin)/Control$thin]=tau.alpha
         Alpha.mc[,(iiter-Control$burnin)/Control$thin]=Alpha
       }
-      if(spat.mod>=1)Tau.eta.mc[(iiter-Control$burnin)/Control$thin]=tau.eta
+      if(spat.mod>0){
+        Tau.eta.mc[(iiter-Control$burnin)/Control$thin]=tau.eta
+        Theta.mc[,(iiter-Control$burnin)/Control$thin]=as.numeric(Theta)
+      }
[…]

### 2015-01-20 SpatPred/R/util_funcs.R

@@ -118,15 +118,15 @@ d_biv_normal<-function(Tmp.vec,XY,Sigma){
 #' @param highlight If provided, this vector specifies which cells to separately highlight during plotting
 #' @param cell.width if highlight is provided, cell.width must provide the width of a composite grid cell
 #' @param leg.title Title for legend of plot (if different from the default "Abundance")
+#' @param hcolor If highight provided, gives color for highlighting (default = "yellow")
 #' @return An abundance map
 #' @export 
 #' @keywords abundance map, plot
 #' @author Paul Conn \email{paul.conn@@noaa.gov}
-plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance"){
+plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance",hcolor='yellow',cur.max=NULL){
   library(sp)
   library(RColorBrewer)
-  myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-  
+  myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))  
   Tmp=Grid[[1]]
   if(is.null(highlight)==FALSE){
     midpoints=data.frame(gCentroid(Tmp[highlight,],byid=TRUE))
@@ -138,10 +138,11 @@ plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundan
   new.colnames[1:2]=c("Easting","Northing")
   colnames(Cur.df)=new.colnames
   tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank())
-  p1=ggplot(Cur.df)+aes(Easting,Northing,fill=Abundance)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=myPalette(100))
+  if(is.null(cur.max))p1=ggplot(Cur.df)+aes(Easting,Northing,fill=Abundance)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=myPalette(100),name=leg.title)
+  else p1=ggplot(Cur.df)+aes(Easting,Northing,fill=Abundance)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=myPalette(100),name=leg.title,limits=c(0,cur.max))
   if(is.null(highlight)==FALSE){
[…]

### 2015-01-20 SpatPred/inst/analyze_sim_data.R

@@ -1,0 +1,158 @@
+### ANALYZE SIMULATION DATA FOR SPATIAL PREDICTION PAPER
+
+S=900
+n.sims=1000
+#Vars=matrix(0,S,3)
+#colnames(Vars)=c("GLM","GAM","RSR")
+Bias=matrix(0,n.sims,24)
+colnames(Bias)=c("GLM.sys.all","GAM.sys.all","RSR.sys.all","RSR2.sys.all",
+                 "GLM.sys.gIVH","GAM.sys.gIVH","RSR.sys.gIVH","RSR2.sys.gIVH",
+                 "GLM.sys.gIVH2","GAM.sys.gIVH2","RSR.sys.gIVH2","RSR2.sys.gIVH2",                 
+                 "GLM.clust.all","GAM.clust.all","RSR.clust.all","RSR2.clust.all",
+                 "GLM.clust.gIVH","GAM.clust.gIVH","RSR.clust.gIVH","RSR2.clust.gIVH",
+                 "GLM.clust.gIVH2","GAM.clust.gIVH2","RSR.clust.gIVH2","RSR2.clust.gIVH2")
+Ncell.gIVH=matrix(0,n.sims,8)
+colnames(Ncell.gIVH)=c("GLM.sys","GAM.sys","RSR.sys","RSR2.sys","GLM.clust","GAM.clust","RSR.clust","RSR2.clust")
+Ncell.gIVH2=matrix(0,n.sims,8)
+colnames(Ncell.gIVH2)=c("GLM.sys","GAM.sys","RSR.sys","RSR2.sys","GLM.clust","GAM.clust","RSR.clust","RSR2.clust")
+
+for(igen in 1:2){
+  for(isim in 1:n.sims){
+    cur.file=paste("e:/SpatPred/Sim_data/MCMC_gen",igen,"_sim",isim,".Rda",sep='')
+    load(cur.file)
+    cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    load(cur.file)
+    if(igen==1)cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
+    else cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
+    load(cur.file)
+        
[…]

### 2015-01-20 SpatPred/inst/plot_ribbon_IVH_by_tau.R

@@ -72,7 +72,7 @@ for(itau1 in 1:4){
   for(itau2 in 1:4){
     counter=counter+1
     Precision.pars=list(tau.epsilon=Tau.vec[itau1],tau.eta=Tau.vec[itau2])
-    Control$srr.tol=50
+    Control$srr.tol=0.5
     #Control$srr.tol=200
     Control$fix.tau.epsilon=FALSE
     #formula=Count~1

### 2015-01-20 SpatPred/inst/plot_ribbon_survey_ggplot.R

@@ -1,0 +1,146 @@
+### plot ribbon seal flights and count data using ggplot
+
+library(sp)
+library(rgdal)
+library(nPacMaps)  #from Josh London
+library(rgeos)
+library(doBy)
+library(ggplot2)
+library(plyr)
+
+#First, pull in data
+load('c:/users/paul.conn/git/STabundance/Data_for_ST_plot.Rdata')
+load('c:/users/paul.conn/git/SpatPred/BOSS_Power_photo_data/slr_power_data.RData') #sp data frame including species & area of photo
+load('c:/users/paul.conn/git/BOSS/BOSS/data/AlaskaBeringData.Rdat') #read in "Data" holding grid info
+load('c:/users/paul.conn/git/BOSS/BOSS/data/Effort_points.Rdat')  #read in Sp Points DF "effort_data" for on effort points (use for covariates)
+
+Cur.grid=Data$Grid[[1]]
+
+I.inter=gIntersects(slr_power_data,Data$Grid[[1]],byid=TRUE)
+Points=slr_power_data[which(apply(I.inter,2,'sum')>0),]  #only include effort points that are 'on grid'
+n.points=nrow(Points)
+
+#do some plotting
+plot(Data$Grid[[1]])
+points(Points,pch=19,cex=0.01,col='blue')
+
+n.cells=dim(Cur.grid)[1]
+S=nrow(Cur.grid)  #number of cells
+#first object corresponds to April 4 so start with 17th SpDF
+rownames(Cur.grid@data)=c(1:S)
+Cur.grid<-spChFIDs(Cur.grid,as.character(c(1:S)))  #change ID names to go from 1:S
+Cur.grid[["Ecoregion"]][which(Cur.grid[["Ecoregion"]]==19)]=20  #a few cells are from ecoreg 19 which is never sampled, so convert to 20
+Cur.grid[["Ecoregion"]]=factor(Cur.grid[["Ecoregion"]])
+
+
+data(alaska_dcw)
+data(russia_dcw)
+laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
+                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
+alaska_dcw <- spTransform(alaska_dcw, CRS(laea_180_proj))
+russia_dcw <- spTransform(russia_dcw, CRS(laea_180_proj))
+
+
[…]

### 2015-01-20 SpatPred/inst/plot_sim_example.R

@@ -1,0 +1,80 @@
+#plot maps from a single simulation run
+library(sp)
+library(rgeos)
+library(ggplot2)
+library(gridExtra)
+source("./spatpred/r/util_funcs.R")
+isim=51
+S=900
+
+#load data
+cur.file=paste("e:/SpatPred/Sim_data/MCMC_gen",1,"_sim",isim,".Rda",sep='')
+load(cur.file)
+MCMC.sys=MCMC
+cur.file=paste("e:/SpatPred/Sim_data/MCMC_gen",2,"_sim",isim,".Rda",sep='')
+load(cur.file)
+MCMC.clust=MCMC
+cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+load(cur.file)
+cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
+load(cur.file)
+Effort.sys=Effort
+cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
+load(cur.file)
+Effort.clust=Effort
+
+#plot covariates
+
+##### plot_N_map needs cur.max to be defined and in the global environment to get the N plots on the right scale
+cur.max=NULL
+Grid.list=vector("list",1)
+Grid.list[[1]]=Grid
+Cov1 = plot_N_map(1,matrix(Grid@data$cov1,S,1),Grid=Grid.list,leg.title="Value")+ggtitle("A. Covariate 1")+theme(plot.title = element_text(hjust = 0,size = rel(4)),legend.text=element_text(size=rel(3)),legend.title=element_text(size=rel(3)),text=element_text(size=rel(3)))+xlab('')+ylab('')
+Cov2 = plot_N_map(1,matrix(Grid@data$cov2,S,1),Grid=Grid.list,leg.title="Value")+ggtitle("B. Covariate 2")+theme(plot.title = element_text(hjust = 0,size = rel(4)),legend.text=element_text(size=rel(3)),legend.title=element_text(size=rel(3)),text=element_text(size=rel(3)))+xlab('')+ylab('')
[…]

### 2015-01-20 SpatPred/inst/plot_survey_designs.R

@@ -3,7 +3,7 @@
 S=900
 XY=expand.grid(y=c(sqrt(S):1),x=c(1:sqrt(S)))
 ids.reg=factor(c(1:50))
-ids.clust=factor(c(1:60))
+ids.clust=factor(c(1:50))
 ids.IVH=ids.reg
 isim=1
 
@@ -18,12 +18,20 @@ Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
 load(Cur.file) #load Effort
 Effort.clust=Effort
 
-Cur.file=paste("./Sim_data/Effort_IVH1",isim,".Rda",sep='')
-load(Cur.file) #load Effort
-Effort.IVH=Effort
+
 
 
 Data=Grid
+
+#calculate knot-cell distances for process convolution 
+Knot.locations=expand.grid(x=c(1:6)*7-9,y=c(1:6)*7-9)
+n.knots=nrow(Knot.locations)
+#calculate kernel densities at grid cell centroids 
+Distances=matrix(0,S,n.knots)
+for(iS in 1:S)Distances[iS,]=sqrt((Knot.locations[,1]-XY[iS,"x"])^2+(Knot.locations[,2]-XY[iS,"y"])^2)
+K=matrix(dnorm(Distances,0,7),S,n.knots)  #knot sd=5 
+K=K/rowSums(K)        
+
 
 #compute gIVH for different designs
 
@@ -63,7 +71,7 @@ dev.off()
 
 #my.formula=~(cov1+cov2+cov3)^2+cov1.quad+cov2.quad+cov3.quad
 my.formula=~1
-gIVH.cluster=gIVH(Data=Data,Effort=Effort.clust,Assoc=Assoc,my.formula=my.formula,Tau.beta=c(0.1,2),Tau.eta=c(0.1,10),Tau.epsilon=c(99,100),srr.tol=0.8)
+gIVH.cluster=gIVH(Data=Data,Effort=Effort.clust,Assoc=NULL,K=K,my.formula=my.formula,Tau.beta=c(0.1,2),Tau.eta=c(0.1,10),Tau.epsilon=c(99,100),srr.tol=0.8)
 df<-data.frame(id=ids.clust,Easting=XY[Effort.clust$Mapping,"x"],Northing=XY[Effort.clust$Mapping,"y"])
 Abund=log(gIVH.cluster$Var.pred)
 #Abund=gIVH.cluster$IVH

### 2015-01-20 SpatPred/inst/run_abundance_sims.R

@@ -2,6 +2,7 @@
 # script to run generic spatio-temporal count data simulations
 source("./SpatPred/R/sim_data_abundance.R")
 source("./SpatPred/R/util_funcs.R")
+source("./SpatPred/R/spat_pred.R")
 
 set.seed(123456)
 n.sims=1#number of simulations at each design point
@@ -21,7 +22,7 @@ if(i.sim){
   for(isim in 1:n.sims){
     Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
     load(Cur.file)
-    Effort=sim_effort(Data=Grid,S=S,formula=NULL,type="random",prop.sampled=prop.sampled,n.points=n.transects)
+    Effort=sim_effort(Data=Grid,S=S,type="random",prop.sampled=prop.sampled,n.points=n.transects)
     Cur.file=paste("./Sim_data/Effort_random",isim,".Rda",sep='')
     save(Effort,file=Cur.file)
   }
@@ -34,7 +35,7 @@ if(i.sim){
 
 
 #call estimation routines
-Adj=square_adj(sqrt(S))
+Adj=rect_adj(sqrt(S),sqrt(S))
 for(igen in 1:1){  #loop over generating model to generate data sets
   for(isim in 1:n.sims){  
     Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
@@ -55,7 +56,7 @@ for(igen in 1:1){  #loop over generating model to generate data sets
     
     #estimation model 1: GLM
     formula=~(cov1+cov2+cov3)^2+cov1.quad+cov2.quad+cov3.quad
-    #formula=~cov1+cov1.quad
+    #formula=~cov1+cov2+cov3
     spatmod=0
     Control=list(iter=1000,burnin=500,thin=100,srr.tol=0.5,predict=TRUE,MH.nu=rep(0.2,n.transects),adapt=TRUE,fix.tau.epsilon=FALSE,Kern.gam.se=NULL)
     MCMC=spat_pred(formula=formula,Data=Grid,Effort=Effort,spat.mod=0,Offset=Offset,Area.adjust=Area.adjust,Control=Control,Prior.pars=NULL)
@@ -65,7 +66,7 @@ for(igen in 1:1){  #loop over generating model to generate data sets
     Grid.list=vector("list",1)
     Grid.list[[1]]=Grid
[…]

### 2015-01-20 SpatPred/inst/run_design_sims.R

@@ -2,24 +2,24 @@
 # script to run generic spatio-temporal count data simulations
 source("./SpatPred/R/sim_data_abundance.R")
 source("./SpatPred/R/util_funcs.R")
+source("./SpatPred/R/spat_pred.R")
 
 set.seed(123456)
-n.sims=1#number of simulations at each design point
+n.sims=250 #number of simulations at each design point
 S=900
 prop.sampled=0.1
 n.transects=60
 
-i.sim=TRUE
+i.sim=FALSE
 if(i.sim){
   for(isim in 1:n.sims){
     #simulate covariates, abundance
-    Grid=sim_data_generic(S=S,n.covs=4,tau.epsilon=10)
+    Grid=sim_data_generic(S=S,n.covs=3,tau.epsilon=10)
     Grid$cov1.quad=Grid$cov1^2
     Grid$cov2.quad=Grid$cov2^2
     Grid$cov3.quad=Grid$cov3^2
     Grid$cov12=Grid$cov1*Grid$cov2
     Grid$cov13=Grid$cov1*Grid$cov3
-    Grid$cov23=Grid$cov2*Grid$cov3
     Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
     save(Grid,file=Cur.file)
   }
@@ -27,42 +27,42 @@ if(i.sim){
   for(isim in 1:n.sims){
     Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
     load(Cur.file)
-    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="balanced",prop.sampled=prop.sampled,n.points=50)
+    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="balanced",prop.sampled=prop.sampled,n.points=40)
     Cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
     save(Effort,file=Cur.file)
     Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="clustered",prop.sampled=prop.sampled,n.points=60)    
     Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
     save(Effort,file=Cur.file)
-    my.formula=~cov1+cov1.quad
[…]

### 2015-01-20 SpatPred/inst/run_design_simsA.R

@@ -1,0 +1,176 @@
+# run_generic_sims.R
+# script to run generic spatio-temporal count data simulations
+source("./SpatPred/R/sim_data_abundance.R")
+source("./SpatPred/R/util_funcs.R")
+source("./SpatPred/R/spat_pred.R")
+
+set.seed(123456)
+n.sims=250 #number of simulations at each design point
+S=900
+prop.sampled=0.1
+n.transects=45
+
+i.sim=FALSE
+if(i.sim){
+  for(isim in 1:n.sims){
+    #simulate covariates, abundance
+    Grid=sim_data_generic(S=S,n.covs=3,tau.epsilon=10)
+    Grid$cov1.quad=Grid$cov1^2
+    Grid$cov2.quad=Grid$cov2^2
+    Grid$cov3.quad=Grid$cov3^2
+    Grid$cov12=Grid$cov1*Grid$cov2
+    Grid$cov13=Grid$cov1*Grid$cov3
+    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    save(Grid,file=Cur.file)
+  }
+  #now simulate count datasets 
+  for(isim in 1:n.sims){
+    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    load(Cur.file)
+    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="balanced",prop.sampled=prop.sampled,n.points=45)
+    Cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
+    save(Effort,file=Cur.file)
+    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="clustered",prop.sampled=prop.sampled,n.points=45)    
+    Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
+    save(Effort,file=Cur.file)
+    #my.formula=~cov1+cov1.quad
+    #Effort=sim_effort(Data=Grid,S=S,my.formula=my.formula,type="IVH",prop.sampled=prop.sampled,n.points=50)    
+    #Cur.file=paste("./Sim_data/Effort_IVH1",isim,".Rda",sep='')
+    #save(Effort,file=Cur.file)
+    #my.formula=~(cov1+cov2)^2+cov1.quad+cov2.quad
[…]

### 2015-01-20 SpatPred/inst/run_ribbon_models_Jan2015.R

@@ -1,0 +1,301 @@
+source("./SpatPred/R/spat_pred.R")
+source("./SpatPred/R/util_funcs.R")
+require(ggplot2)
+load('Ribbon_data.Rda')
+load('Knot_cell_distances.Rdata')
+K=K.data$K
+set.seed(12345)
+
+#also include quadratic term for ice_conc in dataset
+ice_conc2=(Cur.grid@data[["ice_conc"]])^2
+dist_edge2=(Cur.grid@data[["dist_edge"]])^2
+dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
+dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
+Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
+#attach Ribbon seal counts to data frame
+Count=Ribbon.count
+Cur.grid@data=cbind(Cur.grid@data,Count)
+Which.not.sampled=c(1:S)
+Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
+
+#add in some zero counts in cells with no ice
+Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
+Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
+#Mapping=c(Mapping,Which.no.ice)
+#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
+
+#run glm model
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
+#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
+#formula=Count~1
+#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~dist_mainland+ice_conc+ice_conc2
+formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
+#formula=Count~ice_conc+ice_conc2
+#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
+#formula=Count~dist_shelf+dist_shelf2
+#formula=Count~dist_edge+dist_edge2
[…]

### 2015-01-20 SpatPred/inst/run_ribbon_models_naive.R

@@ -1,4 +1,5 @@
 source("./SpatPred/R/spat_pred.R")
+source("./SpatPred/R/util_funcs.R")
 load('Ribbon_data.Rda')
 load('Knot_cell_distances.Rdata')
 K=K.data$K
@@ -37,7 +38,7 @@ formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
 
 
 
-Control=list(iter=11000,burnin=1000,thin=2,adapt=TRUE,srr.tol=NULL,fix.tau.epsilon=FALSE,predict=TRUE,MH.nu=rep(0.2,length(Mapping)))
+Control=list(iter=11000,burnin=1000,thin=2,adapt=TRUE,srr.tol=NULL,fix.tau.epsilon=FALSE,predict=TRUE,MH.nu=rep(0.5,length(Mapping)))
 Data=Cur.grid
 spat.mod=0
 Offset=Area.photo
@@ -48,13 +49,24 @@ Prior.pars=NULL
 #formula=Count~Ecoregion
 #formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge+Ecoregion
 #formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-glm.out <- spat_pred(formula=formula,Data=Data,spat.mod=spat.mod,Offset=Offset,Mapping=Mapping,Area.adjust=Area.adjust,Control=Control,Assoc=Assoc,Knots=Knots,Prior.pars=Prior.pars)
+Effort=list(Mapping=Mapping,Counts=Data$Count[Mapping])
+glm.out <- spat_pred(formula=formula,Data=Data,spat.mod=spat.mod,Offset=Offset,Effort=Effort,Area.adjust=Area.adjust,Control=Control,Assoc=Assoc,Knots=Knots,Prior.pars=Prior.pars)
 Control=glm.out$Control
 Control$adapt=FALSE
 Control$iter=60000
 Control$burnin=10000
 Control$thin=10
-glm.out.cov <- spat_pred(formula=formula,Data=Data,spat.mod=spat.mod,Offset=Offset,Mapping=Mapping,Area.adjust=Area.adjust,Control=Control,Assoc=Assoc,Knots=Knots,Prior.pars=Prior.pars)
[…]

### 2015-01-20 SpatPred/inst/sim_pred_process_scratch.R

@@ -1,0 +1,20 @@
+
+require(fields)
+
+S=900
+XY=expand.grid(y=c(sqrt(S):1),x=c(1:sqrt(S)))
+Knot.locations=expand.grid(x=c(1:6)*7-9,y=c(1:6)*7-9)
+n.knots=nrow(Knot.locations)
+
+
+
+#calculate kernel densities at grid cell centroids 
+Distances=matrix(0,S,n.knots)
+
+C.star=Exp.cov(Knot.locations,theta=10)
+CrossC=Exp.cov(Knot.locations,XY,theta=10)
+K=crossprod(CrossC,solve(C.star))
+
+#for(iS in 1:S)Distances[iS,]=sqrt((Knot.locations[,1]-XY[iS,"x"])^2+(Knot.locations[,2]-XY[iS,"y"])^2)
+#K=matrix(dnorm(Distances,0,7),S,n.knots)  #knot sd=5 
+#K=K/rowSums(K)        

### 2015-03-24 manuscript/AppendixA.tex
SpatPred draft to Peter and Devin for first time - doesn't have Steller example integrated into text yet
Cleaned up and compiled R package.  Added gIVH function that computes gIVH for a generic model using same procedure used in analysis.  Manuscript is as initially submitted to Ecology, March 2015

@@
 \begin{center}
-Appendix A: Model formulation and Gibbs sampling algorithms for certain classes and extensions of the generalized linear model
+Appendix A: Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model
 \bigskip
@@
 \section{Models}
-\subsection{Generalized linear models (GLM)}
+\subsection{Generalized linear models (GLMs)}
 
@@
 
-\subsection{Generalized additive models (GAM)}
+\subsection{Generalized additive models (GAMs)}
 
@@
 The basic notion is to place $k_j$ knots, which we will denote by $\boldsymbol{\omega}_j = \{ \omega_{1j},\omega_{2j},\hdots,\omega_{k_j j} \}$, throughout the range of a given regressor $j$, where the number of knots controls the level of smoothing.
-For each regressor of interest, we can then calculate a matrix ${\bf K}_j$ of dimension $(n,k_j)$, with entries $\mathcal{N}(X_{ij};\omega_j,\tau_{\omega,j})$, where $\mathcal{N}(x;\mu,\tau)$ denotes a Gaussian distribution with mean $\mu$ and precision $\tau$.
+For each regressor of interest, we can then calculate a matrix ${\bf K}_j$ of dimension $(n,k_j)$ (where $n$ gives the number of observed data locations), with entries $z_{ik}=\mathcal{N}(X_{ij};\omega_{kj},\tau_j)$, where $\mathcal{N}(x;\mu,\tau)$ denotes a Gaussian distribution with mean $\mu$ and precision $\tau$, $i$ indexes a particular observed data location, and $k$ denotes the $k$th knot for covariate $j$, $k \in \{1,2,\hdots,k_j \}$.
 We can then introduce additional regression coefficients $\boldsymbol{\alpha}_j$ to help model the smooth relationship, and incorporate these into the linear predictor.
@@
 
-There are a variety of ways spatial autocorrelation can be included in regression models, depending on (i) the spatial support (i.e., continuous vs.
+\hspace{.5in}There are a variety of ways spatial autocorrelation can be included in regression models, depending on (i) the spatial support (i.e., continuous vs.
 discrete), and (ii) the particular mechanism used to impart correlation.
@@
 \end{linenomath*}
-where $\boldsymbol{\eta}$ represent spatially autocorrelated effects.  In this treatment we shall consider two approaches for
-inducing spatial autocorrelation in $\boldsymbol{\eta}$: process convolution
-\citep[PC;][]{Higdon1998} and restricted spatial regression \citep[RSR;][]{Reich2006,Hodges2010,Hughes2013}.
+where $\boldsymbol{\eta}$ represent spatially autocorrelated effects.  In this treatment we shall limit consideration one approach for
+inducing spatial autocorrelation in $\boldsymbol{\eta}$: restricted spatial regression \citep[RSR;][]{Reich2006,Hodges2010,Hughes2013}.
 
-The PC implementation works by placing $k$ knots throughout the spatial domain being analyzed, and like our GAM formulation, uses distances from the center point of each sample unit to each of these knot locations to induce spatial structure.
-Using similar notation to our GAM formulation, we denote these knots as $\boldsymbol{\omega} = \{ \boldsymbol{\omega_{1}},\boldsymbol{\omega}_{2},\hdots,\boldsymbol{\omega}_{k} \}$; the $\boldsymbol{\omega}_m$ are bivariate in this  instance as space is assumed to be two dimensional.
-We then construct a matrix ${\bf K}$ of dimension $(n,k)$ with entries $\mathcal{BVN}({\bf s}_i;\boldsymbol{\omega}_m,\tau_{\omega})$, where $\mathcal{BVN}({\bf x};{\bf \mu},\tau)$ denotes a bivariate Gaussian distribution with mean $\boldsymbol{\mu}$ and precision $\tau$.
-As with our GAM model we introduce additional regression coefficients $\boldsymbol{\alpha}$ to estimate the level of spatial smoothing conditional on the assumed knot structure and reparameterize the STRM as
-\begin{linenomath*}
-\begin{equation}
-  \label{eq:PC}
-  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + {\bf K} \boldsymbol{\alpha}
-\end{equation}
-\end{linenomath*}
-This model is structurally very similar to the GAM model (indeed it may be interpreted as a GAM with a bivariate smooth on spatial location).
-Similarly, we can rearrange terms to write it similar to the GLM, such that
-$\boldsymbol{\xi}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
-where in this case ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K} \right]$, and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha} \right] ^\prime$.
-
-The RSR approach to spatial regression uses a reduced-rank version of the popular intrinsic conditionally autoregressive \citep[ICAR;][]{Besag1995,Rue2005} model for spatial random effects, reparameterized so that basis vectors are orthogonal to the main effects of interest.
+\hspace{.5in}The RSR approach to spatial regression uses a reduced-rank version of the popular intrinsic conditionally autoregressive \citep[ICAR;][]{Besag1995,RueHeld2005} model for spatial random effects, reparameterized so that basis vectors are orthogonal to the main effects of interest.
 This approach has generated substantial recent interest, as fixed effects retain primacy in explaining variation in the ecological process of interest and problems with spatial confounding between fixed and random effects are eliminated.
@@
   \item Define the residual projection matrix ${\bf P}^\perp={\bf I}-{\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$
-  \item Calculate the Moran operator matrix $\boldsymbol{\Omega}=J{\bf P}^\perp \mathcal{W}{\bf P}^\perp/{\bf 1}^\prime \mathcal{W} {\bf 1}$
+  \item Calculate the Moran operator matrix $\boldsymbol{\Omega}=J{\bf P}^\perp {\bf W}{\bf P}^\perp/{\bf 1}^\prime {\bf W} {\bf 1}$, where $J$ is the number of areal survey units
   \item Define ${\bf K}$ as an $(n \times m)$ matrix, where the columns of ${\bf K}$ are composed of the eigenvectors  associated with the largest $m$ eigenvalues of $\boldsymbol{\Omega}$.
@@
 \end{enumerate}
-Here, ${\bf I}$ an identity matrix, ${\bf 1}$ is a column vector of ones, and $\mathcal{W}$ represents an association matrix describing the spatial neighborhood structure of sampling units.
-For instance, for a first order neighborhood structure, $\mathcal{W}$ would include a 1 for all rows $i$ and columns $j$ where sampling unit $i$ and $j$ are neighbors \citep[see][for alternative association matrices]{Rue2005}.
+Here, ${\bf I}$ an identity matrix, ${\bf 1}$ is a column vector of ones, and ${\bf W}$ represents an association matrix describing the spatial neighborhood structure of sampling units.
+For instance, for a first order neighborhood structure, ${\bf W}$ would include a 1 for all rows $i$ and columns $j$ where sampling unit $i$ and $j$ are neighbors \citep[see][for alternative association matrices]{RueHeld2005}.
 
@@
 
-Setting the prior for the suite of regression parameters as $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$ results in the full conditional distribution
+Setting the prior for the suite of regression parameters as $[\boldsymbol{\beta}] = \mathcal{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$ results in the full conditional distribution
 \begin{linenomath*}
@@
 
-\subsection{Gibbs sampler for the PC STRM model}
+\subsection{Gibbs sampler for the STRM model}
 
-The Gibbs sampler for the process convolution version of the STRM model consists of the following:
+The Gibbs sampler for the RSR version of the STRM model consists of the following:
 
@@
 
-We impose the same Gaussian prior on smoothing kernel weights, $\boldsymbol{\alpha} \sim \mathcal{N}(0,\tau_\alpha^{-1})$ as for the GAM.  The resulting full conditional is simply
-\begin{linenomath*}
-\begin{equation*}
-   [\boldsymbol{\alpha} | \cdot] \equiv \mathcal{MVN}(({\bf K}^\prime {\bf K})^{-1}{\bf K}^\prime (\boldsymbol{\mu}-{\bf X}\boldsymbol{\beta}),({\bf K}^\prime {\bf K}^{-1} (\tau_\epsilon + \tau_\alpha)^{-1}).
-\end{equation*}
-\end{linenomath*}
-
-\subsection{Gibbs sampler for the RSR STRM model}
-
-Although ${\bf K}$ is defined differently in the RSR model, the Gibbs sampler remains the same as for the the PC model, with the exception of a modified full conditional for $[\boldsymbol{\alpha}|\cdot]$ owing to a different prior specification for $\boldsymbol{\alpha}$.
-We must also update an additional precision parameter associated with spatial random effects, $\tau_\eta$.
-
-\underline{Sampling from $[\boldsymbol{\alpha}|\cdot]$}
-
-In the RSR model, the implied prior for reduced dimension spatial effects is $\boldsymbol{\alpha} \sim \text{MVN}(0,\tau_\eta^{-1}{\bf K}^\prime {\bf Q} {\bf K})$.  The resulting full conditional is
+In the RSR model, the implied prior for reduced dimension spatial effects is $\boldsymbol{\alpha} \sim \mathcal{MVN}(0,\tau_\eta^{-1}{\bf K}^\prime {\bf Q} {\bf K})$.
+The resulting full conditional is
 \begin{linenomath*}
@@
 
+\subsection{Calculating the gIVH}
+
+In order to calculate prediction variance on the real scale for generalized models, it is necessary to use a procedure such as the delta method \citep{Dorfman1938}.
+In particular, we can write the variance of the Poisson expectation as
+\begin{linenomath*}
+\begin{eqnarray}
+  \label{eq:var-lambda}
+  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
+  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
+%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
+%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
+%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
+%      \vdots & \vdots & \ddots & \vdots \\
+%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
+%  \end{array} \right].
+\end{eqnarray}
+\end{linenomath*}
+where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).
+Under common univariate link functions (e.g.
+log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g.
+multinomial logit) $\boldsymbol{\Delta}$ will be dense.
+
+\hspace{.5in} Unlike Cook's original formulation, the gIVH will often involve unknown parameters.
+Further, the exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the underlying model structure and estimation procedure.
+We impose a probability distribution $[\boldsymbol{\theta}]$ for unknown parameters, and compute prediction variance as
+\begin{linenomath*}
+\begin{equation}
+  \label{eq:E_var_lambda}
+  {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}) = \int_{\boldsymbol{\theta}} {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}, \boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
+\end{equation}
+\end{linenomath*}
+
+\hspace{.5in} Our approach in this paper is to conduct a Bayesian analysis (using the Gibbs sampling algorithms outlined above) and substitute the posterior distribution $[\boldsymbol{\theta}|{\bf Y}]$ for $[\boldsymbol{\theta}]$ in Eq.
+\ref{eq:E_var_lambda}.
+Under this approach, one simply sets ${\rm Var}(\mu_x)$ equal to the posterior variance of $\mu_x$ (across Markov chain iterations).
+In the case of the log link function, one also needs an estimate for $\mu_x$ to use in delta method calculations; we used the posterior median for each sampling unit.
+Although we use a Bayesian formulation for modeling convenience, one could also consider substituting parametric approximations for ${\rm Var}(\mu_x)$ and $\mu_x$ based on output from standard likelihood-based analyses.
+
 \renewcommand{\refname}{Literature Cited}

### 2015-03-24 manuscript/AppendixB.tex
Cleaned up and compiled R package.  Added gIVH function that computes gIVH for a generic model using same procedure used in analysis.  Manuscript is as initially submitted to Ecology, March 2015

@@
 We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where predictions of animal abundance using these models may be problematic, and to determine whether statistical inferences are more robust when restricted to sampling units within the gIVH.
-Each simulation replicate consisted of several steps, including
+Each simulation replicate consisted of several steps (Fig.
+\ref{}), including
 \begin{enumerate}
@@
 \end{enumerate}
-We now describe each of these tasks and results of the simulation study in further detail.
+We now describe each of these tasks and results of the simulation study in further detail before describing results.
 All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the R package \texttt{SpatPred} that accompanies this article.
@@
 
+As suggested in Appendix A, we used Eq. A.7 to calculate the gIVH, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.
+
+\centerline{\sc Results}
+
+Posterior predictions from simulations indicated that the distribution for proportional error in total abundance was right skewed when statistical inference was made with regard to the entire survey area (Fig.
+\ref{fig:sim_boxplots}).
+This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.
+The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.
+Interestingly, the convolution kernel GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.
+For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude (e.g.
+$\approx 0.3$; Fig.
+\ref{fig:sim_boxplots}) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
+
 \renewcommand{\refname}{Literature Cited}
@@
 \end{flushleft}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=6in]{sim_maps.pdf}
+\end{center}
+\caption{Depiction of a single simulation scenario.
+Panels (A-C) give simulated covariate values, panel D gives true animal abundance, (E) gives estimated abundance from a GLM run on count data from a spatially balanced survey design, and (F) gives abundance from a GLM applied to count data from a convenience survey.
+In (E-F), predictions outside the gIVH are represented by black boxes, and sampling locations are represented with an x.
+For the convenience sample, the median posterior abundance prediction for the entire survey area is $57\%$ greater than true abundance when inference is made to the whole study area.
+When inference is limited to the gIVH, median posterior abundance was just $16\%$ greater than true abundance.}
+\label{fig:sim_maps}
+\end{figure}
 
@@
 
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=6in]{bias_generic.pdf}
+\end{center}
+\caption{Boxplots summarizing distribution of proportional error in the posterior predictive median of abundance for the simulation study as a function of estimation model (x-axis), survey design (columns) and whether or not inference was restricted to the gIVH (rows).
+The lower and upper limits of each box correspond to first and third quartiles, while whiskers extend to the lowest and highest observed bias within 1.5 interquartile range units from the box.
+Outliers outside of this range are denoted with points.
+Horizontal lines within boxes denote median bias.
+The two numbers located below each boxplot indicate mean bias (upper number) and the number of additional outliers for which proportional bias was greater than 2.0 (lower number).
+}
+\label{fig:sim_boxplots}
+\end{figure}
+
 \end{document}

### 2015-03-24 manuscript/AppendixC.tex
Cleaned up and compiled R package.  Added gIVH function that computes gIVH for a generic model using same procedure used in analysis.  Manuscript is as initially submitted to Ecology, March 2015

\documentclass[12pt,fleqn]{article}
%\documentclass[12pt,a4paper]{article}
\usepackage{natbib}
\usepackage{lineno}
%\usepackage{lscape}
%\usepackage{rotating}
%\usepackage{rotcapt, rotate}
\usepackage{amsmath,epsfig,epsf,psfrag}
\usepackage{setspace}
\usepackage{ulem}
\usepackage{xcolor}
\usepackage[labelfont=bf,labelsep=period]{caption} %for making figure and table numbers bold
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\hypersetup{pdfpagemode=UseNone}

%\usepackage{a4wide,amsmath,epsfig,epsf,psfrag}

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bX{{\ensuremath\mathbf{X}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\renewcommand{\refname}{\hspace{2.3in} \normalfont \normalsize LITERATURE CITED}
%this tells it to put 'Literature Cited' instead of 'References'
\bibpunct{(}{)}{,}{a}{}{;}
\oddsidemargin 0.0in
\evensidemargin 0.0in
\textwidth 6.5in
\headheight 0.0in
\topmargin 0.0in
\textheight=9.0in
%\renewcommand{\tablename}{\textbf{Table}}
%\renewcommand{\figurename}{\textbf{Figure}}
\renewcommand{\em}{\it}
\renewcommand\thefigure{C\arabic{figure}}
\renewcommand\thetable{C\arabic{table}}
\renewcommand\theequation{C.\arabic{equation}}

\begin{document}

\begin{center} \bf {\large On extrapolating past the range of observed data when making statistical predictions in ecology}

\vspace{0.7cm}
Paul B. Conn$^{1*}$, Devin S. Johnson$^1$, and Peter L. Boveng$^1$
\end{center}
\vspace{0.5cm}

\rm
\small

\it $^1$National Marine Mammal Laboratory, Alaska Fisheries Science Center,
NOAA National Marine Fisheries Service,
Seattle, Washington 98115 U.S.A.\\

\rm \begin{flushleft}

\raggedbottom
\vspace{.5in}

\begin{center}
Appendix C: Additional details on ribbon seal count data and analyses
\bigskip
\end{center}
\vspace{.3in}

\doublespacing

As part of an international effort, researchers with the U.S. National Marine Fisheries Service conducted aerial surveys over the eastern Bering Sea in 2012 and 2013.  Agency scientists used infrared video to detect seals that were on ice, and collected simultaneous digital photographs to provide information on species identity. For this study, we use spatially referenced count data from photographed ribbon seals, {\it Phoca fasciata} on a subset of 10 flights flown over the Bering Sea from April 20-27, 2012.  We limited flights to a one week period because sea ice melts rapidly in the Bering Sea in the spring, and modeling counts over a longer duration would likely require addressing how sea ice and seal abundance changes over both time and space \citep[see][]{ConnEtAl2015}. However, limiting analysis to a one week period makes the assumption of static sea ice and seal densities tenable \citep{ConnEtAl2014}.

\hspace{.5in} Our objective with this dataset will be to model seal counts on transects through 25km by 25km grid cells as a function of habitat covariates and possible spatial autocorrelation. Estimates of apparent abundance can then be obtained by summing predictions across grid cells. Figure \ref{fig:covs} show explanatory covariates gathered to help predict ribbon seal abundance.  These data are described in fuller detail by \citet{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors.  Since our focus in this paper is on illustrating spatial modeling concepts, we devote our efforts to the comparably easier problem of estimating apparent abundance (i.e., uncorrected for vagaries of the detection process).

\hspace{0.5in} We fit GLMs, an STRMs, and GAMs to the ribbon seal data.  The fixed effects component of the GLM and STRM included linear effects of all explanatory covariates (Fig. \ref{fig:covs}, as well as a quadratic effect for sea ice concentration.  For the STRM, we imposed a restricted spatial regression (RSR) formulation for spatially autocorrelated random effects, where dimension reduction was accomplished by only selecting eigenvectors of the spectral decomposition associated with eigenvalues that were greater than 0.5 (see Appendix A for additional information on model structure).  For the GAM, we employed smooth terms with 8 knots for each covariate.  This included 4 knots at the minimum, 1/3 and 2/3 quantiles, and the maximum observed covariate values, in addition to 4 knots above and below the range of observed data to reduce the magnitude of known biased endpoint behavior that can occur with radial basis function implementations of GAMs.

As with simulated data examples, we adopted a Bayesian perspective and conducted estimation using MCMC (see Appendix A for algorithm details and information on prior distributions) with 60,000 iterations where the first 10,000 iterations were discarded as a burn-in.  We generated posterior predictions of ribbon seal abundance across the landscape as
\begin{equation}
  N_i \sim {\rm Poisson}(A_i \lambda_i),
\end{equation}
and calculated the gIVH as in Eqn. \ref{eq:gIVH}, with delta method modifications as specified in Eqns. \ref{eq:var-mu}-\ref{eq:E_var_lambda}.

\renewcommand{\refname}{Literature Cited}
\bibliographystyle{ecology}

\bibliography{master_bib}

\end{flushleft}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6in]{covariates_col.pdf}
\end{center}
\caption{Assembled covariates used to help explain and predict ribbon seal relative abundance in the eastern Bering Sea.  Covariates include distance from mainland (\texttt{dist\_mainland}), distance from 1000m depth contour (\texttt{dist\_shelf}), average remotely sensed sea ice concentration while surveys were being conducted (\texttt{ice\_conc}), and distance from the southern sea ice edge (\texttt{dist\_edge}).  All covariates except ice concentration were standardized to have a mean of 1.0 prior to plotting and analysis.
}
\label{fig:covs}
\end{figure}

\end{document}

### 2015-03-24 manuscript/SpatPred_report.tex
Cleaned up and compiled R package.  Added gIVH function that computes gIVH for a generic model using same procedure used in analysis.  Manuscript is as initially submitted to Ecology, March 2015

--- 2015-01-20 manuscript/SpatPred3.tex
@@
 
+\bibpunct{(}{)}{,}{a}{}{;}
+
 \bibliographystyle{ecology}
@@
 
-{\em Abstract.\ } We'll do this later
+{\em Abstract.\ } Ecologists are increasingly using statistical models to predict animal abundance and occurrence in unsampled locations.
+The reliability of such predictions depends on a number of factors, including sample size, how far prediction locations are from the observed data, and similarity of predictive covariates in locations where data are gathered to locations where predictions are desired.
+In this paper, we propose extending Cook's notion of an independent variable hull (IVH), developed originally for application with linear regression models, to generalized regression models as a way to help assess the potential reliability of predictions in unsampled areas.
+Predictions occurring inside the generalized independent variable hull (gIVH) can be regarded as interpolations, while predictions occurring outside the gIVH can be regarded as extrapolations worthy of additional investigation or skepticism.
+We conduct a simulation study to demonstrate the usefulness of this metric for limiting the scope of spatial inference when conducting model-based abundance estimation from survey counts.
+In this case, limiting inference to the gIVH substantially reduces bias, especially when survey designs are spatially imbalanced.
+We also demonstrate the utility of the gIVH in diagnosing problematic extrapolations when estimating the relative abundance of ribbon seals in the Bering Sea as a function of predictive covariates.
+We suggest that ecologists routinely use diagnostics such as the gIVH and related metrics (e.g., prediction variance) to help gauge the reliability of predictions from statistical models.
 
-{\em Key words: Abundance, Extrapolation, General additive models, Generalized linear models, Independent Variable Hull, Leverage, Prediction variance, Spatial regression, Species distribution model}
+{\em Key words: Abundance, Extrapolation, Independent Variable Hull, Prediction variance, Spatial regression, Species distribution model}
 
@@
 In ecology and conservation, a common goal is to make predictions about an unsampled random variable given a limited sample from the target population.
-For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_s$, we often desire to predict a new observation $y_s$ at $s$.
-For instance, we might use a generalized linear model \citep{McCullaghNelder1989} or one of its extensions to predict species density or occurrence in a new location, or to predict the future trend of a population.
+For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_i$, we often desire to predict a new observation $y_i$ at $i$ (where $i$ can be a design point or a spatial location).
+For instance, we might use a generalized linear model \citep[GLM;][]{McCullaghNelder1989} or one of its extensions to predict species density or occurrence in a new location.
+Such predictions can be of direct use to conservation and management, for instance, in estimating population abundance or distribution, and for projecting shifts in species range as a function of climate change.
+Spatially explicit estimates of abundance are also useful for testing theory related to biogeography or biodiversity \citep[e.g., neutral theory;][]{Hubbell2001}, and for accurately estimating the strength of density dependence \citep{ThorsonEtAlInPress}.
 
@@
 past the range of observed data.
-This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new point $y_s$ whenever $x_s < \min({\bf x})$ or $x_s > \max({\bf x})$.
+This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new response $y_i$ whenever $x_i < \min({\bf x})$ or $x_i > \max({\bf x})$ (where $x_i$ is an independent variable measured at point $i$).
 But what about more complicated situations
@@
 
-Our aim in this paper is to investigate one possibility for defining extrapolation in the generalized linear model and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatial, temporal, or spatio-temporal regression models (STRMs).
+Our aim in this paper is to investigate one possibility for defining extrapolation in the GLM and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatio-temporal regression models (STRMs).
 In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers \citep{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points.
-Application of the gIVH and related criterion (e.g.
-prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction and survey design.
-We illustrate use of the gIVH on simulated count data, on a species distribution model (SDM) for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).
+Application of the gIVH and related criterion (e.g., prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction and survey design.
+We illustrate use of the gIVH on simulated count data, and on a species distribution model (SDM) for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea.
 
-\section{The generalized independent variable hull (gIVH)}
+\centerline{\sc GENERALIZING THE INDEPENDENT VARIABLE HULL}
 
@@
 In the context of outlier detection, Cook (\citeyear{Cook1979}) defined an independent variable hull (IVH) as the smallest convex set containing all design points of a full-rank linear regression model.
-Linear regression models are often written in matrix form, i.e.
+Linear regression models are often written in matrix form, i.e.,
 \begin{linenomath*}
@@
 \end{linenomath*}
-where ${\bf Y}$ give observed data, ${\bf X}$ is a so-called design matrix that includes explanatory variables \citep[see e.g.][]{Draper1966}, and $\boldsymbol{\epsilon}$ represent normally distributed residuals (here and throughout the paper, bold symbols will be used to denote vectors and matrices).
+where ${\bf Y}$ are observed responses, ${\bf X}$ is a so-called design matrix that includes explanatory variables \citep[see e.g.,][]{Draper1966}, and $\boldsymbol{\epsilon}$ represent normally distributed residuals (here and throughout the paper, bold symbols will be used to denote vectors and matrices).
 Under this formulation, the IVH is defined relative to the hat matrix, ${\bf V}_{\rm LR}={\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$ (where the subscript ``LR" denotes linear regression).
@@
 \ref{fig:IVH}).
-In simple linear regression with one predictor variable, predictions on a hypothetical response variable obtained at covariate values below the lowest observed value or above the highest observed value are primarily outside of the IVH.
-We suspect this result conforms to most ecologists intuition about what constitutes ``extrapolating past the observed data."  However, fitting a quadratic model exhibits more nuance; if there is a large gap between design points, it is entirely possible that intermediate covariate values will also be outside of the IVH and thus more likely to result in problematic predictions.
+In simple linear regression with one predictor variable, predictions on a hypothetical response variable obtained at covariate values slightly outside the range of observed data are also outside the IVH.
+However, fitting a quadratic model exhibits more nuance; if there is a large gap between design points, intermediate covariate values may also be outside of the IVH and thus more likely to result in problematic predictions.
 Fitting a model with two covariates and both linear and quadratic effects, the shape of the IVH is somewhat more irregular, and even includes a hole in the middle of the surface when interactions are modeled (Fig.
@@
 
-Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes iid Gaussian error.
+Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes independent and identically distributed ($iid$) Gaussian error.
 Thus, it is not directly applicable to generalized models, such as those including alternative error distributions (e.g., Poisson, binomial) or spatial random effects.
@@
   \label{eq:gIVH}
-  {\rm var}(\lambda_s) \le {\rm max} [{\rm var}(\boldsymbol{\lambda}_{\mathcal{S}})],
+  {\rm var}(\lambda_i) \le {\rm max} [{\rm var}(\boldsymbol{\lambda}_{\mathcal{S}})],
 \end{equation}
 \end{linenomath*}
-where $s \in \mathcal{S}_0$, $\lambda_s$ corresponds to the mean prediction at $s$, $\mathcal{S}$ denotes the set of observed design points, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.
+where $i \in \mathcal{S}_0$, $\lambda_i$ corresponds to the mean prediction at $i$, $\mathcal{S}$ denotes the set of observed design points, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.
 
@@
 \begin{equation}
-  Y_s \sim f_Y(g(\mu_s)),
+  Y_i \sim f_Y(g^{-1}(\mu_i)),
 \end{equation}
 \end{linenomath*}
-where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives an inverse link function, and
-$\mu_s$ is a linear predictor.  For many such generalizations, it is possible to specify the $\mu_s$ as
+where $f_Y$ denotes a probability density or mass function (e.g., Bernoulli, Poisson), $g$ gives a link function, and
+$\mu_i$ is a linear predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
 \begin{linenomath*}
@@
 where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; Appendix A).
-This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method \citep{Dorfman1938,VerHoef2012} to approximate prediction variance on the real scale (i.e.
-the scale in which data are measured) as
-\begin{linenomath*}
-\begin{eqnarray}
-  \label{eq:var-lambda}
-  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
-  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
-%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
-%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
-%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
-%      \vdots & \vdots & \ddots & \vdots \\
-%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
-%  \end{array} \right].
-\end{eqnarray}
-\end{linenomath*}
-where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).
-Under common univariate link functions (e.g.
-log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g.
-multinomial logit) $\boldsymbol{\Delta}$ will be dense.
-
+This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method \citep{Dorfman1938,VerHoef2012b} to approximate prediction variance on the real scale (i.e., the scale in which data are measured; see Appendix A).
 Unlike Cook's original formulation, the gIVH will often involve unknown parameters.
-Further, the exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the underlying model structure and estimation procedure.
 For most of the following treatment, we shall assume that data have already been collected (see Discussion for comments on the potential use of the gIVH in survey planning), and that informative probability distributions can be specified for these parameters.
-Specifically, we impose a probability distribution $[\boldsymbol{\theta}]$ for unknown parameters, and compute prediction variance as
-\begin{linenomath*}
-\begin{equation}
-  \label{eq:E_var_lambda}
-  {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}) = \int_{\boldsymbol{\theta}} {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}, \boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
-\end{equation}
-\end{linenomath*}
-Our approach in this paper is to conduct a Bayesian analysis and substitute the posterior distribution $[\boldsymbol{\theta}|{\bf Y}]$ for $[\boldsymbol{\theta}]$ in Eq.
-\ref{eq:E_var_lambda}.
-However, one could also consider substituting parametric approximations based on output from standard likelihood-based analyses.
 
@@
 In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).
-The gIVH, together with posterior prediction variance, seem ideally situated to diagnosing potential extrapolation issues.
+The gIVH, together with posterior prediction variance, seem ideally suited to diagnosing potential extrapolation issues.
 Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions will be based on.
@@
 We developed a package \texttt{SpatPred} in the R statistical programming environment \citep{RTeam2012} to simulate data and conduct all analyses.
-The seal and sea lion datasets are included as part of this package, and are available at \url{https://github.com/pconn/SpatPred}.
+The seal dataset is included as part of this package, and is available at \url{https://github.com/pconn/SpatPred}.
 The R package has also been archived via figshare [WE PLAN TO DO THIS WHEN THE MANUSCRIPT IS ACCEPTED FOR PUBLICATION].
 
-\section{Examples}
-
-We used simulation to investigate the usefulness of the gIVH for limiting the scope of inference when fitting models to count data that have already been collected.
-We also applied the gIVH to examining the robustness of a species distribution model (SDM) of ribbon seals in the eastern Bering Sea, and to time series projections of Steller sea lion population trends.
+\centerline{\sc Examples}
 
@@
 Animal abundance was generated as a function of three hypothetical spatially autocorrelated habitat covariates (Appendix B).
-For each simulated landscape, we conducted virtual surveys of $n=45$ survey units using two different designs: (1) a spatially balanced sample using a random tesselation design \citep{StevensOlsen2004}, and (2) a convenience sample where the probability of sampling was greater for cells closer to a ``base of operations" located in the middle of the survey grid.
-The former approach preserves randomness while seeking a degree or regularity when distributing sampling locations across the landscape, while the latter may be easier to implement.
+For each simulated landscape, we conducted virtual surveys of $n=45$ survey units using two different designs: (1) a spatially balanced sample \citep{StevensOlsen2004}, and (2) a convenience sample where the probability of sampling was greater for cells closer to a ``base of operations" located in the middle of the survey grid.
+The former approach preserves randomness while seeking a degree of regularity when distributing sampling locations across the landscape, while the latter may be easier to implement logistically.
 
@@
 Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (Appendix A).
-The GLM and STRM expressed log-density as a function of linear and quadratic covariate effects (as well as first order interactions), while the GAM used a kernel smoother with 4 knots for each covariate (Appendix A).
+The fixed effects components of the GLM and STRM were configured to have both linear and quadratic covariate effects and first-order interactions, while the GAM expressed log-density as a function of smooth terms for each covariate (Appendix B).
 Each model was provided with two of the three covariates used to generate the data.
@@
 Specifically, the performance of $\text{gIVH}$ may help decide its utility in limiting the scope of inference once data have been collected and analyzed, and perhaps point out areas worthy of additional sampling.
-A fuller, technical description of the simulation study design is provided in Appendix B; for illustration of a single simulation replicate see Fig.
-\ref{fig:sim_maps}.
+A fuller, technical description of the simulation study design and several illustrations are provided in Appendix B.
 
 Posterior predictions from simulations indicated that the distribution for proportional error in total abundance was right skewed when statistical inference was made with regard to the entire survey area (Fig.
-\ref{fig:sim_boxplots}).
+B3).
 This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.
 The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.
-Interestingly, the convolution kernel GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.
-For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude (e.g.
-$\approx 0.3$; Fig.
-\ref{fig:sim_boxplots}) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
+For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude ($\approx 0.3$) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
+By contrast, proportional bias was close to zero when inference was restricted to the gIVH.
+Interestingly, the radial basis GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.
 
 \subsection{Ribbon seal SDM}
-As part of an international effort, researchers with the U.S.
-National Marine Fisheries Service conducted aerial surveys over the eastern Bering Sea in 2012 and 2013.
-Agency scientists used infrared video to detect seals that were on ice, and collected simultaneous digital photographs to provide information on species identity.
-Here, we use spatially referenced count data from photographed ribbon seals, {\it Phoca fasciata} (Fig.
-\ref{fig:ribbon}) on a subset of 10 flights flown over the Bering Sea from April 20-27, 2012.
-We limited flights to a one week period because sea ice melts rapidly in the Bering Sea in the spring, and modeling counts over a longer duration would likely require addressing how sea ice and seal abundance changes over both time and space \citep[see][]{ConnEtAl2015}.
-However, limiting analysis to a one week period makes the assumption of static sea ice and seal densities tenable \citep{ConnEtAl2014}.
 
-Our objective with this dataset will be to model seal counts on transects through 25km by 25km grid cells as a function of habitat covariates and possible spatial autocorrelation.
-Estimates of apparent abundance can then be obtained by summing predictions across grid cells.
-Figure \ref{fig:flights} shows the transects flown and the number of ribbon seals encountered in each cell, and Figure \ref{fig:covs} show explanatory covariates gathered to help predict ribbon seal abundance.
-These data are described in fuller detail by \citet{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors.
-Since our focus in this paper is on illustrating spatial modeling concepts, we devote our efforts to the comparably easier problem of estimating apparent abundance (i.e., uncorrected for vagaries of the detection process).
+To demonstrate application of the gIVH on a real dataset, we consider the case of estimating ribbon seal relative abundance from count data obtained during aerial surveys of the Bering Sea in April, 2012 (Fig.
+\ref{fig:flights}).
+In this case, we wish to model seal counts as a function of habitat covariates and possible spatial autocorrelation, and to use predictions from this fitted model to estimate relative abundance for the entire study area.
+For further background and details on these data, we refer the reader to Appendix C.
 
@@
 Predicting abundance in areas further south and east may thus prove problematic, as the values of several explanatory covariates (Fig.
-\ref{fig:covs}) are also maximized in these regions.
+C1) are also maximized in these regions.
 To illustrate, let $Y_i$ denote the ribbon seal count ($Y_i$) obtained in sampled grid cell $i$.
@@
 \end{eqnarray}
-where $P_i$ gives the proportion of area photographed in grid cell $i$, $A_i$ gives the proportion of suitable habitat in cell $i$ (ribbon seals do not often utilize land), $\theta_i$ is a linear predictor, and $\epsilon_i$ is normally distributed $iid$ error.
-By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs, GAMs, and STRMs (see Appendix S1).
+where $P_i$ gives the proportion of area surveyed in grid cell $i$, $A_i$ gives the proportion of cell $i$ that is seal habitat, $\theta_i$ is a linear predictor, and $\epsilon_i$ is normally distributed $iid$ error.
+By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs, GAMs, and STRMs (see Appendix A).
 
 We start by fitting a hierarchical GLM, GAM, and STRM to the full suite of predictor covariates (Fig.
-\ref{fig:covs}).
-For the GLM and STRM, we fit a model with linear effects of all predictor variables, and with an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}.
-For the STRM, we imposed a restricted spatial regression (RSR) formulation for spatially autocorrelated random effects, where dimension reduction was accomplished by only selecting eigenvectors of the spectral decomposition associated with eigenvalues that were greater than 0.5 (see Appendix S1 for additional information on model structure).
-For the GAM, we employed smooth terms with 8 knots for each covariate.
-This included 4 knots at the minimum, 1/3 and 2/3 quantiles, and the maximum observed covariate values, in addition to 4 knots above and below the range of observed data to reduce the magnitude of biased endpoint behavior.
+C1).
+For the GLM and STRM, we fit a model with linear effects of all predictor variables and an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}.
+For the GAM, smooth effects were included for each covariate, with 8 knots each.
+As with simulated data examples, we adopted a Bayesian perspective and conducted estimation and posterior prediction using MCMC (see Appendix A for estimation algorithm details and Appendix C for further numerical details related to the ribbon seal example).
 
-As with simulated data examples, we adopted a Bayesian perspective and conducted estimation using MCMC (see Appendix S1 for algorithm details) with 60,000 iterations where the first 10,000 iterations were discarded as a burn-in.
-We generated posterior predictions of ribbon seal abundance across the landscape as
-\begin{equation}
-  N_i \sim {\rm Poisson}(A_i \lambda_i),
-\end{equation}
-and calculated the gIVH as in Eqn. \ref{eq:gIVH}, with delta method modifications as specified in Eqns. \ref{eq:var-mu}-\ref{eq:E_var_lambda}.
-Initial spatial predictions using the three models (GLM, GAM, STRM) all produced extremely high, unbelievable predictions along the southern boundary of the study area \ref{fig:ribbon_plot}.
+Initial spatial predictions using the three models (GLM, GAM, STRM) all produced extremely high, unbelievable predictions along the southern boundary of the study area (Fig.
+\ref{fig:ribbon_plot}).
 Predictions in this region were also largely out of the gIVH, indicating the potential utility for the gIVH in revealing problematic extrapolations.
-
 We considered several possible alternatives for trying to obtain more robust abundance estimates before settling on a preferred alternative.
@@
 However, this is not ideal in that one does not get an abundance estimate for the whole study area, and it may be difficult to compare abundance from one year to the next using this approach.
-Second, one could consider a simpler estimation model, such as one post-stratifying by ecoregion as was used in a previous analysis of ribbon seal data to try to avoid this problem \citep{ConnEtAl2014}.
+Second, one could consider a simpler estimation model, such as one post-stratifying by ecoregion \citep[as in][]{ConnEtAl2014}.
 Finally, one could build in a priori knowledge of habitat preferences into the model structure.
@@
 This solution seemed the most logical, as many of the large, anomalous predictions were over open water along the southern edge of the study area, where we would have obtained zero counts had they been surveyed.
+This approach effectively requires that sea ice concentration be included as a predictive covariate to help model absences in cells without ice.
 
@@
 We note that these estimates are for example illustration only, as they are uncorrected for imperfect detection \citep[e.g., incomplete detection of thermal cameras, animals that were unavailable for sampling because they were in the water, species misidentification; see][]{ConnEtAl2014}.
+Our approach here was to examine extrapolation and prediction error using relatively simple models, with the understanding that such effects are also likely to occur in complex models with more realistic observation processes.
 
-\subsection{Steller sea lion trends}
-
-\section{Discussion}
+\centerline{\sc Discussion}
 
 We have demonstrated the capacity for certain classes of statistical models to produce biased predictions of animal abundance when extrapolating past the range of observed data.
-In simulations, commonly used models (GLMs, STRMs) exhibited positive bias when predictions were required for the entire study area, particularly when convenience sampling was employed; in the ribbon seal example, naive extrapolation of fitted statistical relationships produced high positive bias along the southern boundary of the study area; and in the Steller sea lion example, projected abundance trends were increasingly implausible the farther the predictions were from the realm of observed data.
+In simulations, commonly used models (GLMs, STRMs) exhibited positive bias when predictions were required for the entire study area, particularly when convenience sampling was employed (with mean proportional bias $\approx 30\%$).
+In the ribbon seal example, naive extrapolation of fitted statistical relationships produced high positive bias along the southern boundary of the study area.
 However, the gIVH appeared useful in diagnosing places where extrapolations from the fitted statistical model were problematic.
@@
 
-Use of the gIVH in seal and sea lion applications to identify extrapolation issues was not particularly profound, as experts could easily use ecological knowledge of these populations to identify biologically implausible predictions and come to the conclusion that models either needed to be reformulated or inference limited to a smaller set of design points.
-Likewise, one could argue that the high positive bias in the simulation study could likely have been reduced had an expert meticulously examined diagnostics from each simulation replicate and reconfigured estimation models to better match the range of covariates observed.
-However, such determinations are likely to be quite subjective, and may prove insufficient when there are large number of regression coefficients and interaction terms.
-Even relatively simple regression models may exhibit non-intuitive patterns (e.g.
-Fig.
-\ref{fig:IVH}).
-Further, relying on expert opinion alone in successive rounds of model formulation and fitting may lead to investigators choosing models based on how much they like the results, which is clearly not ideal scientific practice.
-
 Our intent in writing this paper is to raise awareness of potential problems with extrapolation bias in statistical models, and to provide an additional tool (the gIVH) to help diagnose its presence.
-Other approaches for selecting models to reduce overfitting, such as cross validation \citep[e.g.][]{PicardCook1984}, are also useful for this purpose, but may not entirely eliminate the problem (particularly for sparse datasets).
+Other approaches for selecting models to reduce overfitting, such as cross validation \citep[e.g.,][]{PicardCook1984}, are also useful for this purpose, but may not entirely eliminate the problem (particularly for sparse datasets).
 In fields such as species distribution modeling and model-based abundance estimation, the goal for analysts is often to build predictive maps of species abundance and/or occurrence using a limited number of sample locations.
@@
 When inference is extended to unsampled locations, the gIVH appears to be a useful diagnostic for whether bias for predictions in unsampled locations can be expected.
+In some cases, biological knowledge and intuition may be sufficient to diagnose anomalous predictions.
+However, such determinations are likely to be quite subjective, and may prove insufficient when there are a large number of regression coefficients and interaction terms.
+For instance, even relatively simple regression models may exhibit non-intuitive patterns (e.g., Fig.
+\ref{fig:IVH}).
+Further, relying on expert opinion alone in successive rounds of model formulation and fitting may lead to investigators choosing models based on how much they like the results, which is clearly not ideal scientific practice.
 
@@
 However, one could also apply the gIVH when predicting species occurrence from presence/absence data.
-In this case, common link functions (e.g.
-probit or logit) are symmetric, and potential for positive and negative bias in predictive maps seem equally likely.
+In this case, common link functions (e.g., probit or logit) are symmetric, and potential for positive and negative bias in predictive maps seem equally likely.
 Future research should be directed to examine conditions under which the gIVH is a useful diagnostic in such applications.
@@
 One area that gIVH ideas may also prove useful is in formulating survey designs.
-The topic of optimal (or near-optimal) spatial design has received considerable attention in the statistical literature, often in the context of designing environmental monitoring systems \citep{DiggleLophaven2006,XiaEtAl2006,Muller2007}.
-Optimal designs can differ depend on whether one wishes to minimize a criterion such as integrated prediction variance, or a summary of the information matrix (i.e.
-if one wishes to optimize regression parameter performance).
-Optimal designs can also be sensitive to the structure of the estimation model that is used, so that tailoring a survey design to a particular model can be somewhat dangerous if there is uncertainty about the ultimate ``best" structure for the model used to relate animal abundance and occurrence to landscape or habitat covariates.
-Nevertheless, one could still think about augmenting a spatially balanced sampling design with a number of locations which are known or thought to have high leverage as a function of available covariates.
+The topic of optimal (or near-optimal) spatial design has received considerable attention in the statistical literature, often in the context of designing environmental monitoring systems \citep{Muller2007}.
+Optimal designs can be sensitive to the structure of the estimation model that is used, so that tailoring a survey design to a particular model can be somewhat dangerous if there is uncertainty about the ultimate ``best" structure for the model used to relate animal abundance and occurrence to available covariates.
+Nevertheless, one could still think about augmenting a spatially balanced sampling design with a number of locations which are known or thought to have high prediction variance as a function of available covariates.
 We are excited about this prospect, and it is a subject of current research.
-
-With regards to ribbon seal abundance, we once again caution that the estimates presented here are for illustration only, as numerous factors that influence the detectability of seals was not included in count models used in this paper \citep[but see][]{ConnEtAl2014}.
-Our approach here was to examine extrapolation and prediction error using relatively simple models, with the understanding that such effects are also likely to occur in complex models with more realistic observation components.
-Indeed, our experience is that the magnitude of extrapolation bias can increase when additional data on imperfect detection or species misclassification are included.
 
@@
 
+\centerline{\sc Supplemental Material}
+
+\centerline{\bf Appendix A}
+Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model ({\it Ecological Archives} XXXX-XXX-A1).
+
+\centerline{\bf Appendix B}
+Full details of simulation study examining predictive extrapolation ({\it Ecological Archives} XXXX-XXX-A2).
+
+\centerline{\bf Appendix C}
+Additional details on ribbon seal count data ({\it Ecological Archives} XXXX-XXX-A3).
+
 \clearpage
+%FIGURE 1. A ribbon seal, {\it Phoca fasciata}; the focus of spatial %modeling efforts in this paper.
 
-FIGURE 1. A ribbon seal, {\it Phoca fasciata}; the focus of spatial modeling efforts in this paper.
+FIGURE 1.
+Example IVHs constructed from simulated data.
+In (A) and (B), linear regression is used to relate a response variable to a single covariate, $x$, obtained at locations denoted with an ``x".
+Using $x$ as a simple linear effect (A), only predictions less than the minimum observed value of $x$ or greater than the maximum value of $x$ are outside the IVH (shaded area), as scaled prediction variance in these areas (solid line) is greater than the maximum scaled prediction variance for observed data (dashed line).
+Using both linear and quadratic effects (B), some intermediate points are also outside the IVH.
+When both linear and quadratic effects of two covariates ($x_1$ and $x_2$) are modeled, the IVH is more nuanced and depends on whether interactions are omitted (C) or included (D).
 
 FIGURE 2.
-Example IVHs constructed from simulated data.
-In panels A and B, the investigator plans to model a hypothetical (unmeasured) response variable using a linear regression model as a function of a single covariate, $x$, obtained at a number of design points (denoted with an ``x").
-Using $x$ as a simple linear effect (A), only predictions less than the minimum observed value of $x$ or greater than the maximum value of $x$ are outside the IVH (shaded area), as scaled prediction variance in these areas (solid line) are greater than the maximum scaled prediction variance for observed data (dashed line).
-Using both linear and quadratic effects of $x$, some intermediate points are also outside the IVH; predictions at these points should also be viewed with some caution.
-Panels C $\&$ D show a more complicated IVH when the investigator wishes to relate an unmeasured response variable to linear and quadratic effects of two covariates, $x$ and $y$, either without interactions (C) or with interactions (D).
-Any potential predictions in the shaded area are outside of the IVH.
-
-FIGURE 3.
-Depiction of a single simulation scenario.
-Panels (A-C) give simulated covariate values, panel D gives true animal abundance, (E) gives estimated abundance from a GLM run on count data from a spatially balanced survey design, and (F) gives abundance from a GLM applied to count data from a convenience survey.
-In (E-F), predictions outside the gIVH are represented by black boxes, and sampling locations are represented with an x.
-For the convenience sample, the median posterior abundance prediction for the entire survey area is $57\%$ greater than true abundance when inference is made to the whole study area.
-When inference is limited to the gIVH, median posterior abundance was just $16\%$ greater than true abundance.
-
-FIGURE 4.
-Boxplots summarizing distribution of proportional error in the posterior predictive median of abundance for the simulation study as a function of estimation model (x-axis), survey design (columns) and whether or not inference was restricted to the gIVH (rows).
-The lower and upper limits of each box correspond to first and third quartiles, while whiskers extend to the lowest and highest observed bias within 1.5 interquartile range units from the box.
-Outliers outside of this range are denoted with points.
-Horizontal lines within boxes denote median bias.
-The two numbers located below each boxplot indicate mean bias (upper number) and the number of additional outliers for which proportional bias was greater than 2.0 (lower number).
-
-FIGURE 5.
 Aerial survey tracks over the Bering Sea, April 22-29, 2012 (blue lines) overlayed on a tesselated study area consisting of 25km by 25km grid cells (gray lines).
-Dark gray indicates land, while the brown line indicates a 1000m depth contour, and the orange line shows the U.S Exclusive Economic Zone (EEZ) boundary (which was used to construct the easterly edge of the survey area).
+Dark gray indicates land, while the brown line indicates a 1000m depth contour, and the orange line shows the U.S Exclusive Economic Zone (EEZ) boundary.
 Colored pixels indicate ribbon seal counts along aerial transects.
@@
 
-FIGURE 6.
-Potential covariates gathered to help explain and predict ribbon seal abundance in the eastern Bering Sea.
-Covariates include distance from mainland (\texttt{dist\_mainland}), distance from 1000m depth contour (\texttt{dist\_shelf}), average remotely sensed sea ice concentration while surveys were being conducted (\texttt{ice\_conc}), and distance from the southern sea ice edge (\texttt{dist\_edge}).
-All covariates except ice concentration were standardized to have a mean of 1.0 prior to plotting and analysis.
-
-FIGURE 7  Posterior mean predictions of ribbon seal apparent abundance across the eastern Bering sea from models fit to survey data.
+FIGURE 3.
+Posterior mean predictions of ribbon seal apparent abundance across the eastern Bering sea from models fit to survey data.
 Each row gives result for different model types (GLM, GAM, or STRM, respectively); left column plots give results for naive runs without pseudo-absences, while plots in the right column give predictions for runs where psuedo-absence data (i.e., 0 counts in cells with less than $0.1\%$ ice) were included.
@@
 
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=6in]{ribbon}
-\end{center}
-\caption{ }
-\label{fig:ribbon}
-\end{figure}
+%\begin{figure}[!h]
+%\begin{center}
+%\includegraphics[width=6in]{ribbon}
+%\end{center}
+%\caption{ }
+%\label{fig:ribbon}
+%\end{figure}
 
@@
 \begin{center}
-\includegraphics[width=6in]{sim_maps.pdf}
-\end{center}
-\caption{ }
-\label{fig:sim_maps}
-\end{figure}
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=6in]{bias_generic.pdf}
-\end{center}
-\caption{ }
-\label{fig:sim_boxplots}
-\end{figure}
-
-\begin{figure}[!h]
-\begin{center}
 \includegraphics[width=6in]{BOSS_survey_ribbon.pdf}
@@
 \label{fig:flights}
-\end{figure}
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=6in]{covariates.pdf}
-\end{center}
-\caption{ }
-\label{fig:covs}
 \end{figure}

### 2015-03-24 SpatPred/R/gIVH.R

@@ -1,84 +1,44 @@
-### calculate prior prediction variance/gIVH via Monte Carlo
-
-gIVH <- function(Data,Effort,spat.mod=1,K=NULL,Assoc=NULL,my.formula,Tau.beta,Tau.eta,Tau.epsilon,srr.tol=0.8){
-  n.reps=1000
-  X.pred=model.matrix(my.formula,data=Data@data)
-  X.obs=model.matrix(my.formula,data=Data@data[Effort$Mapping,]) 
-  n.obs=length(Effort$Mapping)
-  XpXinv=solve(crossprod(X.obs))
-  n.beta=ncol(X.pred)
-  I.beta=diag(nrow(XpXinv))
+#' a generic function to calculate the generalized independent variable hull for spatial models when
+#' model is of form Data ~ f(g^{-1}(mu)), where f is a probability density/mass function, g is a link function, and
+#' mu = X*Beta + K*Theta + Epsilon.  Here, X is a design matrix and Beta give fixed effects,
+#' K is an expansion matrix for spatial (or GAM) random effects, Theta are random effects (if modeled), and Epsilon is normally distributed error
+#' @param X A design matrix for fixed effects (including sampled and unsampled locations)
+#' @param Beta A matrix holding a sample from the joint (prior or posterior) distribution of regression parameters. The number of rows = the number of samples, while the number of columns gives the number of regression coefficients.
+#' @param Sampled An integer-valued vector indicating which cells were actually sampled (these should correspond to rows of X)
+#' @param K An expansion matrix for random effects (if not provided, a diagonal matrix is used)
+#' @param Theta If random effects, a matrix holding a sample from the joint distribution of spatial random random effects (# rows = sample size, # columns = # of random effects).  Default is NULL.
[…]

### 2015-03-24 SpatPred/R/mcmc_abundance.R

@@ -1,204 +1,0 @@
-#' function to perform Bayesian analysis of count data using a spatial log Gaussian Cox process model
-#' @param model A formula object specifying the linear predictor for abundance intensity
-#' @param Data   A list holding the following objects:
-#'        Adj - Adjacency matrix of ones and zeros describing spatial connectivity (for ICAR modeling)
-#'        Grid - A data frame holding covariate values for all cells one wants to make predictions on (including sampled cells)
-#'        Count.data - A data frame with the following columns: (1) "Cell" - entries this column provides the numerical index (in 1:S) for the cell surveyed
-#'              (3) "AreaSurveyed" - provides
-#'              the proportion of the target cell that was surveyed, (4) "Count" Number of target species counted
-#' @param Area.adjust   A vector allowing for differences in suitable habitat for each cell.  Can be used for different grid cell sizes or different
-#'        proportions of suitable habitat (e.g., 1.0 = 100% of habitat is suitable for the focal species)
-#' @param Control A list giving MCMC controls and additional model options
-#'  "iter": number of MCMC iterations;
-#'  "burnin": number of MCMC burnin iterations;
-#'	"thin": if specified, how many iterations to skip between recorded posterior samples;
-#'  "srr.tol": Threshold eigenvalue level for SRR; only eigenvectors with higher eigenvalues than srr.tol are included in SRR formulation (default is 0.5)
-#'  "predict" If TRUE (default), calculate posterior predictions across the entire grid
-#'  "MH.mu" A vector providing continuous Uniform half-range values for Metropolis-Hastings proposals
[…]

### 2015-03-24 SpatPred/R/sim_data_abundance.R

@@ -1,4 +1,4 @@
-#' function to simulate spatial count data over a simulated landscape
+#' function to simulate abundance over a simulated landscape
 #' @param S Number of cells (must be a square number)
 #' @param n.covs Number of habitat covariates (default 4)
 #' @param tau.epsilon Precision of exchangeable errors in log space
@@ -7,19 +7,13 @@
 #' @keywords spatio-temporal, simulation
 #' @author Paul B. Conn
 sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
-  require(ggplot2)
-  require(Matrix)
-  require(spatstat)
-  require(sp)
-  require(RandomFields)
-  
   DEBUG=FALSE   
   PLOT=FALSE
   if(PLOT)set.seed(11111)
   tau.epsilon=100
   #if(DEBUG)tau.epsilon=100
   
-  source('./SpatPred/R/util_funcs.R')
+  #source('./SpatPred/R/util_funcs.R')
    
   if(sqrt(S)%%1>0)cat("error in sim_data_generic; S must be a square number \n")
   
@@ -54,22 +48,13 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
     Grid2=data.frame(y=rep(1:30,each=30),x=rep(c(1:30),30),cov=as.vector(Covs[,2]))
     Grid3=data.frame(y=rep(1:30,each=30),x=rep(c(1:30),30),cov=as.vector(Covs[,3]))
     Grid=rbind(Grid1,Grid2,Grid3)
-    #Covariate=c(rep(1,S),rep(2,S),rep(3,S))
-    #Grid=cbind(Grid,Covariate)
-    library(ggplot2)
-    library(RColorBrewer)
     myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
-    #Grid.plot=ggplot()+geom_raster(data=Grid,aes(x=x,y=y,fill=cov))+
-    #  theme(axis.text=element_blank(),axis.title=element_blank(),plot.title = element_text(hjust = 0))+
-    #  ggtitle("A.")+scale_fill_gradientn(colours=myPalette(100))+facet_wrap(~Covariate)   
[…]

### 2015-03-24 SpatPred/R/spat_pred.R

@@ -3,7 +3,7 @@
 #' @param Data   A SpatialPolygonsDataFrame holding covariate values for all cells one wants to make predictions on (including sampled cells)
 #' @param Effort A list including "Counts" which gives the response variable for surveyed cells, and "Mapping" which is a vector indicating which sample unit is associated with which count
 #' @param Area.adjust   A vector allowing for differences in suitable habitat for each cell.  Can be used for different grid cell sizes or different
-#'        proportions of suitable habitat (e.g., 1.0 = 100% of habitat is suitable for the focal species)
+#'        proportions of suitable habitat (e.g., 1.0 = 100 percent of habitat is suitable for the focal species)
 #' @param Offset	A vector giving the fraction of each grid cell in Mapping that was surveyed
 #' @param spat.mod Controls whether (and what type) of spatial autocorrelation is implemented (spatmod = 0: no spatial autocorrelation; spatmod=1: process convolution; spatmod=2: srr ICAR) 
 #'        Note that if spatmod=1, the argument Knots needs to be specified; if spatmod=2, Adj must be specified
@@ -31,15 +31,14 @@
 #' 	MCMC: A list object containing posterior samples;
 #'  Accept: A list object indicating the number of proposals that were accepted for parameters updated via Metropolis-Hastings;
 #'  Control: A list object giving MCMC tuning parameters (which are updated if the 'adapt' alorithm is used)
+#'  gIVH: A binary vector, with length equal to the number of survey units, describing whether predictions for each survey unit are (=1) or are not (=0) within the gIVH
+#'  K.gam: A ``design matrix" associated with smooth effects, should the GAM procedure be employed
 #' @export
 #' @import Matrix
 #' @keywords areal model, data augmentation, mcmc, spatial prediction
 #' @author Paul B. Conn \email{paul.conn@@noaa.gov} 
 #' @examples print("Later!")
 spat_pred<-function(formula,Data,Effort,spat.mod=0,Offset,Area.adjust,Control,Assoc=NULL,K=NULL,Names.gam=NULL,Knots.gam=NULL,Prior.pars=NULL,Precision.pars=NULL){
-  require(Matrix)
-  require(mvtnorm)
-  require(rgeos)
   #do a little checking
[…]

### 2015-03-24 SpatPred/R/util_funcs.R

@@ -119,13 +119,12 @@ d_biv_normal<-function(Tmp.vec,XY,Sigma){
 #' @param cell.width if highlight is provided, cell.width must provide the width of a composite grid cell
 #' @param leg.title Title for legend of plot (if different from the default "Abundance")
 #' @param hcolor If highight provided, gives color for highlighting (default = "yellow")
+#' @param cur.max If provided, sets maximum value for plotted abundance
 #' @return An abundance map
 #' @export 
 #' @keywords abundance map, plot
 #' @author Paul Conn \email{paul.conn@@noaa.gov}
 plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance",hcolor='yellow',cur.max=NULL){
-  library(sp)
-  library(RColorBrewer)
   myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))  
   Tmp=Grid[[1]]
   if(is.null(highlight)==FALSE){
@@ -162,79 +161,6 @@ rrw <- function(Q){
   if(sum(val.inv==0)==2) X <- cbind(X, 1:length(sim))
   sim <- sim-X%*%solve(crossprod(X), crossprod(X,sim))
   return(sim)
-}
-
-#' stack list of vectors into one big vector
-#' @param L list vector of matrices
-#' @return X one vector consisting of stacked elements of L
-#' @export 
-#' @author Paul Conn \email{paul.conn@@noaa.gov}
-stack_list_vector<-function(L){
-  X=L[[1]]
-  len.L=length(L)
-  if(len.L>1){
-    for(ilen in 2:len.L)X=c(X,L[[ilen]])  #could be better optimized
-  }
-  X
-}
-
-#' stack list into matrix
-#' @param L list vector of matrices
-#' @return X one matrix consisting of stacked elements of L
-#' @export 
-#' @author Paul Conn \email{paul.conn@@noaa.gov}
-stack_list<-function(L){
-  X=L[[1]]
-  len.L=length(L)
-  if(len.L>1){
-    for(ilen in 2:len.L)X=rbind(X,L[[ilen]])  #could be better optimized
-  }
-  X
-}
-
-#' neg log likelihood function for generating initial estimates of Beta to start off Markov chain
-#' @param Beta parameter values
-#' @param Counts Animal counts in sampled cells
-#' @param P probability of seeing an animal by cell
-#' @param X model matrix
-#' @return multinomial log likelihood
-#' @export 
[…]

### 2015-03-24 SpatPred/inst/analyze_sim_data.R

@@ -1,9 +1,7 @@
 ### ANALYZE SIMULATION DATA FOR SPATIAL PREDICTION PAPER
-
+### note: this has some hardwiring for local directory structures
 S=900
 n.sims=1000
-#Vars=matrix(0,S,3)
-#colnames(Vars)=c("GLM","GAM","RSR")
 Bias=matrix(0,n.sims,24)
 colnames(Bias)=c("GLM.sys.all","GAM.sys.all","RSR.sys.all","RSR2.sys.all",
                  "GLM.sys.gIVH","GAM.sys.gIVH","RSR.sys.gIVH","RSR2.sys.gIVH",
@@ -26,17 +24,7 @@ for(igen in 1:2){
     else cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
     load(cur.file)
         
-    #calculate posterior prediction variance on count scale
-    #Vars[,"GLM"]=apply(MCMC$GLM$MCMC$Pred,1,"var")
-    #Vars[,"GAM"]=apply(MCMC$GAM$MCMC$Pred,1,"var")
-    #Vars[,"RSR"]=apply(MCMC$RSR$MCMC$Pred,1,"var")
-    #Vars[,"RSR"]=apply(MCMC$RSR$MCMC$Pred,1,"var")
-    
-    
-    #determine gIVH membership
-    #IVH.glm=which(Vars[,"GLM"]<=max(Vars[Effort$Mapping,"GLM"]))
-    #IVH.gam=which(Vars[,"GAM"]<=max(Vars[Effort$Mapping,"GAM"]))
-    #IVH.rsr=which(Vars[,"RSR"]<=max(Vars[Effort$Mapping,"RSR"])) 
+     #determine gIVH membership
     IVH.glm=which(MCMC$GLM$gIVH==1)  #gIVH on real scale
     IVH.gam=which(MCMC$GAM$gIVH==1)
     IVH.rsr=which(MCMC$RSR$gIVH==1)   #2 covariates
@@ -135,8 +123,6 @@ bias.plot = ggplot(Bias.df,aes(factor(Model),Bias))+geom_boxplot()+facet_grid(gI
 bias.plot=bias.plot + theme(text=element_text(size=20))
 bias.plot=bias.plot + theme(axis.text.y=element_text(size=14))
 bias.plot=bias.plot + coord_cartesian(ylim=c(-1.2,2))
[…]

### 2015-03-24 SpatPred/inst/compile_plot_ribbon_data.R

@@ -53,6 +53,9 @@ Tmp<-Cur.grid
 library(ggplot2)
 library(plyr)
 library(grid)
+library(RColorBrewer)
+greenPalette <- colorRampPalette(brewer.pal(9, "Greens"))
+
 Tmp@data$id=rownames(Tmp@data)
 tmp1<-fortify(Tmp,region='id')
 tmp2<-join(tmp1,Tmp@data,by="id")
@@ -60,18 +63,18 @@ tmp2[,"Ecoregion"]=as.numeric(tmp2[,"Ecoregion"])
 new.colnames=colnames(tmp2)
 new.colnames[1:2]=c("Easting","Northing")
 colnames(tmp2)=new.colnames
-tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank())
-pdf(file="covariates.pdf")
+tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank(),plot.title = element_text(hjust = 0))
+pdf(file="covariates_col.pdf")
 pushViewport(viewport(layout=grid.layout(2,2)))
-p1=ggplot(tmp2)+aes(Easting,Northing,fill=dist_mainland)+geom_raster()+tmp.theme+scale_fill_gradient(low="black", high="white")
+p1=ggplot(tmp2)+aes(Easting,Northing,fill=dist_mainland)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=greenPalette(100),name="Value")+ggtitle("A. Distance from mainland")
 print(p1,vp=viewport(layout.pos.row=1,layout.pos.col=1))
-p2=ggplot(tmp2)+aes(Easting,Northing,fill=dist_shelf)+geom_raster()+tmp.theme+scale_fill_gradient(low="black", high="white")
+p2=ggplot(tmp2)+aes(Easting,Northing,fill=dist_shelf)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=greenPalette(100),name="Value")+ggtitle("B. Distance from shelf")
 print(p2,vp=viewport(layout.pos.row=1,layout.pos.col=2))
-p3=ggplot(tmp2)+aes(Easting,Northing,fill=ice_conc)+geom_raster()+tmp.theme+scale_fill_gradient(low="black", high="white")
[…]

### 2015-03-24 SpatPred/inst/compile_plot_spotted_data.R

@@ -1,116 +1,0 @@
-# script to compile count and effort data for ribbon seals from SLR data and make plots
-#
-# Paul B. Conn
-# 
-
-library(sp)
-library(rgdal)
-library(nPacMaps)  #from Josh London
-library(rgeos)
-library(doBy)
-library(ggplot2)
-
-#First, pull in data
-load('c:/users/paul.conn/git/SpatPred/BOSS_Power_photo_data/slr_power_data.RData') #sp data frame including species & area of photo
-load('c:/users/paul.conn/git/BOSS/BOSS/data/AlaskaBeringData.Rdat') #read in "Data" holding grid info
-load('c:/users/paul.conn/git/BOSS/BOSS/data/Effort_points.Rdat')  #read in Sp Points DF "effort_data" for on effort points (use for covariates)
-
-I.inter=gIntersects(slr_power_data,Data$Grid[[1]],byid=TRUE)
-Points=slr_power_data[which(apply(I.inter,2,'sum')>0),]  #only include effort points that are 'on grid'
-n.points=nrow(Points)
-
-#do some plotting
-plot(Data$Grid[[1]])
-points(Points,pch=19,cex=0.01,col='blue')
-
-data(alaska_dcw)
-data(russia_dcw)
-laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
-                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
-alaska_dcw <- spTransform(alaska_dcw, CRS(laea_180_proj))
-russia_dcw <- spTransform(russia_dcw, CRS(laea_180_proj))
-plot(alaska_dcw, col = "gray", add = TRUE)
-plot(russia_dcw, col = "gray", add = TRUE)
-                                                                     
-                                                                     
-
-##create a single grid object for this analysis, averaging time varying covariates over the dates considered
-Cur.grid=Data$Grid[[1]]
-n.cells=dim(Cur.grid)[1]
-S=nrow(Cur.grid)  #number of cells
-#first object corresponds to April 4 so start with 17th SpDF
-Mean.covs=matrix(0,S,3)
[…]

### 2015-03-24 SpatPred/inst/plot_IVH_simple.R

@@ -21,14 +21,14 @@ plot(x0,Y)
 PredVar.df=data.frame(x=x0,pred.var=diag(X0%*%solve(crossprod(X))%*%t(X0)))
 
 library(ggplot2)
-my.theme=theme(plot.title = element_text(hjust = 0),axis.ticks = element_blank(), axis.text.y = element_blank(),axis.title.y=element_blank(),strip.text=element_text(face="bold"),axis.title=element_text(size=16),text=element_text(size=16),legend.text=element_text(size=16),legend.key.height=unit(1.1,"line"))
+my.theme=theme(plot.title = element_text(hjust = 0),axis.ticks = element_blank(), axis.text.y = element_blank(),axis.text.y=element_blank(),strip.text=element_text(face="bold"),axis.title=element_text(size=16),text=element_text(size=16),legend.text=element_text(size=16),legend.key.height=unit(1.1,"line"))
 
 #note the next line is hardwired with values obtained by examining elements of Extrap
 Rect.df=data.frame(x.min=c(0,2.98),x.max=c(1.01,4),y.min=c(0,0),y.max=c(1,1))
 linear.plot=ggplot(Rect.df)+geom_rect(aes(xmin=x.min,xmax=x.max,ymin=y.min,ymax=y.max),fill='gray')
 linear.plot=linear.plot+my.theme+geom_point(data=Point.df,aes(x=x,y=y),size=3,shape=4,colour='black')
 linear.plot=linear.plot+geom_line(data=PredVar.df,aes(x=x,y=pred.var),size=1)
-linear.plot=linear.plot+geom_hline(yintercept=max.IVH,linetype=2)+ggtitle("A.")
+linear.plot=linear.plot+geom_hline(yintercept=max.IVH,linetype=2)+ggtitle("A.")+ylab("Scaled variance")
 linear.plot
 
 
@@ -53,13 +53,13 @@ plot(x0,Y)
 PredVar.df=data.frame(x=x0,pred.var=diag(X0%*%solve(crossprod(X2))%*%t(X0)))
 
 library(ggplot2)
[…]

### 2015-03-24 SpatPred/inst/plot_ribbon_IVH_by_tau.R

@@ -1,183 +1,0 @@
-source("./SpatPred/R/spat_pred.R")
-require(maptools)
-require(ggplot2)
-load('Ribbon_data.Rda')
-load('Knot_cell_distances.Rdata')
-K=K.data$K
-set.seed(12345)
-
-#also include quadratic term for ice_conc in dataset
-ice_conc2=(Cur.grid@data[["ice_conc"]])^2
-dist_edge2=(Cur.grid@data[["dist_edge"]])^2
-dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
-dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
-Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
-#attach Ribbon seal counts to data frame
-Count=Ribbon.count
-Cur.grid@data=cbind(Cur.grid@data,Count)
-
-#add in some zero counts in cells with no ice
-Which.not.sampled=c(1:S)
-Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
-Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
-Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
-Mapping=c(Mapping,Which.no.ice)
-Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
-
-Which.not.sampled=c(1:S)
-Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
-
-#run glm model
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
-#formula=Count~1
-#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~dist_mainland+ice_conc+ice_conc2
-formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~ice_conc+ice_conc2
-#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
-#formula=Count~dist_shelf+dist_shelf2
[…]

### 2015-03-24 SpatPred/inst/plot_ribbon_survey_ggplot.R

@@ -1,146 +1,0 @@
-### plot ribbon seal flights and count data using ggplot
-
-library(sp)
-library(rgdal)
-library(nPacMaps)  #from Josh London
-library(rgeos)
-library(doBy)
-library(ggplot2)
-library(plyr)
-
-#First, pull in data
-load('c:/users/paul.conn/git/STabundance/Data_for_ST_plot.Rdata')
-load('c:/users/paul.conn/git/SpatPred/BOSS_Power_photo_data/slr_power_data.RData') #sp data frame including species & area of photo
-load('c:/users/paul.conn/git/BOSS/BOSS/data/AlaskaBeringData.Rdat') #read in "Data" holding grid info
-load('c:/users/paul.conn/git/BOSS/BOSS/data/Effort_points.Rdat')  #read in Sp Points DF "effort_data" for on effort points (use for covariates)
-
-Cur.grid=Data$Grid[[1]]
-
-I.inter=gIntersects(slr_power_data,Data$Grid[[1]],byid=TRUE)
-Points=slr_power_data[which(apply(I.inter,2,'sum')>0),]  #only include effort points that are 'on grid'
-n.points=nrow(Points)
-
-#do some plotting
-plot(Data$Grid[[1]])
-points(Points,pch=19,cex=0.01,col='blue')
-
-n.cells=dim(Cur.grid)[1]
-S=nrow(Cur.grid)  #number of cells
-#first object corresponds to April 4 so start with 17th SpDF
-rownames(Cur.grid@data)=c(1:S)
-Cur.grid<-spChFIDs(Cur.grid,as.character(c(1:S)))  #change ID names to go from 1:S
-Cur.grid[["Ecoregion"]][which(Cur.grid[["Ecoregion"]]==19)]=20  #a few cells are from ecoreg 19 which is never sampled, so convert to 20
-Cur.grid[["Ecoregion"]]=factor(Cur.grid[["Ecoregion"]])
-
-
-data(alaska_dcw)
-data(russia_dcw)
-laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
-                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
-alaska_dcw <- spTransform(alaska_dcw, CRS(laea_180_proj))
-russia_dcw <- spTransform(russia_dcw, CRS(laea_180_proj))
-
-
[…]

### 2015-03-24 SpatPred/inst/run_abundance_sims.R

@@ -1,152 +1,0 @@
-# run_generic_sims.R
-# script to run generic spatio-temporal count data simulations
-source("./SpatPred/R/sim_data_abundance.R")
-source("./SpatPred/R/util_funcs.R")
-source("./SpatPred/R/spat_pred.R")
-
-set.seed(123456)
-n.sims=1#number of simulations at each design point
-S=900
-prop.sampled=0.1
-n.transects=60
-
-i.sim=TRUE
-if(i.sim){
-  for(isim in 1:n.sims){
-    #simulate covariates, abundance
-    Grid=sim_data_generic(S=S,n.covs=4,tau.epsilon=10)
-    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
-    save(Grid,file=Cur.file)
-  }
-  #now simulate count datasets (could possibly use clustered "convenience" sampling, but just using random right now)
-  for(isim in 1:n.sims){
-    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
-    load(Cur.file)
-    Effort=sim_effort(Data=Grid,S=S,type="random",prop.sampled=prop.sampled,n.points=n.transects)
-    Cur.file=paste("./Sim_data/Effort_random",isim,".Rda",sep='')
-    save(Effort,file=Cur.file)
-  }
-}
-
-
-
-
-#s
-
-
-#call estimation routines
-Adj=rect_adj(sqrt(S),sqrt(S))
-for(igen in 1:1){  #loop over generating model to generate data sets
-  for(isim in 1:n.sims){  
-    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
-    load(Cur.file) #load abundance,covariate grid
-    
-    Cur.file=paste("./Sim_data/Effort_random",isim,".Rda",sep='')
-    load(Cur.file) #load Effort 
-    
-    n.transects=length(Effort$Mapping)
-    Grid$cov1.quad=Grid$cov1^2
-    Grid$cov2.quad=Grid$cov2^2
-    Grid$cov3.quad=Grid$cov3^2
-    Grid$cov12=Grid$cov1*Grid$cov2
-    Grid$cov13=Grid$cov1*Grid$cov3
-    Grid$cov23=Grid$cov2*Grid$cov3
[…]

### 2015-03-24 SpatPred/inst/run_design_sims.R

@@ -1,16 +1,15 @@
 # run_generic_sims.R
 # script to run generic spatio-temporal count data simulations
-source("./SpatPred/R/sim_data_abundance.R")
-source("./SpatPred/R/util_funcs.R")
-source("./SpatPred/R/spat_pred.R")
-
+mainDir=getwd()
+subDir="Sim_data"
+dir.create(file.path(mainDir, subDir), showWarnings = FALSE)
 set.seed(123456)
-n.sims=250 #number of simulations at each design point
+n.sims=1000 #number of simulations at each design point
 S=900
 prop.sampled=0.1
-n.transects=60
+n.transects=45
 
-i.sim=FALSE
+i.sim=TRUE
 if(i.sim){
   for(isim in 1:n.sims){
     #simulate covariates, abundance
@@ -27,31 +26,19 @@ if(i.sim){
   for(isim in 1:n.sims){
     Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
     load(Cur.file)
-    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="balanced",prop.sampled=prop.sampled,n.points=40)
+    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="balanced",prop.sampled=prop.sampled,n.points=45)
     Cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
     save(Effort,file=Cur.file)
-    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="clustered",prop.sampled=prop.sampled,n.points=60)    
+    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="clustered",prop.sampled=prop.sampled,n.points=45)    
     Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
     save(Effort,file=Cur.file)
-    #my.formula=~cov1+cov1.quad
-    #Effort=sim_effort(Data=Grid,S=S,my.formula=my.formula,type="IVH",prop.sampled=prop.sampled,n.points=50)    
-    #Cur.file=paste("./Sim_data/Effort_IVH1",isim,".Rda",sep='')
-    #save(Effort,file=Cur.file)
[…]

### 2015-03-24 SpatPred/inst/run_design_simsA.R

@@ -1,176 +1,0 @@
-# run_generic_sims.R
-# script to run generic spatio-temporal count data simulations
-source("./SpatPred/R/sim_data_abundance.R")
-source("./SpatPred/R/util_funcs.R")
-source("./SpatPred/R/spat_pred.R")
-
-set.seed(123456)
-n.sims=250 #number of simulations at each design point
-S=900
-prop.sampled=0.1
-n.transects=45
-
-i.sim=FALSE
-if(i.sim){
-  for(isim in 1:n.sims){
-    #simulate covariates, abundance
-    Grid=sim_data_generic(S=S,n.covs=3,tau.epsilon=10)
-    Grid$cov1.quad=Grid$cov1^2
-    Grid$cov2.quad=Grid$cov2^2
-    Grid$cov3.quad=Grid$cov3^2
-    Grid$cov12=Grid$cov1*Grid$cov2
-    Grid$cov13=Grid$cov1*Grid$cov3
-    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
-    save(Grid,file=Cur.file)
-  }
-  #now simulate count datasets 
-  for(isim in 1:n.sims){
-    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
-    load(Cur.file)
-    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="balanced",prop.sampled=prop.sampled,n.points=45)
-    Cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
-    save(Effort,file=Cur.file)
-    Effort=sim_effort(Data=Grid,S=S,my.formula=NULL,type="clustered",prop.sampled=prop.sampled,n.points=45)    
-    Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
-    save(Effort,file=Cur.file)
-    #my.formula=~cov1+cov1.quad
-    #Effort=sim_effort(Data=Grid,S=S,my.formula=my.formula,type="IVH",prop.sampled=prop.sampled,n.points=50)    
-    #Cur.file=paste("./Sim_data/Effort_IVH1",isim,".Rda",sep='')
-    #save(Effort,file=Cur.file)
-    #my.formula=~(cov1+cov2)^2+cov1.quad+cov2.quad
[…]

### 2015-03-24 SpatPred/inst/run_ribbon_models.R

@@ -1,7 +1,6 @@
-source("./SpatPred/R/spat_pred.R")
-load('Ribbon_data.Rda')
-load('Knot_cell_distances.Rdata')
-K=K.data$K
+
+require(ggplot2)
+data(Ribbon_data)
 set.seed(12345)
 
 #also include quadratic term for ice_conc in dataset
@@ -19,25 +18,12 @@ Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
 #add in some zero counts in cells with no ice
 Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
 Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
-#Mapping=c(Mapping,Which.no.ice)
-#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
 
 #run glm model
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
-#formula=Count~1
-#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~dist_mainland+ice_conc+ice_conc2
 formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~ice_conc+ice_conc2
-#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
-#formula=Count~dist_shelf+dist_shelf2
-#formula=Count~dist_edge+dist_edge2
-#formula=Count~dist_mainland+dist_mainland2
 
 
-
-Control=list(iter=11000,burnin=1000,thin=2,adapt=TRUE,srr.tol=NULL,fix.tau.epsilon=FALSE,predict=TRUE,MH.nu=rep(0.2,length(Mapping)))
+Control=list(iter=11000,burnin=1000,thin=2,adapt=TRUE,srr.tol=NULL,fix.tau.epsilon=FALSE,predict=TRUE,MH.nu=rep(0.5,length(Mapping)))
 Data=Cur.grid
 spat.mod=0
 Offset=Area.photo
[…]

### 2015-03-24 SpatPred/inst/run_ribbon_models_Jan2015.R

@@ -1,301 +1,0 @@
-source("./SpatPred/R/spat_pred.R")
-source("./SpatPred/R/util_funcs.R")
-require(ggplot2)
-load('Ribbon_data.Rda')
-load('Knot_cell_distances.Rdata')
-K=K.data$K
-set.seed(12345)
-
-#also include quadratic term for ice_conc in dataset
-ice_conc2=(Cur.grid@data[["ice_conc"]])^2
-dist_edge2=(Cur.grid@data[["dist_edge"]])^2
-dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
-dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
-Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
-#attach Ribbon seal counts to data frame
-Count=Ribbon.count
-Cur.grid@data=cbind(Cur.grid@data,Count)
-Which.not.sampled=c(1:S)
-Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
-
-#add in some zero counts in cells with no ice
-Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
-Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
-#Mapping=c(Mapping,Which.no.ice)
-#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
-
-#run glm model
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
-#formula=Count~1
-#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~dist_mainland+ice_conc+ice_conc2
-formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~ice_conc+ice_conc2
-#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
-#formula=Count~dist_shelf+dist_shelf2
-#formula=Count~dist_edge+dist_edge2
[…]

### 2015-03-24 SpatPred/inst/run_ribbon_models_ice0.R

@@ -1,415 +1,0 @@
-source("./SpatPred/R/spat_pred.R")
-load('Ribbon_data.Rda')
-load('Knot_cell_distances.Rdata')
-K=K.data$K
-set.seed(12345)
-
-#also include quadratic term for ice_conc in dataset
-ice_conc2=(Cur.grid@data[["ice_conc"]])^2
-dist_edge2=(Cur.grid@data[["dist_edge"]])^2
-dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
-dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
-Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
-#attach Ribbon seal counts to data frame
-Count=Ribbon.count
-Cur.grid@data=cbind(Cur.grid@data,Count)
-Which.not.sampled=c(1:S)
-Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
-
-#add in some zero counts in cells with no ice
-Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
-Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
-Mapping=c(Mapping,Which.no.ice)
-Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
-
-#run glm model
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
-#formula=Count~1
-#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~dist_mainland+ice_conc+ice_conc2
-formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~ice_conc+ice_conc2
-#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
-#formula=Count~dist_shelf+dist_shelf2
-#formula=Count~dist_edge+dist_edge2
[…]

### 2015-03-24 SpatPred/inst/run_ribbon_models_naive.R

@@ -1,437 +1,0 @@
-source("./SpatPred/R/spat_pred.R")
-source("./SpatPred/R/util_funcs.R")
-load('Ribbon_data.Rda')
-load('Knot_cell_distances.Rdata')
-K=K.data$K
-set.seed(12345)
-
-#also include quadratic term for ice_conc in dataset
-ice_conc2=(Cur.grid@data[["ice_conc"]])^2
-dist_edge2=(Cur.grid@data[["dist_edge"]])^2
-dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
-dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
-Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
-#attach Ribbon seal counts to data frame
-Count=Ribbon.count
-Cur.grid@data=cbind(Cur.grid@data,Count)
-Which.not.sampled=c(1:S)
-Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
-
-#add in some zero counts in cells with no ice
-Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
-Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
-#Mapping=c(Mapping,Which.no.ice)
-#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
-
-#run glm model
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
-#formula=Count~1
-#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~dist_mainland+ice_conc+ice_conc2
-formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~ice_conc+ice_conc2
-#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
-#formula=Count~dist_shelf+dist_shelf2
[…]

### 2015-03-24 SpatPred/inst/run_ribbon_models_noRE.R

@@ -1,457 +1,0 @@
-source("./SpatPred/R/spat_pred.R")
-load('Ribbon_data.Rda')
-load('Knot_cell_distances.Rdata')
-K=K.data$K
-set.seed(12345)
-
-#also include quadratic term for ice_conc in dataset
-ice_conc2=(Cur.grid@data[["ice_conc"]])^2
-dist_edge2=(Cur.grid@data[["dist_edge"]])^2
-dist_shelf2=(Cur.grid@data[["dist_shelf"]])^2
-dist_mainland2=(Cur.grid@data[["dist_mainland"]])^2
-Cur.grid@data=cbind(Cur.grid@data,ice_conc2,dist_edge2,dist_shelf2,dist_mainland2)
-#attach Ribbon seal counts to data frame
-Count=Ribbon.count
-Cur.grid@data=cbind(Cur.grid@data,Count)
-Which.not.sampled=c(1:S)
-Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
-
-#add in some zero counts in cells with no ice
-Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
-Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
-#Mapping=c(Mapping,Which.no.ice)
-#Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
-
-#run glm model
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge
-#formula=Count~1
-#formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~dist_mainland+ice_conc+ice_conc2
-formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge
-#formula=Count~ice_conc+ice_conc2
-#formula=Count~dist_shelf+dist_shelf2+ice_conc+ice_conc2+dist_edge+dist_edge2
-#formula=Count~dist_shelf+dist_shelf2
-#formula=Count~dist_edge+dist_edge2
[…]

### 2015-03-24 SpatPred/inst/run_spotted_models.R

@@ -1,214 +1,0 @@
-source("./SpatPred/R/spat_pred.R")
-load('Spotted_data.Rda')
-
-#also include quadratic term for ice_conc in dataset
-ice_conc2=(Cur.grid@data[["ice_conc"]])^2
-Cur.grid@data=cbind(Cur.grid@data,ice_conc2)
-#attach Ribbon seal counts to data frame
-Count=Spotted.count
-Cur.grid@data=cbind(Cur.grid@data,Count)
-Which.not.sampled=c(1:S)
-Which.not.sampled=Which.not.sampled[-which(Which.not.sampled %in% Mapping)]
-
-#add in some zero counts in cells with no ice
-Which.no.ice=which(Cur.grid@data[,"ice_conc"]<0.001)
-Which.no.ice=Which.no.ice[which((Which.no.ice%in%Which.not.sampled)==TRUE)]
-Mapping=c(Mapping,Which.no.ice)
-Area.photo=c(Area.photo,rep(0.1,length(Which.no.ice)))
-
-#run glm model
-Control=list(iter=11000,burnin=1000,thin=2,adapt=TRUE,srr.tol=NULL,fix.tau.epsilon=FALSE,predict=TRUE,MH.nu=rep(0.2,length(Mapping)))
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-formula=Count~1
-
-Data=Cur.grid
-spat.mod=0
-Offset=Area.photo
-Area.adjust=Area.hab
-Assoc=NULL
-Knots=NULL
-Prior.pars=NULL
-#formula=Count~Ecoregion
-formula=Count~dist_mainland+dist_shelf+ice_conc+ice_conc2+dist_edge+Ecoregion
-#formula=Count~dist_mainland+dist_shelf+depth+ice_conc+ice_conc2+dist_edge+Ecoregion
-glm.out <- spat_pred(formula=formula,Data=Data,spat.mod=spat.mod,Offset=Offset,Mapping=Mapping,Area.adjust=Area.adjust,Control=Control,Assoc=Assoc,Knots=Knots,Prior.pars=Prior.pars)
-Control=glm.out$Control
-Control$adapt=FALSE
-Control$iter=120000
-Control$burnin=20000
-Control$thin=50
[…]

### 2015-03-24 SpatPred/inst/sim_pred_process_scratch.R

@@ -1,20 +1,0 @@
-
-require(fields)
-
-S=900
-XY=expand.grid(y=c(sqrt(S):1),x=c(1:sqrt(S)))
-Knot.locations=expand.grid(x=c(1:6)*7-9,y=c(1:6)*7-9)
-n.knots=nrow(Knot.locations)
-
-
-
-#calculate kernel densities at grid cell centroids 
-Distances=matrix(0,S,n.knots)
-
-C.star=Exp.cov(Knot.locations,theta=10)
-CrossC=Exp.cov(Knot.locations,XY,theta=10)
-K=crossprod(CrossC,solve(C.star))
-
-#for(iS in 1:S)Distances[iS,]=sqrt((Knot.locations[,1]-XY[iS,"x"])^2+(Knot.locations[,2]-XY[iS,"y"])^2)
-#K=matrix(dnorm(Distances,0,7),S,n.knots)  #knot sd=5 
-#K=K/rowSums(K)        

### 2015-03-24 SpatPred/inst/test_GAM.R

@@ -1,39 +1,0 @@
-#test GAM 
-library(mvtnorm)
-X=c(0:100)/100
-n=length(X)
-knot.loc=c(0,0.33,0.66,1)
-Y=X+X^3+rnorm(n,0,0.1)
-sd=0.5
-mcmc.iter=10000
-
-K=matrix(0,101,4)
-for(icol in 1:4){
-  K[,icol]=dnorm(X,knot.loc[icol],sd)
-}
-K=K/apply(K,1,'sum')
-
-Alpha.mc=matrix(0.5,4,mcmc.iter)
-Beta=0
-Beta.mc=rep(0,mcmc.iter)
-tau.epsilon=100
-
-KpKinv=solve(crossprod(K))
-for(iiter in 2:mcmc.iter){
-  Delta=Y-K%*%Alpha.mc[,iiter]
-  #update intercept
-  Beta.mc[iiter]=rnorm(1,sum(Delta)/n,1/(n*tau.epsilon+0.01*n))
-
-  #update tau.epsilon
-  Delta=Y-K%*%Alpha.mc[,iiter]-Beta.mc[iiter]
-  tau.epsilon=rgamma(1,0.5*n+1,0.5*crossprod(Delta)+0.01)
-  
-  #update Alpha
-  Alpha.mc[,iiter]=rmvnorm(1,KpKinv%*%crossprod(K,Y-Beta.mc[iiter]),KpKinv/(tau.epsilon+0.01))
-  
-}
-
-Alpha.mean=apply(Alpha.mc,1,'mean')
-plot(X,Y)
-lines(X,K%*%Alpha.mean+mean(Beta.mc))
-#SWEET!

### 2015-04-16 manuscript/SpatPred_PLoS.tex
Commit prior to PLoS One submission

--- 2015-03-24 manuscript/SpatPred_report.tex
@@
-\documentclass[12pt,english]{article}
+% Template for PLoS
+% Version 3.1 February 2015
+%
+% To compile to pdf, run:
+% latex plos.template
+% bibtex plos.template
+% latex plos.template
+% latex plos.template
+% dvipdf plos.template
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended
+% to minimize problems and delays during our production
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% Once your paper is accepted for publication,
+% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
+% the final text of your manuscript.
+%
+% There are no restrictions on package use within the LaTeX files except that
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file.
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig." instead of "Figure".
+% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - tabs/spacing/line breaks within cells to alter layout or alignment
+% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
+% - colors, shading, or graphic objects
+% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
+%
+% Please be sure to include all portions of an equation in the math environment.
+%
+% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column.
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
 
-\usepackage{amsmath,amssymb,amsthm,epsfig,lineno,rotfloat,psfrag,natbib,caption,setspace,url,bm,geometry}
-\usepackage{ecology}
-\usepackage{graphicx}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
-\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
 
-%\setlength{\evensidemargin}{0in} \setlength{\oddsidemargin}{0in}
-%\setlength{\topmargin}{-0.65in} \setlength{\textwidth}{6.5in}
-%\setlength{\textheight}{9.5in} \setlength{\topskip}{0in}
-%\setlength{\headheight}{0in}
+% Use Unicode characters when possible
+\usepackage[utf8]{inputenc}
 
-%\newcommand{\yst}{\ensuremath{y_{st}}}
-%\newcommand{\tyst}{\ensuremath{\tilde y_{st}}}
-%\newcommand{\tzs}{\ensuremath{\tilde z_{s}}}
-%\newcommand{\zs}{\ensuremath{z_{s}}}
-%\newcommand{\bg}{\ensuremath{\boldsymbol{\gamma}}}
-%\newcommand{\bxs}{\ensuremath{\mathbf{x}_{s}}}
-%\newcommand{\bxst}{\ensuremath{\mathbf{x}_{st}}}
-%\newcommand{\bK}{\ensuremath{\mathbf{K}}}
-%\newcommand{\bn}{\ensuremath{\boldsymbol{\eta}}}
-%\newcommand{\bb}{\ensuremath{\boldsymbol{\beta}}}
-%\newcommand{\ba}{\ensuremath{\boldsymbol{\alpha}}}
-%\newcommand{\by}{\ensuremath{\mathbf{y}}}
-%\newcommand{\bz}{\ensuremath{\mathbf{z}}}
-%\newcommand{\bX}{\ensuremath{\mathbf{X}}}
-%\newcommand{\be}{\ensuremath{\boldsymbol{\epsilon}}}
-%\newcommand{\bQ}{\ensuremath{\mathbf{Q}}}
-%\newcommand{\bO}{\ensuremath{\boldsymbol{\Omega}}}
-%\newcommand{\bm}{\ensuremath{\boldsymbol{\mu}}}
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
 
-\bibpunct{(}{)}{,}{a}{}{;}
+% fixltx2e package for \textsubscript
+\usepackage{fixltx2e}
 
-\bibliographystyle{ecology}
-\raggedbottom
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
 
-\captionsetup[table]{margin=0pt,font=small,labelfont={sc},justification=justified,labelsep=period}
-\captionsetup[figure]{margin=0pt,font=small,labelfont={sc},justification=justified,labelsep=period,name=Fig}
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% rotating package for sideways tables
+\usepackage{rotating}
+
+% Remove comment for double spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
 
 \begin{document}
-\begin{spacing}{1.9}
+\vspace*{0.35in}
 
-\begin{center}
-On extrapolating past the range of observed data when making statistical predictions in ecology
-\bigskip\\
-\normalsize
-{\sc Paul B. Conn$^{1,}$\footnotemark[2] and
-Devin S. Johnson$^1$, Peter L. Boveng$^1$ }\smallskip\\
-$^1${\em National Marine Mammal Laboratory, NOAA, National Marine Fisheries Service,
+% Title must be 250 characters or less.
+% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
+\begin{flushleft}
+{\Large
+\textbf\newline{On extrapolating past the range of observed data when making statistical predictions in ecology}
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Paul B. Conn\textsuperscript{1,*},
+Devin S. Johnson\textsuperscript{1},
+Peter L. Boveng\textsuperscript{1},
+\\
+\bigskip
+\bf{1} National Marine Mammal Laboratory, NOAA, National Marine Fisheries Service,
 Alaska Fisheries Science Center, 7600 Sand Point Way NE, Seattle,
-WA 98115 USA }\\ \medskip
-\end{center}
-\footnotetext[2]{Email: paul.conn@noaa.gov}
+WA 98115 USA
+\\
+\bigskip
 
-\raggedright \setlength{\parindent}{0.3in}
-%\renewcommand{\baselinestretch}{1.8}\normalsize
-\clubpenalty=0
- \linenumbers
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* paul.conn@noaa.gov
 
-{\em Abstract.\ } Ecologists are increasingly using statistical models to predict animal abundance and occurrence in unsampled locations.
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
+Ecologists are increasingly using statistical models to predict animal abundance and occurrence in unsampled locations.
 The reliability of such predictions depends on a number of factors, including sample size, how far prediction locations are from the observed data, and similarity of predictive covariates in locations where data are gathered to locations where predictions are desired.
@@
 We suggest that ecologists routinely use diagnostics such as the gIVH and related metrics (e.g., prediction variance) to help gauge the reliability of predictions from statistical models.
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step.
+% Author Summary not valid for PLOS ONE submissions.
+\linenumbers
 
-{\em Key words: Abundance, Extrapolation, Independent Variable Hull, Prediction variance, Spatial regression, Species distribution model}
-
-\centerline{\sc Introduction}
-
+\section*{Introduction}
 In ecology and conservation, a common goal is to make predictions about an unsampled random variable given a limited sample from the target population.
-For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_i$, we often desire to predict a new observation $y_i$ at $i$ (where $i$ can be a design point or a spatial location).
-For instance, we might use a generalized linear model \citep[GLM;][]{McCullaghNelder1989} or one of its extensions to predict species density or occurrence in a new location.
+For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_i$, we often desire to predict a new observation $y_i$ at \textit{i} (where \textit{i} can be a design point or a spatial location).
+For instance, we might use a generalized linear model (GLM;~\cite{McCullaghNelder1989}) or one of its extensions to predict species density or occurrence in a new location.
 Such predictions can be of direct use to conservation and management, for instance, in estimating population abundance or distribution, and for projecting shifts in species range as a function of climate change.
-Spatially explicit estimates of abundance are also useful for testing theory related to biogeography or biodiversity \citep[e.g., neutral theory;][]{Hubbell2001}, and for accurately estimating the strength of density dependence \citep{ThorsonEtAlInPress}.
+Spatially explicit estimates of abundance are also useful for testing theory related to biogeography or biodiversity (e.g., neutral theory;~\cite{Hubbell2001}), and for accurately estimating the strength of density dependence~\cite{ThorsonEtAlInPress}.
 
@@
 past the range of observed data.
-This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new response $y_i$ whenever $x_i < \min({\bf x})$ or $x_i > \max({\bf x})$ (where $x_i$ is an independent variable measured at point $i$).
+This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new response $y_i$ whenever $x_i < \min({\bf x})$ or $x_i > \max({\bf x})$ (where $x_i$ is an independent variable measured at point \textit{i}).
 But what about more complicated situations
-where there are multiple explanatory variables, or when one uses a spatial regression model to account for the residual spatial autocorrelation that is inevitably present in patchy ecological data \citep{LichsteinEtAl2002}?
+where there are multiple explanatory variables, or when one uses a spatial regression model to account for the residual spatial autocorrelation that is inevitably present in patchy ecological data~\cite{LichsteinEtAl2002}?
 How reliable are spatially- or temporally-explicit predictions in sophisticated models for animal abundance and occurrence?
@@
 making robust predictions at new combinations of explanatory variables.  The issue is sometimes considered more of a
-philosophical problem than a statistical one, and has even been likened to soothsaying \citep{EhrenbergBound1993}.
+philosophical problem than a statistical one, and has even been likened to soothsaying~\cite{EhrenbergBound1993}.
 To our mind, the reliability of predictions from statistical models is likely a function of several factors, including (i) the intensity of sampling,
@@
 
-Our aim in this paper is to investigate one possibility for defining extrapolation in the GLM and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatio-temporal regression models (STRMs).
-In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers \citep{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points.
+In this paper, we investigate one possibility for defining extrapolation in the GLM and its extensions, including generalized additive models (GAMs;~\cite{HastieTibshirani1999,Wood2006}) and spatio-temporal regression models (STRMs).
+In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers~\cite{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points.
 Application of the gIVH and related criterion (e.g., prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction and survey design.
 We illustrate use of the gIVH on simulated count data, and on a species distribution model (SDM) for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea.
-
-\centerline{\sc GENERALIZING THE INDEPENDENT VARIABLE HULL}
+% You may title this section "Methods" or "Models".
+% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
+% authors may use "Analysis"
+\section*{Materials and Methods}
+\subsection*{Generalizing the independent variable hull}
 
@@
 
-In the context of outlier detection, Cook (\citeyear{Cook1979}) defined an independent variable hull (IVH) as the smallest convex set containing all design points of a full-rank linear regression model.
-Linear regression models are often written in matrix form, i.e.,
+In the context of outlier detection, Cook~\cite{Cook1979} defined an independent variable hull (IVH) as the smallest convex set containing all design points of a full-rank linear regression model.
+Linear regression models are often written in matrix form; that is,
 \begin{linenomath*}
@@
 \end{linenomath*}
-where ${\bf Y}$ are observed responses, ${\bf X}$ is a so-called design matrix that includes explanatory variables \citep[see e.g.,][]{Draper1966}, and $\boldsymbol{\epsilon}$ represent normally distributed residuals (here and throughout the paper, bold symbols will be used to denote vectors and matrices).
+where ${\bf Y}$ are observed responses, ${\bf X}$ is a so-called design matrix that includes explanatory variables~\cite{Draper1966}, and $\boldsymbol{\epsilon}$ represent normally distributed residuals (here and throughout the paper, bold symbols will be used to denote vectors and matrices).
 Under this formulation, the IVH is defined relative to the hat matrix, ${\bf V}_{\rm LR}={\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$ (where the subscript ``LR" denotes linear regression).
@@
 \end{linenomath*}
-Cook (\citeyear{Cook1979}) used this concept to identify influential observations and possible outliers, arguing that design points near the edge of the IVH are deserving of special attention.
+Cook~\cite{Cook1979} used this concept to identify influential observations and possible outliers, arguing that design points near the edge of the IVH are deserving of special attention.
 Similarly, points outside the IVH should be interpreted with caution.
@@
 
-Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes independent and identically distributed ($iid$) Gaussian error.
+Cook's~\cite{Cook1979} formulation for the IVH is particular to linear regression analysis, which assumes independent and identically distributed (\textit{iid}) Gaussian error.
 Thus, it is not directly applicable to generalized models, such as those including alternative error distributions (e.g., Poisson, binomial) or spatial random effects.
 Further, the hat matrix is not necessarily well defined for models with more general spatial structure.
-However, since the hat matrix is proportional to prediction variance, Cook (1979) notes that design points with maximum prediction variance will be located on the boundary of the IVH.
+However, since the hat matrix is proportional to prediction variance, Cook~\cite{Cook1979} notes that design points with maximum prediction variance will be located on the boundary of the IVH.
 We therefore define a generalized independent variable hull (gIVH) as the set of all potential design points $\mathcal{S}_0$ (note that $\mathcal{S}_0$ can include both observed and unobserved design points) such that
@@
 \end{linenomath*}
-where $i \in \mathcal{S}_0$, $\lambda_i$ corresponds to the mean prediction at $i$, $\mathcal{S}$ denotes the set of observed design points, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.
+where $i \in \mathcal{S}_0$, $\lambda_i$ corresponds to the mean prediction at \textit{i}, $\mathcal{S}$ denotes the set of observed design points, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.
 
@@
   Y_i \sim f_Y(g^{-1}(\mu_i)),
+  \label{eq:Y}
 \end{equation}
 \end{linenomath*}
-where $f_Y$ denotes a probability density or mass function (e.g., Bernoulli, Poisson), $g$ gives a link function, and
+where $f_Y$ denotes a probability density or mass function (e.g., Bernoulli, Poisson), \textit{g} gives a link function, and
 $\mu_i$ is a linear predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
@@
 where the $\boldsymbol{\epsilon}$ represent Gaussian error, ${\bf X}_{aug}$ denotes an augmented design matrix, and $\boldsymbol{\beta}_{aug}$ denote an augmented vector of parameters.
-For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see Appendix A for examples of how numerous types of models can be written in this form).
+For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see \nameref{S1_Text} for examples of how numerous types of models can be written in this form).
 
@@
 \end{linenomath*}
-where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; Appendix A).
-This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method \citep{Dorfman1938,VerHoef2012b} to approximate prediction variance on the real scale (i.e., the scale in which data are measured; see Appendix A).
-Unlike Cook's original formulation, the gIVH will often involve unknown parameters.
-For most of the following treatment, we shall assume that data have already been collected (see Discussion for comments on the potential use of the gIVH in survey planning), and that informative probability distributions can be specified for these parameters.
+where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; \nameref{S1_Text}).
+This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method~\cite{Dorfman1938,VerHoef2012b} to approximate prediction variance on the real scale (i.e., the scale in which data are measured; see \nameref{S1_Text}).
+If a Bayesian approach to estimation is used, computation is even easier, as one can simply base thresholding on posterior predictive variance of $g^{-1}(\mu_i)$.
+A Bayesian mode of inference is perhaps the most natural to use in these problems, for the simple reason that unlike Cook's original formulation, the gIVH will often involve unknown parameters.
+For most of the following treatment, we shall assume that data have already been collected (see Discussion for comments on the potential use of the gIVH in survey planning), and that informative prior probability distributions can be specified for unknown parameters.
 
-We propose to use the gIVH in much the same manner as Cook (1979).
+We propose to use the gIVH in much the same manner as Cook~\cite{1979}.
 In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).
@@
 
-\subsection{Computing}
+\subsection*{Computing}
 
-We developed a package \texttt{SpatPred} in the R statistical programming environment \citep{RTeam2012} to simulate data and conduct all analyses.
+We developed a package \texttt{SpatPred} in the R statistical programming environment~\cite{RTeam2012} to simulate data and conduct all analyses.
 The seal dataset is included as part of this package, and is available at \url{https://github.com/pconn/SpatPred}.
@@
 
-\centerline{\sc Examples}
-
-\subsection{Simulation study}
+\subsection*{Simulation study}
 
@@
 For each of 100 simulations, we generated animal abundance over a $30 \times 30$ grid assuming that animal density was homogeneous in each grid cell.
-Animal abundance was generated as a function of three hypothetical spatially autocorrelated habitat covariates (Appendix B).
-For each simulated landscape, we conducted virtual surveys of $n=45$ survey units using two different designs: (1) a spatially balanced sample \citep{StevensOlsen2004}, and (2) a convenience sample where the probability of sampling was greater for cells closer to a ``base of operations" located in the middle of the survey grid.
+Animal abundance was generated as a function of three hypothetical spatially autocorrelated habitat covariates (\nameref{S2_Text}).
+For each simulated landscape, we conducted virtual surveys of $n=45$ survey units using two different designs: (1) a spatially balanced sample~\cite{StevensOlsen2004}, and (2) a convenience sample where the probability of sampling was greater for cells closer to a ``base of operations" located in the middle of the survey grid.
 The former approach preserves randomness while seeking a degree of regularity when distributing sampling locations across the landscape, while the latter may be easier to implement logistically.
@@
 For ease of presentation and analysis, we assumed detection probability was 100\% in each quadrat.
-Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (Appendix A).
-The fixed effects components of the GLM and STRM were configured to have both linear and quadratic covariate effects and first-order interactions, while the GAM expressed log-density as a function of smooth terms for each covariate (Appendix B).
+Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (\nameref{S1_Text}).
+The fixed effects components of the GLM and STRM were configured to have both linear and quadratic covariate effects and first-order interactions, while the GAM expressed log-density as a function of smooth terms for each covariate (\nameref{S2_Text}).
 Each model was provided with two of the three covariates used to generate the data.
 
-For each simulated data set and model structure, we calculated the posterior predictive variance and resulting $\text{gIVH}$ as in Eq.
+For each simulated data set and model structure, we calculated the posterior predictive variance and resulting gIVH as in Eq.
 \ref{eq:gIVH}.
 We then calculated posterior predictions of animal abundance within and outside of each gIVH in order to gauge bias as a function of this restriction.
-Specifically, the performance of $\text{gIVH}$ may help decide its utility in limiting the scope of inference once data have been collected and analyzed, and perhaps point out areas worthy of additional sampling.
-A fuller, technical description of the simulation study design and several illustrations are provided in Appendix B.
+Specifically, the performance of the gIVH may help decide its utility in limiting the scope of inference once data have been collected and analyzed, and perhaps point out areas worthy of additional sampling.
+A fuller, technical description of the simulation study design is provided in \nameref{S2_Text}; a visual depiction of a single simulation replicate is displayed in
+Fig. \ref{fig:sim_maps}.
 
-Posterior predictions from simulations indicated that the distribution for proportional error in total abundance was right skewed when statistical inference was made with regard to the entire survey area (Fig.
-B3).
-This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.
-The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.
-For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude ($\approx 0.3$) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
-By contrast, proportional bias was close to zero when inference was restricted to the gIVH.
-Interestingly, the radial basis GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.
+\subsection*{Ribbon seal SDM}
 
-\subsection{Ribbon seal SDM}
+As part of an international effort, researchers with the U.S.
+National Marine Fisheries Service conducted aerial surveys over the eastern Bering Sea in 2012 and 2013.
+Agency scientists used infrared video to detect seals that were on ice, and collected simultaneous digital photographs to provide information on species identity.
+For this study, we use spatially referenced count data from photographed ribbon seals, {\it Phoca fasciata} on a subset of 10 flights flown over the Bering Sea from April 20-27, 2012.
+We limited flights to a one week period because sea ice melts rapidly in the Bering Sea in the spring, and modeling counts over a longer duration would likely require addressing how sea ice and seal abundance changes over both time and space~\cite{ConnEtAl2015}.
+However, limiting analysis to a one week period makes the assumption of static sea ice and seal densities tenable~\cite{ConnEtAl2014}.
 
-To demonstrate application of the gIVH on a real dataset, we consider the case of estimating ribbon seal relative abundance from count data obtained during aerial surveys of the Bering Sea in April, 2012 (Fig.
-\ref{fig:flights}).
-In this case, we wish to model seal counts as a function of habitat covariates and possible spatial autocorrelation, and to use predictions from this fitted model to estimate relative abundance for the entire study area.
-For further background and details on these data, we refer the reader to Appendix C.
+Our objective with this dataset will be to model seal counts on transects through 25km by 25km grid cells as a function of habitat covariates and possible spatial autocorrelation.
+Estimates of apparent abundance can then be obtained by summing predictions across grid cells.
+Figure \ref{fig:covs} show explanatory covariates gathered to help predict ribbon seal abundance.
+These data are described in fuller detail by~\cite{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors.
+Since our focus in this paper is on illustrating spatial modeling concepts, we devote our efforts to the comparably easier problem of estimating apparent abundance (i.e., uncorrected for vagaries of the detection process).
 
@@
 issue with spatial prediction: abundance of ribbon seals appears to be maximized in the southern and/or southeast quadrant of the surveyed area.
-Predicting abundance in areas further south and east may thus prove problematic, as the values of several explanatory covariates (Fig.
-C1) are also maximized in these regions.
-To illustrate, let $Y_i$ denote the ribbon seal count ($Y_i$) obtained in sampled grid cell $i$.
-Suppose that counts arise according to a log-Gaussian Cox process, such that
+Predicting abundance in areas farther south and east may thus prove problematic, as the values of several explanatory covariates (Fig.
+\ref{fig:covs}) are also maximized in these regions.
+
+We start by fitting hierarchical GLMs, STRMs, and GAMs to the ribbon seal data.
+To accommodate incomplete coverage of grid cells and account for non-target habitat, we adapted Eqs.
+\ref{eq:Y}-\ref{eq:mu-aug} as follows.
+First, let $Y_i$ denote the ribbon seal count ($Y_i$) obtained in sampled grid cell $i$.  Suppose that counts arise according to a log-Gaussian Cox process, such that
 \begin{eqnarray}
   \label{eqn:Cox}
-  Y_i \sim {\rm Poisson}(\lambda_i) \mbox{ and } \\
-  \log(\lambda_i) = \log(P_i) + \log(A_i) + \theta_i + \epsilon_i, \nonumber
+  Y_i & \sim & {\rm Poisson}(\lambda_i) \mbox{ and } \\
+  \log(\lambda_i) & = & \log(P_i) + \log(A_i) + \theta_i + \epsilon_i, \nonumber
 \end{eqnarray}
 where $P_i$ gives the proportion of area surveyed in grid cell $i$, $A_i$ gives the proportion of cell $i$ that is seal habitat, $\theta_i$ is a linear predictor, and $\epsilon_i$ is normally distributed $iid$ error.
-By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs, GAMs, and STRMs (see Appendix A).
+By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs, GAMs, and STRMs (see \nameref{S1_Text})
 
-We start by fitting a hierarchical GLM, GAM, and STRM to the full suite of predictor covariates (Fig.
-C1).
-For the GLM and STRM, we fit a model with linear effects of all predictor variables and an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}.
-For the GAM, smooth effects were included for each covariate, with 8 knots each.
-As with simulated data examples, we adopted a Bayesian perspective and conducted estimation and posterior prediction using MCMC (see Appendix A for estimation algorithm details and Appendix C for further numerical details related to the ribbon seal example).
+The fixed effects component of the GLM and STRM included linear effects of all explanatory covariates (Fig.
+\ref{fig:covs}, as well as a quadratic effect for sea ice concentration.
+For the STRM, we imposed a restricted spatial regression (RSR) formulation for spatially autocorrelated random effects, where dimension reduction was accomplished by only selecting eigenvectors of the spectral decomposition associated with eigenvalues that were greater than 0.5 (see Text S1 for additional information on model structure).
+For the GAM, we employed smooth terms with 8 knots for each covariate.
+This included 4 knots at the minimum, 1/3 and 2/3 quantiles, and the maximum observed covariate values, in addition to 4 knots above and below the range of observed data to reduce the magnitude of known biased endpoint behavior that can occur with radial basis function implementations of GAMs.
+
+As with simulated data examples, we adopted a Bayesian perspective and conducted estimation using MCMC (see Text S1 for algorithm details and information on prior distributions) with 60,000 iterations where the first 10,000 iterations were discarded as a burn-in.
+We generated posterior predictions of ribbon seal abundance across the landscape as
+\begin{equation}
+  N_i \sim {\rm Poisson}(A_i \lambda_i),
+\end{equation}
+and calculated the gIVH as in Eqn. \ref{eq:gIVH}, with delta method modifications as specified in \nameref{S1_Text}.
 
@@
 However, this is not ideal in that one does not get an abundance estimate for the whole study area, and it may be difficult to compare abundance from one year to the next using this approach.
-Second, one could consider a simpler estimation model, such as one post-stratifying by ecoregion \citep[as in][]{ConnEtAl2014}.
+Second, one could consider a simpler estimation model, such as one post-stratifying by ecoregion (as in~\cite{ConnEtAl2014}).
 Finally, one could build in a priori knowledge of habitat preferences into the model structure.
 We adopted the latter solution, incorporating ``pseudo-absences" (i.e., zero counts where sampling was not conducted) in locations where it would have been (nearly) impossible to detect seals.
-Specifically, we inserted pseudo-absences in cells where ice concentrations were $<0.1\%$.
+Specifically, we inserted pseudo-absences in cells where ice concentrations were \textless 0.1\%.
 This solution seemed the most logical, as many of the large, anomalous predictions were over open water along the southern edge of the study area, where we would have obtained zero counts had they been surveyed.
@@
 
-Fitting our three models to the augmented dataset with pseudo-absences, most predictions occurred within the gIVH (Fig.
+\section*{Results}
+
+\subsection*{Simulation study}
+
+Posterior predictions from simulations indicated that the distribution for proportional error in total abundance was right skewed when statistical inference was made with regard to the entire survey area (Fig.
+\ref{fig:sim_boxplots}).
+This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.
+The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.
+For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude ($\approx 0.3$) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
+By contrast, proportional bias was close to zero when inference was restricted to the gIVH.
+Interestingly, the radial basis GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.
+
+\subsection*{Ribbon Seal SDM}
+
+Fitting our three ribbon seal SDMs to the augmented dataset with pseudo-absences, most predictions occurred within the gIVH (Fig.
 \ref{fig:ribbon_plot}).
@@
 Posterior summaries of abundance across the study area were very comparable (albeit with a longer right hand tail for the STRM), with $5\%$, $50\%$, and $95\%$ posterior prediction quantiles as follows: GLM (48686,64836,93927); GAM(46120,61214,90873); STRM(41039,63717,194095).
-We note that these estimates are for example illustration only, as they are uncorrected for imperfect detection \citep[e.g., incomplete detection of thermal cameras, animals that were unavailable for sampling because they were in the water, species misidentification; see][]{ConnEtAl2014}.
+We note that these estimates are for example illustration only, as they are uncorrected for imperfect detection (e.g., incomplete detection of thermal cameras, animals that were unavailable for sampling because they were in the water, species misidentification;~\cite{ConnEtAl2014}).
 Our approach here was to examine extrapolation and prediction error using relatively simple models, with the understanding that such effects are also likely to occur in complex models with more realistic observation processes.
 
-\centerline{\sc Discussion}
+\section*{Discussion}
 
@@
 
-Our intent in writing this paper is to raise awareness of potential problems with extrapolation bias in statistical models, and to provide an additional tool (the gIVH) to help diagnose its presence.
-Other approaches for selecting models to reduce overfitting, such as cross validation \citep[e.g.,][]{PicardCook1984}, are also useful for this purpose, but may not entirely eliminate the problem (particularly for sparse datasets).
+Our intent is to raise awareness of potential problems with extrapolation bias in statistical models, and to provide an additional tool (the gIVH) to help diagnose its presence.
+Other approaches for selecting models to reduce overfitting, such as cross validation~\cite{PicardCook1984}, are also useful for this purpose, but may not entirely eliminate the problem (particularly for sparse datasets).
 In fields such as species distribution modeling and model-based abundance estimation, the goal for analysts is often to build predictive maps of species abundance and/or occurrence using a limited number of sample locations.
@@
 One area that gIVH ideas may also prove useful is in formulating survey designs.
-The topic of optimal (or near-optimal) spatial design has received considerable attention in the statistical literature, often in the context of designing environmental monitoring systems \citep{Muller2007}.
+The topic of optimal (or near-optimal) spatial design has received considerable attention in the statistical literature, often in the context of designing environmental monitoring systems~\cite{Muller2007}.
 Optimal designs can be sensitive to the structure of the estimation model that is used, so that tailoring a survey design to a particular model can be somewhat dangerous if there is uncertainty about the ultimate ``best" structure for the model used to relate animal abundance and occurrence to available covariates.
@@
 
+\section*{Supporting Information}
+
+% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
+
+\subsection*{S1 Text}
+\label{S1_Text}
+{\bf Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model.}
+
+\subsection*{S2 Text}
+\label{S2_Text}
+{\bf Full details of simulation study examining predictive extrapolation.}
+
+\section*{Acknowledgments}
+We thank J. Thorson, J. Ver Hoef, and P. De Valpine for helpful discussion on an earlier version of this mnauscript.  Funding
+for Bering Sea aerial surveys was provided by the U.S. National Oceanic and Atmospheric
+Administration and by the U.S. Bureau of Ocean Energy Management (Interagency
+Agreement M12PG00017).
+The views and conclusions in this article do not necessarily represent findings or policy of the U.S.
+National Oceanic and Atmospheric Administration.
+Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.
+Government.
+
+\nolinenumbers
+
+%\section*{References}
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% OR
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here.
+%
+\bibliographystyle{plos2015}
 \bibliography{master_bib}
 
-\centerline{\sc Supplemental Material}
+%\begin{thebibliography}{10}
 
-\centerline{\bf Appendix A}
-Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model ({\it Ecological Archives} XXXX-XXX-A1).
-
-\centerline{\bf Appendix B}
-Full details of simulation study examining predictive extrapolation ({\it Ecological Archives} XXXX-XXX-A2).
-
-\centerline{\bf Appendix C}
-Additional details on ribbon seal count data ({\it Ecological Archives} XXXX-XXX-A3).
-
-\clearpage
-%FIGURE 1. A ribbon seal, {\it Phoca fasciata}; the focus of spatial %modeling efforts in this paper.
-
-FIGURE 1.
-Example IVHs constructed from simulated data.
-In (A) and (B), linear regression is used to relate a response variable to a single covariate, $x$, obtained at locations denoted with an ``x".
-Using $x$ as a simple linear effect (A), only predictions less than the minimum observed value of $x$ or greater than the maximum value of $x$ are outside the IVH (shaded area), as scaled prediction variance in these areas (solid line) is greater than the maximum scaled prediction variance for observed data (dashed line).
-Using both linear and quadratic effects (B), some intermediate points are also outside the IVH.
-When both linear and quadratic effects of two covariates ($x_1$ and $x_2$) are modeled, the IVH is more nuanced and depends on whether interactions are omitted (C) or included (D).
-
-FIGURE 2.
-Aerial survey tracks over the Bering Sea, April 22-29, 2012 (blue lines) overlayed on a tesselated study area consisting of 25km by 25km grid cells (gray lines).
-Dark gray indicates land, while the brown line indicates a 1000m depth contour, and the orange line shows the U.S Exclusive Economic Zone (EEZ) boundary.
-Colored pixels indicate ribbon seal counts along aerial transects.
-The average effective area surveyed in each grid cell was approximately $2.6km^2$ ($0.4\%$).
-Note that surveys were designed to target multiple seal species, several of which had high densities further north (results not shown).
-
-FIGURE 3.
-Posterior mean predictions of ribbon seal apparent abundance across the eastern Bering sea from models fit to survey data.
-Each row gives result for different model types (GLM, GAM, or STRM, respectively); left column plots give results for naive runs without pseudo-absences, while plots in the right column give predictions for runs where psuedo-absence data (i.e., 0 counts in cells with less than $0.1\%$ ice) were included.
-Cells highlighted in black indicate those where predictions were outside the generalized independent variable hull (gIVH).
-
-\clearpage
-\centerline{\sc Figures}
-
-%\begin{figure}[!h]
-%\begin{center}
-%\includegraphics[width=6in]{ribbon}
-%\end{center}
-%\caption{ }
-%\label{fig:ribbon}
-%\end{figure}
+%\end{thebibliography}
 
 \begin{figure}[!h]
-\begin{center}
-\includegraphics[width=6in]{IVH_simple2.pdf}
-\end{center}
-\caption{ }
+\caption{{\bf Example IVHs constructed from simulated data}.
+In (A) and (B), linear regression is used to relate a response variable to a single covariate, \textit{x}, obtained at locations denoted with an ``x".
+Using \textit{x} as a simple linear effect (A), only predictions less than the minimum observed value of \textit{x} or greater than the maximum value of \textit{x} are outside the IVH (shaded area), as scaled prediction variance in these areas (solid line) is greater than the maximum scaled prediction variance for observed data (dashed line).
+Using both linear and quadratic effects (B), some intermediate points are also outside the IVH.
+When both linear and quadratic effects of two covariates (\textit{x}\textsubscript{1} and \textit{x}\textsubscript{2}) are modeled, the IVH is more nuanced and depends on whether interactions are omitted (C) or included (D).
+}
 \label{fig:IVH}
@@
 \begin{figure}[!h]
-\begin{center}
-\includegraphics[width=6in]{BOSS_survey_ribbon.pdf}
-\end{center}
-\caption{ }
+\caption{{\bf Depiction of a single simulation scenario}.
+Panels (A-C) give simulated covariate values, panel D gives true animal abundance, (E) gives estimated abundance from a GLM run on count data from a spatially balanced survey design, and (F) gives abundance from a GLM applied to count data from a convenience survey.
+In (E-F), predictions outside the gIVH are represented by black boxes, and sampling locations are represented with an x.
+For the convenience sample, the median posterior abundance prediction for the entire survey area is 57\% greater than true abundance when inference is made to the whole study area.
+When inference is limited to the gIVH, median posterior abundance was just 16\% greater than true abundance.}
+\label{fig:sim_maps}
+\end{figure}
+
+\begin{figure}[!h]
+\caption{{\bf Assembled covariates used to help explain and predict ribbon seal relative abundance in the eastern Bering Sea}.
+Covariates include distance from mainland (\texttt{dist\_mainland}), distance from 1000m depth contour (\texttt{dist\_shelf}), average remotely sensed sea ice concentration while surveys were being conducted (\texttt{ice\_conc}), and distance from the southern sea ice edge (\texttt{dist\_edge}).
+All covariates except ice concentration were standardized to have a mean of 1.0 prior to plotting and analysis.
+}
+\label{fig:covs}
+\end{figure}
+
+\begin{figure}[!h]
+\caption{{\bf Aerial survey tracks over the Bering Sea, April 22-29, 2012}.
+Survey tracks are shown in blue, and are overlayed on a tesselated study area consisting of 25km by 25km grid cells (gray lines).
+Dark gray indicates land, while the brown line indicates a 1000m depth contour, and the orange line shows the U.S Exclusive Economic Zone (EEZ) boundary.
+Colored pixels indicate ribbon seal counts along aerial transects.
+The average effective area surveyed in each grid cell was approximately 2.6km\textsuperscript{2} (0.4\%).
+Note that surveys were designed to target multiple seal species, several of which had high densities further north (results not shown).
+}
 \label{fig:flights}
@@
 \begin{figure}[!h]
-\begin{center}
-\includegraphics[width=6in]{ribbon_maps.pdf}
-\end{center}
-\caption{ }
+\caption{{\bf Boxplots summarizing proportional error in abundance from the simulation experiment}.
+Each boxplot summarizes the distribution of proportional error in the posterior predictive median of abundance as a function of estimation model (x-axis), survey design (columns) and whether or not inference was restricted to the gIVH (rows).
+The lower and upper limits of each box correspond to first and third quartiles, while whiskers extend to the lowest and highest observed bias within 1.5 interquartile range units from the box.
+Outliers outside of this range are denoted with points.
+Horizontal lines within boxes denote median bias.
+The two numbers located below each boxplot indicate mean bias (upper number) and the number of additional outliers for which proportional bias was greater than 2.0 (lower number).
+}
+\label{fig:sim_boxplots}
+\end{figure}
+
+\begin{figure}[!h]
+\caption{{\bf Posterior mean predictions of ribbon seal apparent abundance across the eastern Bering sea from models fit to survey data}.
+Each row gives result for different model types (GLM, GAM, or STRM, respectively); left column plots give results for naive runs without pseudo-absences, while plots in the right column give predictions for runs where psuedo-absence data (i.e., 0 counts in cells with \textless 0.1\% ice) were included.
+Cells highlighted in black indicate those where predictions were outside the generalized independent variable hull (gIVH) }
 \label{fig:ribbon_plot}
@@
 
-\clearpage
-
-\end{spacing}
 \end{document}

### 2015-04-16 manuscript/Text_S1.tex
Commit prior to PLoS One submission

--- 2015-03-24 manuscript/AppendixA.tex
@@
 \renewcommand{\em}{\it}
-\renewcommand\thefigure{A\arabic{figure}}
-\renewcommand\thetable{A\arabic{table}}
-\renewcommand\theequation{A.\arabic{equation}}
+\renewcommand\thefigure{S1\arabic{figure}}
+\renewcommand\thetable{S1\arabic{table}}
+\renewcommand\theequation{S1.\arabic{equation}}
 
 \begin{document}
-
-\begin{center} \bf {\large On extrapolating past the range of observed data when making statistical predictions in ecology}
-
-\vspace{0.7cm}
-Paul B. Conn$^{1*}$, Devin S. Johnson$^1$, and Peter L. Boveng$^1$
-\end{center}
-\vspace{0.5cm}
-
-\rm
-\small
-
-\it $^1$National Marine Mammal Laboratory, Alaska Fisheries Science Center,
-NOAA National Marine Fisheries Service,
-Seattle, Washington 98115 U.S.A.\\
 
@@
 \begin{center}
-Appendix A: Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model
+S1 Text: Model formulation, Gibbs sampling algorithms, and gIVH calculations for certain classes and extensions of the generalized linear model
 \bigskip
@@
 In addition, reduced dimension spatial models such as RSR lighten computational burden while still accounting for course-scale spatial autocorrelation \citep[see e.g.][]{LatimerEtAl2009,Wikle2010,Hughes2013}.
-It turns out that RSR models can also be written in the form of Eq.
-\ref{eq:PC}, using a
-different choice of ${\bf K}$ \citep{Hughes2013}, constructed as follows:
+This approach works by reformulating the spatial random effects in Eq.
+\ref{eq:spat_reg} as $\boldsymbol{\eta} = {\bf K} \boldsymbol{\alpha}$, where $\alpha$ is a vector of $m$ random effects in a reduced dimension subspace.
+The matrix ${\bf K}$ can be constructed as follows \citep{Hughes2013}:
 \begin{enumerate}
@@
 
-Although $\boldsymbol{\mu}$ could potentially be integrated out of the likelihood, its inclusion allows for easier computation of full conditionals for $\boldsymbol{\beta}$ and $\tau_\epsilon$ (and for other full conditionals once the model is expanded, e.g.
-for GAMs and STRMs).
+Although $\boldsymbol{\mu}$ could potentially be integrated out of the likelihood, its inclusion allows for easier computation of full conditionals for $\boldsymbol{\beta}$ and $\tau_\epsilon$ (and for other full conditionals once the model is expanded, for example, to GAMs and STRMs).
 During the main MCMC phase, the $\boldsymbol{\mu}$ only need to be updated for grid cells that are exposed to sampling (see a subsequent section for information on posterior prediction for details on unsampled areas).
@@
 
-In order to calculate prediction variance on the real scale for generalized models, it is necessary to use a procedure such as the delta method \citep{Dorfman1938}.
-In particular, we can write the variance of the Poisson expectation as
+In order to calculate prediction variance on the real scale for generalized models, one can use a procedure such as the delta method \citep{Dorfman1938}, or in the case of Bayesian inference, can simply calculate posterior predictive variance of $\boldsymbol{\lambda}_x$.
+For the delta method approach, we can write the variance of the Poisson expectation as
 \begin{linenomath*}
@@
 
-\hspace{.5in} Our approach in this paper is to conduct a Bayesian analysis (using the Gibbs sampling algorithms outlined above) and substitute the posterior distribution $[\boldsymbol{\theta}|{\bf Y}]$ for $[\boldsymbol{\theta}]$ in Eq.
-\ref{eq:E_var_lambda}.
-Under this approach, one simply sets ${\rm Var}(\mu_x)$ equal to the posterior variance of $\mu_x$ (across Markov chain iterations).
-In the case of the log link function, one also needs an estimate for $\mu_x$ to use in delta method calculations; we used the posterior median for each sampling unit.
+\hspace{.5in} Our approach in this paper is to conduct a Bayesian analysis (using the Gibbs sampling algorithms outlined above) and to use posterior predictive inference to summarize ${\rm var}(\boldsymbol{\lambda}_x | {\bf Y})$.
 Although we use a Bayesian formulation for modeling convenience, one could also consider substituting parametric approximations for ${\rm Var}(\mu_x)$ and $\mu_x$ based on output from standard likelihood-based analyses.

### 2015-04-16 manuscript/Text_S2.tex
Commit prior to PLoS One submission

--- 2015-01-20 manuscript/AppendixB.tex
@@
 \renewcommand{\em}{\it}
-\renewcommand\thefigure{B\arabic{figure}}
-\renewcommand\thetable{B\arabic{table}}
-\renewcommand\theequation{B.\arabic{equation}}
+\renewcommand\thefigure{S2.\arabic{figure}}
+\renewcommand\thetable{S2.\arabic{table}}
+\renewcommand\theequation{S2.\arabic{equation}}
 
 \begin{document}
-
-\begin{center} \bf {\large On extrapolating past the range of observed data when making statistical predictions in ecology}
-
-\vspace{0.7cm}
-Paul B. Conn$^{1*}$, Devin S. Johnson$^1$, and Peter L. Boveng$^1$
-\end{center}
-\vspace{0.5cm}
-
-\rm
-\small
-
-\it $^1$National Marine Mammal Laboratory, Alaska Fisheries Science Center,
-NOAA National Marine Fisheries Service,
-Seattle, Washington 98115 U.S.A.\\
 
@@
 \begin{center}
-Appendix B: Full details of simulation study examining predictive extrapolation
+S2 Text: Full details of simulation study examining predictive extrapolation
 \bigskip
@@
 \begin{enumerate}
-  \item Simulate three hypothetical, statistically dependent, spatially autocorrelated environmental covariates over a 30 by 30 grid,
+  \item Simulate three hypothetical, statistically dependent, spatially autocorrelated environmental covariates over a 30$\times$30 grid,
   \item Simulate animal abundance across the landscape as a function of environmental covariates,
@@
 For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models, corresponding to a generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a generalized additive model (GAM; smooth effects of covariates on the log scale), and a spatio-temporal regression model (with both linear, fixed effects and spatially autocorrelated random effects).
-Further details on these models, together with the procedure used for posterior simulation, are presented in Appendix A.
-For this study, we used the RSR implementation of the STRM outlined in Appendix A.
+Further details on these models, together with the procedure used for posterior simulation, are presented in Text S1.
+For this study, we used the RSR implementation of the STRM outlined in Text S1.
 Prior distributions for each precision parameter, $\tau$, were set to $\text{Gamma}(1.0,0.01)$, which is diffuse while maintaining a flat shape near the origin.
@@
 \underline{5. Calculating the gIVH} \\
+
+As suggested in Text S1, we used Eq. S1.7 to calculate the gIVH, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.
 

### 2015-04-16 SpatPred/inst/analyze_sim_data.R

@@ -138,7 +138,5 @@ Summary$Bias=-1.1
 bias.plot=bias.plot + geom_text(data=Means,aes(label=Bias.mean))
 bias.plot=bias.plot + geom_text(data=Summary,aes(label=N.above.2))
 
-pdf("bias_generic.pdf")
-bias.plot
-dev.off()
+ggsave(filename="bias_generic.tiff")
 

### 2015-04-16 SpatPred/inst/compile_plot_ribbon_data.R

@@ -64,7 +64,8 @@ new.colnames=colnames(tmp2)
 new.colnames[1:2]=c("Easting","Northing")
 colnames(tmp2)=new.colnames
 tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank(),plot.title = element_text(hjust = 0))
-pdf(file="covariates_col.pdf")
+
+tiff(file="covariates_col.tiff",height=8,width=8,units='in',res=300)
 pushViewport(viewport(layout=grid.layout(2,2)))
 p1=ggplot(tmp2)+aes(Easting,Northing,fill=dist_mainland)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=greenPalette(100),name="Value")+ggtitle("A. Distance from mainland")
 print(p1,vp=viewport(layout.pos.row=1,layout.pos.col=1))

### 2015-04-16 SpatPred/inst/plot_IVH_simple.R

@@ -87,7 +87,7 @@ XpXinv=solve(crossprod(X))
 for(i in 1:160000)Xnew.eval[i]=t(Xnew[i,])%*%XpXinv%*%Xnew[i,]
 Extrap=(Xnew.eval>max.IVH)
 
-my.theme=theme(plot.title = element_text(hjust = -.1),strip.text=element_text(face="bold"),axis.title=element_text(size=16),text=element_text(size=16),legend.position="none")
+my.theme=theme(plot.title = element_text(hjust = 0),strip.text=element_text(face="bold"),axis.title=element_text(size=16),text=element_text(size=16),legend.position="none")
 
 DF=data.frame(x=Xnew[,2],y=Xnew[,3],resp=Extrap)
 DF.data=data.frame(x=x,y=y)
@@ -133,7 +133,7 @@ XpXinv=solve(crossprod(X))
 for(i in 1:160000)Xnew.eval[i]=t(Xnew[i,])%*%XpXinv%*%Xnew[i,]
 Extrap=(Xnew.eval>max.IVH)
 
-my.theme=theme(plot.title = element_text(hjust = -.1),strip.text=element_text(face="bold"),axis.title=element_text(size=16),text=element_text(size=16),legend.position="none")
+my.theme=theme(plot.title = element_text(hjust = 0),strip.text=element_text(face="bold"),axis.title=element_text(size=16),text=element_text(size=16),legend.position="none")
 
 DF=data.frame(x=Xnew[,2],y=Xnew[,3],resp=Extrap)
 DF.data=data.frame(x=x,y=y)
@@ -146,7 +146,7 @@ pdf(file="IVH_simple2.pdf")
 grid.arrange(arrangeGrob(linear.plot,quad.plot,plot.biv,plot.biv.2,nrow=2,widths=unit.c(unit(.5,"npc"))))
 dev.off()
 
-tiff(file="IVH_simple2.tiff")
+tiff(file="IVH_simple2.tiff",res=300,width=8,height=8,units='in')
 grid.arrange(arrangeGrob(linear.plot,quad.plot,plot.biv,plot.biv.2,nrow=2,widths=unit.c(unit(.5,"npc"))))
 dev.off()
 

### 2015-04-16 SpatPred/inst/plot_sim_example.R

@@ -54,7 +54,7 @@ Est.clust=plot_N_map(1,N,Grid=Grid.list,leg.title="N",highlight=which(MCMC.clust
 df<-data.frame(id=ids,Easting=XY[Effort.clust$Mapping,"x"],Northing=XY[Effort.clust$Mapping,"y"],Abundance=1)
 Est.clust=Est.clust+geom_point(data=df,aes(x=Easting,y=Northing),shape=3,size=2)+ggtitle("F. Convenience")+theme(plot.title = element_text(hjust = 0,size = rel(4)),legend.text=element_text(size=rel(3)),legend.title=element_text(size=rel(3)),text=element_text(size=rel(3)))+xlab('')+ylab('') #+theme(plot.title = element_text(hjust = 0,size = rel(1.5)),legend.position="none",axis.ticks = element_blank(), axis.text = element_blank())
 
-pdf(file="sim_maps.pdf")
+tiff(file="sim_maps.tiff",res=300,height=8,width=8,units='in')
 grid.arrange(arrangeGrob(Cov1,Cov2,Cov3,N.plot,Est.sys,Est.clust,widths=unit(0.5,"npc"),heights=unit(0.33,"npc"),nrow=3))
 dev.off()
 

### 2015-04-16 SpatPred/inst/run_ribbon_models.R

@@ -219,7 +219,7 @@ rsr.ice0.plot=rsr.ice0.plot+ggtitle("F. STRM (Pseudo-absence)")+scale_fill_gradi
 
 
 library(gridExtra)
-pdf(file="ribbon_maps.pdf")
+tiff(file="ribbon_maps.tiff",res=300,height=8,width=8,units='in')
 grid.arrange(arrangeGrob(naive.glm.plot,ice0.glm.plot,naive.gam.plot,ice0.gam.plot,naive.rsr.plot,rsr.ice0.plot,nrow=3,widths=unit(0.5,"npc"),heights=unit(0.33,"npc")))
 dev.off()
 

### 2015-07-30 manuscript/Revision/ReviewerResponses.tex
Commit prior to revision of PLoS One ms. Switched Bayesian GAM out for frequentist one using mgcv, changed color scheme for plot_N_map

%  This document must be used with LaTeX 2e!!!

%  See the documentation ``The biometrics class'' by
%  Josephine Collis for more information

%  Be sure to put the files biometrics.sty and biometrics.bst
%  in the same directory as your latex file.  Also put the .bib
%  file -- see below

\documentclass[12pt,fleqn]{article}
%\documentclass[12pt,a4paper]{article}
\usepackage{biometrics}
\usepackage{amsmath,epsfig,epsf,psfrag,color}
%\usepackage{a4wide,amsmath,epsfig,epsf,psfrag}
\usepackage[nolists]{endfloat}

%  put your commands here

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bthet{{\ensuremath\boldsymbol{\theta}}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

\begin{document}

\bf \hspace{-.32in} Conn et al. responses to reviewer comments

%  make sure that the document has 25 lines per page (it is 12 pt)

\setlength{\textheight}{575pt} \setlength{\baselineskip}{23pt}

%  IMPORTANT -- place the \title, \author, etc, \begin{summary}...
%  \end{summary} \keywords statements in the order demonstrated below
\rm
Here, we provide line by line responses to reviewer comments, suggestions, and edits.  For completeness, we include all reviewer comments in addition to our responses (our responses are in blue).

\section{General}
The authors present a potentially extremely useful extension of a classical model checking technique. I would like to
congratulate the authors on a succinct and clear paper. On the whole I think the paper is publishable aside from a couple of
clarifications. However, I would like to make the suggestions below to (hopefully) make the paper more easily understandable
to a wider audience and to put it into a broader context.

I would also like to thank the authors for putting their data and code online during the review process, this made it very easy
to understand their methods.

I have included comments below and direct comments as annotations on the PDF of the paper and appendices. Please note
that the annotations are not simply typos, there are major points there too.

\textcolor{blue}{
We thank the reviewer for the effort they put forth to review our submission and for their constructive criticism (particularly with regard to appealing to a larger audience by including frequentist GAMs).  We really appreciate the thought that went into this - hopefully it will make it into a much better paper.
}
\textcolor{blue}{
We include point by point responses to comments and annotations made on the PDFs after responding to major comments below.
}

\section{Context}
The authors may be interested in a recent paper by Mannocci et al (2015) where "envelope models" (i.e. max/min of
covariates used as cutoff in predictions) are developed (disclosure: I reviewed that paper). It may be interesting to compare
the differences in the resulting abundance estimates given the envelope approach. I had suggested in my review of Mannocci's
paper that they think about quantiles rather than max/min but this proved to exclude too many observations in their examples.

\textcolor{blue}{
We had not seen this paper and thank the reviewer for bringing it to our attention.  In the Discussion (lines 330-340), we now cite Mannocci et al (2015) and try to situate the gIVH as an alternative method for envelope specification.  We now state that
}

\textcolor{blue}{
``In the field of SDM modeling, researchers often stress the need for prediction locations to be similar to the locations used for model development~\cite{ElithLeathwick2009}.  One way to accomplish this is through a prediction envelope, whereby a specific criterion is used to limit predictions of animal density or occurrence to the range of conditions and covariates encountered during surveys~\cite{MannocciEtAl2015}.  Using the gIVH for this purpose will likely be more conservative than envelope specifications based on other criterion (e.g., in contrast to minimum an maximum observed covariate values as in~\cite{MannocciEtAl2015}), but is more in line with linear modeling theory.  A comparison of envelope specification methods is beyond the scope of this paper, but we suspect there are cases where specific strategies that seem intuitive can result in poor predictions, particularly when the form of prediction models is of high dimension and/or includes multiple interaction terms."
}

\textcolor{blue}{
As we indicate in the text, we believe a comparison of alternative envelope specifications, while of general interest, is beyond the scope of the paper. In particular, we suspect that relative performance of different envelopes will likely depend on a large number of factors (including, but not limited to, the complexity of the estimation model).
}

\section{GLM/GAM Theory}

In Wood (2006) p. 245 and elsewhere, {\bf x} is referred to as ``${\bf X}_P$" or the ``lpmatrix ". That is, the matrix that maps the
parameters to the predictions on the link scale. The authors may want to incorporate this terminology and explanation into the
article.
In a frequentist GAM, one can then apply the delta method (as the authors have done) to obtain the prediction variance on the
response scale or use a parametric bootstrap (see e.g. Marra, Miller and Zanin 2011, Augustin et al 2013 or Miller et al 2013,
Appendix B). See below for more on this.

\textcolor{blue}{
We have maintained our original notation in the main article, but now specifically indicate in S1 Text that \cite{Wood2006} terms {\bf x} the ``lpmatrix" and provides methods to output it in the \texttt{mgcv} package.  In the main article text, we also indicate the possibility of using parametric bootstrapping as an alternative to the delta method and cite the articles the reviewer suggested (lines 147-151).  Further information, including new details regarding frequentist GAMs, is provided in S1 Text.
}

\section{Other models}

{\it Flexible frequentist models}. Examples in the paper are all in a Bayesian setting. I'm not about to argue that there is any
problem with this at all BUT it might draw more people to the paper if there were a frequentist example too. Since the authors
were so kind as to put their data and code online, I took the liberty of fitting a GAM to their data and calculating the gIVH for a
frequentist GAM (using the R package mgcv ). Though I am sure that I have made some slight mistakes in the precise format
of the data, the predicted abundance I obtained using the gIVH was about 47,827, down from 301,557 without using gIVH.
This estimate seems on the low side compared to the posterior prediction quantiles reported (see next comment for more on
this). I have included the code to do this with the review, in the hope this will be useful (with apologies it is not sufficiently well
commented).

\textcolor{blue}{
We agree that an exploration of applying gIVH ideas to frequentist GAMs would be useful for drawing in a wider audience.  We went ahead and replaced our Bayesian GAMs (which didn't really work very well anyway) with results from applying frequentist GAMs via \texttt{mgcv}.  We did this both with regards to simulations and the ribbon seal example and include details and implementation instructions in S1 Text.  We have reused portions of the reviewer's R code for this purpose, and give him credit in both the relevant R script as well as the acknowledgments section of the main article (we hope that's okay - thanks for providing it!).
}

\textcolor{blue}{
There is undoubtedly some substantial model-to-model uncertainty here, likely because the epicenter of ribbon seal abundance was poorly sampled. Some of the discrepancy that the reviewer notes between predicted abundance from the gIVH-restricted GAM and the Bayesian estimation approaches likely has to do with a considerably larger number of locations that fall out of the gIVH in the GAM model.  A better comparison would be to restrict comparison of total abundance for cells that fall within the gIVH for {\it all} fitted models so that the total number of cells contributing to the abundance estimate is the same for each model.  We now make this comparison when describing results from the ribbon seal example.  Although the GAM produced an estimate that bordered on ``significantly" higher than the GLM model when estimating total abundance for the entire study area, estimates are actually remarkably similar when restricted to this multi-model gIVH envelope.
}

Models with large number of parameters. Cook (1979) says "since the v are increasing functions of p, outliers become more
difficult to detect as the model is enlarged" (p. 172). It would be interesting to know what happens when we have complex
models (i.e. some GAMs I fit have 100s if not 1000s of parameters) and see how well the gIVH performs. This could be done
simply by increasing the spline basis size. My observation was that spline basis size made quite a big difference to the result;
increasing the basis size of my GAM from 5 to 9 reduced the abundance estimate considerably. This probably bears more
thought than I have given it so far.

\textcolor{blue}{
We're not sure what can be said theoretically about the relationship of the gIVH and model complexity in the generalized case.  For example, playing around with basis size for a quasipoisson GAM in the ribbon seal example, we obtained the following:}

\vspace{0.3in}
\begin{tabular}{lll}
\hline
Knots & gIVH size (no. cells) & $\hat{N}$ - all cells \\
\hline
3 & 30 & 92000 \\
4 & 47 & 118000 \\
5 & 56 & 121000 \\
7 & 61 & 288000 \\
9 & 46 & 219000 \\
\hline
\end{tabular}
\vspace{0.3in}

\textcolor{blue}{
 This result conformed to our a priori expectation that increasing the basis size increases the number of effective parameters and thereby (a) increases the chances that a given sampling unit will be out of the gIVH (after all, with a ``dot" model, all cells are within the gIVH by definition), and (b) will tend to increase abundance since there is a greater potential for extrapolation past the range of observed data (that, and effectively increasing the variance on a right skewed distribution).  However, it's not quite so clear cut, as the $s=9$ point shows.
}

\textcolor{blue}{
 Given that it may be difficult to say anything generally about model complexity and the gIVH, we have not attempted to address this issue directly.  We do however note that, with regards to future work on ribbon seals, that ``Although not reported here, additional model fitted to the data suggested sensitivity to model complexity and choice of basis, both of which are worthy of additional investigation."  (end of Ribbon Seal results section, lines 304-306).
}

\section{Plots}

As commented in the paper, I would advise against the use of the "rainbow" colour scheme. For folks like myself who are
colourblind, these are a stuggle. Even for those who are not disabled, they are problematic. See e.g. Borland and Taylor
(2007).

\textcolor{blue}{
We have change our color schemes from rainbow to monochromatic ones available in RColorBrewer.  Hopefully this will make plots more discernable to colorblind folks and will help with viewing if printed in black and white.
}

\section{Annotated PDF comments}

\subsection{Article}

Abstract.  May want to say via what method here?

\textcolor{blue}{We now state (last sentence in abstract, lines 17-20) that ``We suggest that ecologists routinely use diagnostics such as the gIVH to help gauge the reliability of predictions from statistical models (such as generalized linear, generalized additive, and spatio-temporal regression models)."}

Line 60.  May want to be more specific about the kind of model(s) here.

\textcolor{blue}{We have added the following sentence to the end of the introduction (lines 62-64): ``In particular, we examine the performance of the gIVH in identifying problematic extrapolations when modeling survey counts using GAMs, GLMs, and STRMs."}

Line 71.  ``In both geographical and covariate space"
\textcolor{blue}{parenthetical added}

Clarify: you mean $v=max(diag(V_{LR}))$? 
\textcolor{blue}{definition added}

Lines 104-106.  Do you mean response distributions?
\textcolor{blue}{Our understanding is that ``error distribution" and ``response distribution" are somewhat interchangeable, but we prefer the authors suggestion - changed to ``response distribution."}

Lines 106-107.  citation/expand on this?
\textcolor{blue}{This was poorly stated.  We've replaced ``for models with more general spatial structure" with "for more complicated models with prior distributions on parameters, as with hierarchical models."}

Lines 114-115  “observed design points” is kind of odd terminology: design points in a linear model may be observed but e.g. a GAM or spline or quadratic model the basis evaluations are not really "observed” in the same way.

I understand why you have this observed/unobserved given the models you talk about next but it would be better IMHO to use more precise terminology here to keep it open to a more general audience?

\textcolor{blue}{This is a good point.  We chose this terminology for generality (i.e. one need not necessarily be talking about spatially referenced data), but given the motivating topic of SDM modeling, this generality probably isn't needed.  Changed to ``the set of locations where data are observed." (lines 116-117)}

Line 129. See $L_p$ matrix stuff in Wood (2006), see long comments.

\textcolor{blue}{
We have maintained our original notation in the main article, but now specifically indicate in S1 Text that \cite{Wood2006} terms {\bf x} the ``lpmatrix" and provides methods to output it in the \texttt{mgcv} package.
}

Line 132. Or simulation based methods? See long comments.

\textcolor{blue}{
We now indicate the possibility of using parametric bootstrapping as an alternative to the delta method in the main text, citing the articles the reviewer suggested (lines 147-151).  Further details, including new details regarding frequentist GAMs, are provided in S1 Text.
}

Line 138.  Expand on ``. . . will often involve unknown parameters. . . "  I'm not sure what you mean.

\textcolor{blue}{
If trying to use the gIVH for survey planning, one potentially won't have any data, just a proposed model structure.  In this case, one would need to integrate across prior parameter distributions to produce the gIVH.  That's what we were (admittedly obtusely) trying to get at.  However, I think this is implicit in posterior prediction and we now omit this description.
}

Line 169.  1.0 not 100\%
\textcolor{blue}{
changed}

Line 224.  Picky question: was there sensitivity to knot choice? I know this is not the point of the paper, but knot selection is hard.
\textcolor{blue}{
Yes, there was sensitivity to knot choice, especially with regards to endpoint behavior.  We weren't really happy with how this model turned out, and have eliminated it from the study (replacing it with frequentist GAMs fitted with \texttt{mgcv}.}

Line 248.  [RE: pseudo-absence stuff] This seems a little ad hoc, do you have some references to back this up? Can you show the sensitivity in the results to this and what it’s going to the gIVH.
\textcolor{blue}{We agree that this is a little ad hoc but feel that this approach results in much better estimates. The main issue was that models without these pseudo-absences (which we've renamed ``presumed absences" in the revision to differentiate it from ``pseudo-absences" as used in the presence-only SDM literature) is that there were no surveys conducted over open water so there are few data to inform the y-intercept of the ice concentration - seal count function. Although this tactic is quite specific to our study population, we believe it is a reasonable one.  We do show the resulting estimates and gIVHs with and without ``presumed absences" in Fig. 6.
}

Line 265. Was this down to knot placement or number issues?
\textcolor{blue}{Again, we've removed our kludgy Bayesian GAM model in favor of \texttt{mgcv}.
}

Line 276. Commas separating thousands and spaces after commas separating numbers! Otherwise this is very hard to read.
\textcolor{blue}{Changed as as suggested.
}

Figure 2 Caption.  Hard to tell from the low res graphics but the cells outside the gIVH are all overestimates?  Might be easier to plot the bottom two as error surfaces i.e. prediction-truth?  Are the latter two plots an average of just one realisation from the simulation?
 \textcolor{blue}{It's unfortunate that the PLoS changes figures to low res for the review - although the rainbow coloration probably doesn't help given the reviewer's color blindness.  We have changed the color schemes in this figure to monochromatic ones from RColorBrewer.  This should show that many of the convenience sample estimates in the gIVH are overestimates (but that the spatially balanced estimates look okay for this particular realization).  This plot includes estimates from a single simulation replicate for illustration - we have changed the plot title from ``Depiction of a single simulation scenario" to ``Depiction of a single simulation replicate" to make this more explicit. We considered changing the bottom two plots to bias, but decided that we preferred absolute abundance here.
}

Fig. 4. Should the brown and orange lines be the other way around?
Add someone who is colour-blind it would be better to use more distinct and colour-blind friendly colours.
 \textcolor{blue}{Yes, we had this backwards and have fixed this in the figure caption.  We aren't sure what other colors would be better to use here... however, we have made the orange depth contour line into a dashed line so that all readers should be able to differentiate the two.
}

\subsection{S1 Text}

Page 1.  this is defined backwards, it would be better to say what the linear predictor is before defining mu in terms of it

\textcolor{blue}{
Our preference is to provide equations and then define the elements in them, rather than the other way around.  This seemed like a relatively standard way of doing things, so we have left the text as is.
}

End of section 1.2.  You mean that you can use different basis functions? 
\textcolor{blue}{
Yes, we were getting at different basis functions.  This section has been substantially reformatted, but we now describe ${\bf K}$ as a matrix specified using smooth basis functions.
}

Section 1.3.  Remove “Parameter estimates from”? 
\textcolor{blue}{
Removed.
}

Section 2.  \textcolor{blue}{
Missing citation added
}

Section 2.5.  I think this should be in the paper rather than appendix.
\textcolor{blue}{We have incorporated information from this section into the main body of the paper (see lines 128-151 in main article text). }

Section 2.5.  As in main article, this [regarding unknown parameters when calculating the gIVH] requires further explanation.
\textcolor{blue}{The gIVH is analytically defined for linear regression, but not necessarily for STRMs which have prior distributions, random effects, etc.  We have made an attempt to flesh this out in the main article text.}

\subsection{S2 Text}

Page 1. delete `animal population' in bullet 3
\textcolor{blue}{Done}

Page 4, on simulating count data. Is this realistic given usual aerial survey procedure? Would it not be more realistic to generate lines and find cells underneath rather than randomly generating cells?
\textcolor{blue}{We think this a realistic test of a plot-based sampling, which seems like an adequate simulation test for the gIVH.  Although we used aerial transect surveys in the ribbon seal example, the ideas behind the gIVH pertain to multiple types of sampling, including plot-based counts, transect surveys, etc.}

\bibliographystyle{biometrics}
\bibliography{master_bib}

\end{document}

### 2015-07-30 manuscript/Revision/SpatPred_PLoS_revision.tex
Commit prior to revision of PLoS One ms. Switched Bayesian GAM out for frequentist one using mgcv, changed color scheme for plot_N_map

--- 2015-04-16 manuscript/SpatPred_PLoS.tex
@@
 We also demonstrate the utility of the gIVH in diagnosing problematic extrapolations when estimating the relative abundance of ribbon seals in the Bering Sea as a function of predictive covariates.
-We suggest that ecologists routinely use diagnostics such as the gIVH and related metrics (e.g., prediction variance) to help gauge the reliability of predictions from statistical models.
+We suggest that ecologists routinely use diagnostics such as the gIVH to help gauge the reliability of predictions from statistical models (such as generalized linear, generalized additive, and spatio-temporal regression models).
 % Please keep the Author Summary between 150 and 200 words
@@
 Application of the gIVH and related criterion (e.g., prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction and survey design.
-We illustrate use of the gIVH on simulated count data, and on a species distribution model (SDM) for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea.
+We illustrate use of the gIVH on simulated count data, and on several species distribution model (SDM) formulations for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea.
+In particular, we examine the performance of the gIVH in identifying problematic extrapolations when modeling survey counts using GAMs, GLMs, and STRMs.
 % You may title this section "Methods" or "Models".
@@
 \section*{Materials and Methods}
+
+All data collected and research activities described in this manuscript were performed under National Marine Fisheries Service research permit number 15126.
+
 \subsection*{Generalizing the independent variable hull}
@@
 Predictions outside the range of observed covariates?
-Predictions for locations that are so far from places where data are gathered that we are skeptical that the estimated statistical relationship still holds?
+Predictions for locations that are so far (in either geographical or covariate space) from places where data are gathered that we are skeptical that the estimated statistical relationship still holds?
 To help guide our choice of an operational definition, we turn to early work on outlier detection in simple linear regression analysis.
@@
 Under this formulation, the IVH is defined relative to the hat matrix, ${\bf V}_{\rm LR}={\bf X}({\bf X}^\prime{\bf X})^{-1}{\bf X}^\prime$ (where the subscript ``LR" denotes linear regression).
-Letting $v$ denote the maximum diagonal element of ${\bf V}_{LR}$, one can examine whether a new design point, ${\bf x}_0$ is within the IVH.
+Letting $v$ denote the maximum diagonal element of ${\bf V}_{LR}$ (i.e.
+$v = \max(\textrm{diag}({\bf V}_{LR}))$), one can examine whether a new design point, ${\bf x}_0$ is within the IVH.
 In particular, ${\bf x}_0$ is within the IVH whenever
@@
 Cook's~\cite{Cook1979} formulation for the IVH is particular to linear regression analysis, which assumes independent and identically distributed (\textit{iid}) Gaussian error.
-Thus, it is not directly applicable to generalized models, such as those including alternative error distributions (e.g., Poisson, binomial) or spatial random effects.
-Further, the hat matrix is not necessarily well defined for models with more general spatial structure.
+Thus, it is not directly applicable to generalized models, such as those including alternative response distributions (e.g., Poisson, binomial) or spatial random effects.
+Further, the hat matrix is not necessarily well defined for more complicated models with prior distributions on parameters, as with hierarchical models.
 However, since the hat matrix is proportional to prediction variance, Cook~\cite{Cook1979} notes that design points with maximum prediction variance will be located on the boundary of the IVH.
-We therefore define a generalized independent variable hull (gIVH) as the set of all potential design points $\mathcal{S}_0$ (note that $\mathcal{S}_0$ can include both observed and unobserved design points) such that
+We therefore define a generalized independent variable hull (gIVH) as the set of all predicted locations $\mathcal{S}_0$ for which
 \begin{linenomath*}
@@
 \end{linenomath*}
-where $i \in \mathcal{S}_0$, $\lambda_i$ corresponds to the mean prediction at \textit{i}, $\mathcal{S}$ denotes the set of observed design points, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.
+where $i \in \mathcal{S}_0$, $\lambda_i$ corresponds to the mean prediction at \textit{i}, $\mathcal{S}$ denotes the set of locations where data are observed, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.
 
@@
 \begin{equation}
-  \boldsymbol{\mu} = {\bf X}_{aug}\boldsymbol{\beta}_{aug} + \boldsymbol{\epsilon},
+  \boldsymbol{\mu} = {\bf X}_{aug}\boldsymbol{\beta}_{aug},
   \label{eq:mu-aug}
@@
 \end{linenomath*}
-where the $\boldsymbol{\epsilon}$ represent Gaussian error, ${\bf X}_{aug}$ denotes an augmented design matrix, and $\boldsymbol{\beta}_{aug}$ denote an augmented vector of parameters.
+where ${\bf X}_{aug}$ denotes an augmented design matrix, and $\boldsymbol{\beta}_{aug}$ denote an augmented vector of parameters.
 For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see \nameref{S1_Text} for examples of how numerous types of models can be written in this form).
@@
   \label{eq:var-mu}
-  {\rm var}(\boldsymbol{\mu}_x)= {\bf x} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) {\bf x}^\prime,
+  {\rm var}(\boldsymbol{\mu})= {\bf X}_{aug} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) {\bf X}_{aug}^\prime,
 \end{equation}
 \end{linenomath*}
-where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; \nameref{S1_Text}).
-This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method~\cite{Dorfman1938,VerHoef2012b} to approximate prediction variance on the real scale (i.e., the scale in which data are measured; see \nameref{S1_Text}).
-If a Bayesian approach to estimation is used, computation is even easier, as one can simply base thresholding on posterior predictive variance of $g^{-1}(\mu_i)$.
-A Bayesian mode of inference is perhaps the most natural to use in these problems, for the simple reason that unlike Cook's original formulation, the gIVH will often involve unknown parameters.
-For most of the following treatment, we shall assume that data have already been collected (see Discussion for comments on the potential use of the gIVH in survey planning), and that informative prior probability distributions can be specified for unknown parameters.
+where it is understood that the exact form of ${\bf X}_{aug}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; \nameref{S1_Text}).
+Alternative model structures for $\boldsymbol{\mu}$ can also be accommodated; for instance, in Bayesian models ${\rm var}(\boldsymbol{\mu})$ can be set equal to posterior predictive variance, $\textrm{var}(\boldsymbol{\mu}|{\bf Y},\boldsymbol{\theta})$ (where $\boldsymbol{\theta}$ represent hyperparameters).
+
+The expression for prediction variance in Eqn.
+\ref{eq:var-mu} is on the linear predictor scale.
+If a non-identity link function is used, an additional step is needed to convert prediction variance to the response scale (i.e.
+to calculate $\textrm{var}(\boldsymbol{\lambda})$ as needed to define the gIVH in Eq.
+\ref{eq:gIVH}).
+One approach for calculating variance on the response scale is simply to use the delta method~\cite{Dorfman1938,VerHoef2012b}.
+In particular, we can write the variance of the expected responses as
+\begin{linenomath*}
+\begin{eqnarray}
+  \label{eq:var-lambda}
+  {\rm var}(\boldsymbol{\lambda}) & = & {\rm var}(g(\boldsymbol{\mu})) \nonumber \\
+  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu})\boldsymbol{\Delta}^\prime,
+\end{eqnarray}
+\end{linenomath*}
+where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).
+Under common univariate link functions (e.g.
+log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g.
+multinomial logit) $\boldsymbol{\Delta}$ will be dense.
+
+Alternatively, one can use a simulation based method for determining variance of the predictive mean response vector.
+In Bayesian analysis of hierarchical models, this is easily accomplished via posterior predictive inference~\cite{Gelman2004}.
+In a similar spirit, it is also possible to use parametric bootstrapping instead of the delta method to approximate prediction variance on the response scale for frequentist models~\cite{MarraEtAl2012,AugustinEtAl2013,MillerEtAl2013}.
 
 We propose to use the gIVH in much the same manner as Cook~\cite{1979}.
-In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).
-The gIVH, together with posterior prediction variance, seem ideally suited to diagnosing potential extrapolation issues.
-Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions will be based on.
+In particular, we use the gIVH to determine whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).
+For most of the following treatment, we shall assume that data have already been collected (see Discussion for comments on the potential use of the gIVH in survey planning).
+For further details on how the gIVH was calculated for specific models in this paper, see \nameref{S1_Text}.
 
@@
 We developed a package \texttt{SpatPred} in the R statistical programming environment~\cite{RTeam2012} to simulate data and conduct all analyses.
-The seal dataset is included as part of this package, and is available at \url{https://github.com/pconn/SpatPred}.
-The R package has also been archived via figshare [WE PLAN TO DO THIS WHEN THE MANUSCRIPT IS ACCEPTED FOR PUBLICATION].
+The seal dataset is included as part of this package, and is available at \url{https://github.com/pconn/SpatPred/releases}.
+The R package has also been archived via figshare~\cite{Conn2015}.
 
@@
 We configured virtual sampling quadrats such that they encompassed 10\% of the area of each selected grid cell.
-For ease of presentation and analysis, we assumed detection probability was 100\% in each quadrat.
+For ease of presentation and analysis, we assumed detection probability was 1.0 in each quadrat.
 Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (\nameref{S1_Text}).
@@
 issue with spatial prediction: abundance of ribbon seals appears to be maximized in the southern and/or southeast quadrant of the surveyed area.
-Predicting abundance in areas farther south and east may thus prove problematic, as the values of several explanatory covariates (Fig.
+Predicting abundance in areas farther south and west may thus prove problematic, as the values of several explanatory covariates (Fig.
 \ref{fig:covs}) are also maximized in these regions.
 
-We start by fitting hierarchical GLMs, STRMs, and GAMs to the ribbon seal data.
+We start by fitting hierarchical GLMs and STRMs to the ribbon seal data.
 To accommodate incomplete coverage of grid cells and account for non-target habitat, we adapted Eqs.
@@
 where $P_i$ gives the proportion of area surveyed in grid cell $i$, $A_i$ gives the proportion of cell $i$ that is seal habitat, $\theta_i$ is a linear predictor, and $\epsilon_i$ is normally distributed $iid$ error.
-By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs, GAMs, and STRMs (see \nameref{S1_Text})
+By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs and STRMs (see \nameref{S1_Text}).
 
 The fixed effects component of the GLM and STRM included linear effects of all explanatory covariates (Fig.
-\ref{fig:covs}, as well as a quadratic effect for sea ice concentration.
-For the STRM, we imposed a restricted spatial regression (RSR) formulation for spatially autocorrelated random effects, where dimension reduction was accomplished by only selecting eigenvectors of the spectral decomposition associated with eigenvalues that were greater than 0.5 (see Text S1 for additional information on model structure).
-For the GAM, we employed smooth terms with 8 knots for each covariate.
-This included 4 knots at the minimum, 1/3 and 2/3 quantiles, and the maximum observed covariate values, in addition to 4 knots above and below the range of observed data to reduce the magnitude of known biased endpoint behavior that can occur with radial basis function implementations of GAMs.
-
-As with simulated data examples, we adopted a Bayesian perspective and conducted estimation using MCMC (see Text S1 for algorithm details and information on prior distributions) with 60,000 iterations where the first 10,000 iterations were discarded as a burn-in.
+\ref{fig:covs}), as well as a quadratic effect for sea ice concentration.
+For the STRM, we imposed a restricted spatial regression (RSR) formulation for spatially autocorrelated random effects, where dimension reduction was accomplished by only selecting eigenvectors of the spectral decomposition associated with eigenvalues that were greater than 0.5 (see \nameref{S1_Text} for additional information on model structure).
+Adopting a Bayesian perspective, we estimated parameters for these models using MCMC (see \nameref{S1_Text} for algorithm details and information on prior distributions) with 60,000 iterations where the first 10,000 iterations were discarded as a burn-in.
 We generated posterior predictions of ribbon seal abundance across the landscape as
@@
 \end{equation}
-and calculated the gIVH as in Eqn. \ref{eq:gIVH}, with delta method modifications as specified in \nameref{S1_Text}.
+and calculated the gIVH as in Eqn. \ref{eq:gIVH}, with delta method modifications as specified in Eqn. \ref{eq:var-lambda}.
 
-Initial spatial predictions using the three models (GLM, GAM, STRM) all produced extremely high, unbelievable predictions along the southern boundary of the study area (Fig.
+We also fitted a frequentist GAM to seal data using the \texttt{mgcv} R package~\cite{Wood2006}.
+We included smooth terms for all explanatory covariates; however, owing to relative data sparsity, we chose the minimum number of knots allowable ($s=3$) for the default thin plate spline smoother.
+We used a quasipoisson error structure in \texttt{mgcv} for this analysis, which was the most similar option available to the log-Gaussian Cox formulation chosen for the GLM and STRM models.
+For more information on the procedure used to generate parameter estimates and abundance predictions on the response scale, see \nameref{S1_Text}.
+
+Initial spatial predictions using two of the three models (GLM, STRM) produced extremely high, unbelievable predictions along the southern boundary of the study area (Fig.
 \ref{fig:ribbon_plot}).
@@
 However, this is not ideal in that one does not get an abundance estimate for the whole study area, and it may be difficult to compare abundance from one year to the next using this approach.
-Second, one could consider a simpler estimation model, such as one post-stratifying by ecoregion (as in~\cite{ConnEtAl2014}).
+Second, one could try different predictive covariate models (e.g., by altering the combination or polynomial degree of covariates included in the model).
 Finally, one could build in a priori knowledge of habitat preferences into the model structure.
-We adopted the latter solution, incorporating ``pseudo-absences" (i.e., zero counts where sampling was not conducted) in locations where it would have been (nearly) impossible to detect seals.
-Specifically, we inserted pseudo-absences in cells where ice concentrations were \textless 0.1\%.
+We adopted the latter solution, incorporating presumed absences (i.e., zero counts where sampling was not conducted) in locations where it would have been (nearly) impossible to detect seals.
+Specifically, we inserted presumed absences in cells where ice concentrations were \textless 0.1\%.
 This solution seemed the most logical, as many of the large, anomalous predictions were over open water along the southern edge of the study area, where we would have obtained zero counts had they been surveyed.
@@
 \ref{fig:sim_boxplots}).
-This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.
-The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.
-For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude ($\approx 0.3$) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
-By contrast, proportional bias was close to zero when inference was restricted to the gIVH.
-Interestingly, the radial basis GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.
+Although median bias was close to zero, this right skew translated into positive mean bias, and was exacerbated when convenience sampling was employed.
+The magnitude of mean absolute bias was either the same or reduced (often substantially so) when inference was constrained to the gIVH.
+Positive proportional bias was the rule, and was of concerning magnitude ($\approx 0.3$) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.
+By contrast, proportional bias was close to zero when inference was restricted to the gIVH, although there appeared to be a small negative bias (Fig.
+\ref{fig:sim_boxplots}).
+Interestingly, bias for frequentist GAMs was of smaller magnitude that the Bayesian GLM or STRM models for the particular model structures used here.
 
@@
 
-Fitting our three ribbon seal SDMs to the augmented dataset with pseudo-absences, most predictions occurred within the gIVH (Fig.
+Fitting our three ribbon seal SDMs to the augmented dataset with presumed absences, most predictions occurred within the gIVH (Fig.
 \ref{fig:ribbon_plot}).
-The largest difference among the three models was in the southwest corner of the study area, coinciding with the few cells that were outside the gIVH in the three models.
-Posterior summaries of abundance across the study area were very comparable (albeit with a longer right hand tail for the STRM), with $5\%$, $50\%$, and $95\%$ posterior prediction quantiles as follows: GLM (48686,64836,93927); GAM(46120,61214,90873); STRM(41039,63717,194095).
+Posterior summaries of abundance across the entire study area were of similar magnitude, with $5\%$, $50\%$, and $95\%$ posterior prediction quantiles as follows: GLM (48,686, 64,836, 93,927); STRM(41,039, 63,717, 194,095).
+The GAM produced an estimate of 92,277 (SE 16,549).
+The largest differences among the three models was in the southwest corner of the study area in the area where predictions often occurred outside the gIVH.
+Restricting comparison of abundance to those cells that occur within the gIVH in all three models (i.e.
+only cells not highlighted in the right column of Fig.
+\ref{fig:ribbon_plot}), posterior prediction quantiles for the GLM were (41,750, 52,863, 69,557) and for the STRM were (39,446, 56,520, 135,427); estimated GAM abundance from \texttt{mgcv} was 59,104 (SE 5,376).
+There thus appears to be substantial between-model variation in predicted abundance when summed over the entire study area, but much better agreement (albeit with a heavier right tail for the STRM) when restricting inference to locations where predictions occur within the gIVH.
+
 We note that these estimates are for example illustration only, as they are uncorrected for imperfect detection (e.g., incomplete detection of thermal cameras, animals that were unavailable for sampling because they were in the water, species misidentification;~\cite{ConnEtAl2014}).
 Our approach here was to examine extrapolation and prediction error using relatively simple models, with the understanding that such effects are also likely to occur in complex models with more realistic observation processes.
+Standard diagnostics (e.g.
+q-q plots in \texttt{mgcv}) also suggested some lack of fit associated with the quasipoisson error distribution; future work should investigate alternate error structures such as the Tweedie distribution~\cite{MillerEtAl2013}.
+Although not reported here, additional model fitting suggested sensitivity to model complexity and choice of basis, both of which are worthy of additional investigation.
 
@@
 
-We have demonstrated the capacity for certain classes of statistical models to produce biased predictions of animal abundance when extrapolating past the range of observed data.
-In simulations, commonly used models (GLMs, STRMs) exhibited positive bias when predictions were required for the entire study area, particularly when convenience sampling was employed (with mean proportional bias $\approx 30\%$).
-In the ribbon seal example, naive extrapolation of fitted statistical relationships produced high positive bias along the southern boundary of the study area.
+We have demonstrated the capacity of certain classes of statistical models to produce biased predictions of animal abundance when extrapolating past the range of observed data.
+In simulations, commonly used models exhibited substantial mean positive bias when predictions were required for the entire study area, particularly when convenience sampling was employed.
+Median bias in the simulation study was close to zero, but the bias distribution was right skewed, indicating the possibility of considerably biased overestimates in a substantial proportion of simulation replicates.
+By contrast, restricting inference to locations within the gIVH led to small negative bias.
+Although this negative bias is undesirable, it may be preferable from a conservation and management standpoint.
+For instance, making management decisions (e.g., harvest, restoration efforts) based on estimates that have a small negative bias are much less likely to lead to catastrophic population collapse than are decisions based on overestimates.
+
+In the ribbon seal example, naive extrapolation of fitted statistical relationships produced high positive bias along the southern boundary of the study area for the GAM and STRM models.
 However, the gIVH appeared useful in diagnosing places where extrapolations from the fitted statistical model were problematic.
-For instance, in the simulation study, limiting the scope of inference to locations falling within the gIVH reduced or eliminated bias.
-For ribbon seal relative abundance, it was useful for confirming that the naive models needed to be reformulated, and that reformulated models (with pseudo-absence data) produced predictions in unsurveyed areas that were largely within the realm of observed data.
+For ribbon seal relative abundance, it was useful for confirming that the naive models needed to be reformulated.
+Reformulated models (with presumed absence data) still yielded estimates of total abundance with considerable between-model variation in the southwest corner of the study area.
+However, when inference was restricted to locations within the gIVH for all three fitted models, abundance estimates were quite comparable.
 
-Our intent is to raise awareness of potential problems with extrapolation bias in statistical models, and to provide an additional tool (the gIVH) to help diagnose its presence.
-Other approaches for selecting models to reduce overfitting, such as cross validation~\cite{PicardCook1984}, are also useful for this purpose, but may not entirely eliminate the problem (particularly for sparse datasets).
-In fields such as species distribution modeling and model-based abundance estimation, the goal for analysts is often to build predictive maps of species abundance and/or occurrence using a limited number of sample locations.
-In such applications, the ultimate aim of analysts should be to build models that have low bias and maximal precision.
+In the field of SDM modeling, researchers often stress the need for prediction locations to be similar to the locations used for model development~\cite{ElithLeathwick2009}.
+One way to accomplish this is through a prediction envelope, whereby a specific criterion is used to limit predictions of animal density or occurrence to the range of conditions and covariates encountered during surveys~\cite{MannocciEtAl2015}.
+Using the gIVH for this purpose will likely be more conservative than envelope specifications based on other criterion (e.g., in contrast to minimum an maximum observed covariate values as in~\cite{MannocciEtAl2015}), but is more in line with linear modeling theory.
+A comparison of envelope specification methods is beyond the scope of this paper, but we suspect there are cases where seemingly intuitive envelope strategies result in problematic extrapolations, particularly when the form of prediction models is of high dimension and/or includes multiple interaction terms.
+
+In SDMs and model-based abundance estimation, the goal for analysts is often to build predictive maps of species abundance and/or occurrence using a limited number of sample locations.
+In such applications, the ultimate aim of analysts should be to build models that have low bias and high precision.
 However, traditional approaches to quantifying bias (e.g., goodness-of-fit statistics) only work with observed data points.
@@
 
+Our intent is to raise awareness of potential problems with extrapolation bias in statistical models, and to provide an additional tool (the gIVH) to help diagnose its presence.
+Other methods for selecting models to enhance predictive performance, such as cross validation~\cite{PicardCook1984}, are also useful for this purpose, but may not entirely eliminate the problem (particularly for sparse datasets).
+One approach that might be useful in practice is to combine the cross validation and gIVH paradigms - for instance, using cross validation to narrow down the field to a suite of models with good predictive performance at test locations, and then calculating the gIVH to examine the potential for anomalously high predictions in unsampled locations.
+
 The analyses in this paper focused on abundance estimation, which is necessarily non-negative.
 As such, counts are usually analyzed with a log link function, and there is a much greater potential for positive bias than negative.
+Since prediction variance tends to increase as a function of the magnitude of the prediction, the gIVH will only tend to be able to diagnose predictions that are anomalously large.
 However, one could also apply the gIVH when predicting species occurrence from presence/absence data.
@@
 One area that gIVH ideas may also prove useful is in formulating survey designs.
-The topic of optimal (or near-optimal) spatial design has received considerable attention in the statistical literature, often in the context of designing environmental monitoring systems~\cite{Muller2007}.
+The topic of optimal (or near-optimal) spatial design has received considerable attention in the statistical literature, often in the context of designing environmental monitoring programs~\cite{Muller2007}.
 Optimal designs can be sensitive to the structure of the estimation model that is used, so that tailoring a survey design to a particular model can be somewhat dangerous if there is uncertainty about the ultimate ``best" structure for the model used to relate animal abundance and occurrence to available covariates.
-Nevertheless, one could still think about augmenting a spatially balanced sampling design with a number of locations which are known or thought to have high prediction variance as a function of available covariates.
+Nevertheless, one could still think about augmenting a given sampling design with a number of locations which are known or thought to have high prediction variance as a function of available covariates (e.g., the southwest corner of the study area in the ribbon seal example).
+This could potentially be done as an exercise before any data (or perhaps data from a pilot study) have been collected.
 We are excited about this prospect, and it is a subject of current research.
@@
 \section*{Acknowledgments}
-We thank J. Thorson, J. Ver Hoef, and P. De Valpine for helpful discussion on an earlier version of this mnauscript.  Funding
-for Bering Sea aerial surveys was provided by the U.S. National Oceanic and Atmospheric
-Administration and by the U.S. Bureau of Ocean Energy Management (Interagency
-Agreement M12PG00017).
+We thank J.
+Thorson and J.
+Ver Hoef for helpful discussion on an earlier version of this manuscript.
+David L.
+Miller also provided an insightful review that led us to incorporate frequentist GAMs into the suite of estimation models considered.
 The views and conclusions in this article do not necessarily represent findings or policy of the U.S.
@@
 \begin{figure}[!h]
-\caption{{\bf Depiction of a single simulation scenario}.
+\caption{{\bf Depiction of a single simulation replicate where problematic extrapolation occurs}.
 Panels (A-C) give simulated covariate values, panel D gives true animal abundance, (E) gives estimated abundance from a GLM run on count data from a spatially balanced survey design, and (F) gives abundance from a GLM applied to count data from a convenience survey.
 In (E-F), predictions outside the gIVH are represented by black boxes, and sampling locations are represented with an x.
-For the convenience sample, the median posterior abundance prediction for the entire survey area is 57\% greater than true abundance when inference is made to the whole study area.
-When inference is limited to the gIVH, median posterior abundance was just 16\% greater than true abundance.}
+For the convenience sample, there was considerable positive bias, particularly in cells outside of the gIVH.
+In this case, the median posterior abundance prediction for the entire survey area is 57\% greater than true abundance when inference is made to the whole study area.
+When inference is restricted to cells within the gIVH, median posterior abundance is 16\% greater than true abundance.}
 \label{fig:sim_maps}
@@
 Survey tracks are shown in blue, and are overlayed on a tesselated study area consisting of 25km by 25km grid cells (gray lines).
-Dark gray indicates land, while the brown line indicates a 1000m depth contour, and the orange line shows the U.S Exclusive Economic Zone (EEZ) boundary.
+Dark gray indicates land, while the orange dashed line indicates a 1000m depth contour, and the solid brown line shows the U.S Exclusive Economic Zone (EEZ) boundary.
 Colored pixels indicate ribbon seal counts along aerial transects.
@@
 \begin{figure}[!h]
-\caption{{\bf Posterior mean predictions of ribbon seal apparent abundance across the eastern Bering sea from models fit to survey data}.
-Each row gives result for different model types (GLM, GAM, or STRM, respectively); left column plots give results for naive runs without pseudo-absences, while plots in the right column give predictions for runs where psuedo-absence data (i.e., 0 counts in cells with \textless 0.1\% ice) were included.
-Cells highlighted in black indicate those where predictions were outside the generalized independent variable hull (gIVH) }
+\caption{{\bf Predictions of ribbon seal apparent abundance across the eastern Bering sea from models fit to survey data}.
+Predictions were obtained using the posterior predictive mean for GLM and STRM models, and for the GAM using the \texttt{predict.gam} function in the R \texttt{mgcv} package~\cite{Wood2006}.
+Each row gives result for different model types (GLM, GAM, or STRM, respectively); left column plots give results for naive runs without presumed absences, while plots in the right column give predictions for runs where presumed absence data (i.e., 0 counts in cells with \textless 0.1\% ice) were included.
+Cells highlighted in black indicate those where predictions were outside the generalized independent variable hull (gIVH).
+}
 \label{fig:ribbon_plot}

### 2015-07-30 manuscript/Revision/Text_S1.tex
Commit prior to revision of PLoS One ms. Switched Bayesian GAM out for frequentist one using mgcv, changed color scheme for plot_N_map

--- 2015-04-16 manuscript/Text_S1.tex
@@
 
-We write all statistical models for ecological prediction in the form
+In this paper, we restrict consideration to statistical models for ecological prediction to those that can be written with the form
 \begin{linenomath*}
@@
 \end{linenomath*}
-where $\mu_i=\xi_i + \epsilon_i$, $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function,
-$\xi_i$ is a linear predictor, and $\epsilon_i \sim {\rm Normal}(0,\tau_\epsilon)$ is iid Gaussian error with precision parameter $\tau_\epsilon$.
-The specification in Eq.
-\ref{eq:basic_model} is thus doubly stochastic, in the sense that we assume error associated with $f_Y$ as well as in the location parameter $\mu_i$.
+where $\mu_i=\xi_i + \epsilon_i$, $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function, and
+$\xi_i$ is a linear predictor.
+Many models can be fit with $\epsilon_i = 0$ (such as many frequentist generalized linear and generalized additive models), but the Bayesian models that we develop subsequently involve the specification $\epsilon_i \sim {\rm Normal}(0,\tau_\epsilon)$ where $\tau_\epsilon$ is a precision parameter.
+This specification is doubly stochastic, in the sense that we assume error associated with $f_Y$ as well as in the location parameter $\mu_i$.
 This setup can be useful computationally, and can also be used to approximate singly stochastic system by setting $\tau_\epsilon$ to a large value.
@@
 
-\hspace{.5in}There are a number of ways smooth relationships can be modeled; in order to formulate a GAM in the notation of Eq.
-\ref{eq:basic_model} we employ a knot-based kernel smoother with a Gaussian radial basis function.
-The basic notion is to place $k_j$ knots, which we will denote by $\boldsymbol{\omega}_j = \{ \omega_{1j},\omega_{2j},\hdots,\omega_{k_j j} \}$, throughout the range of a given regressor $j$, where the number of knots controls the level of smoothing.
-For each regressor of interest, we can then calculate a matrix ${\bf K}_j$ of dimension $(n,k_j)$ (where $n$ gives the number of observed data locations), with entries $z_{ik}=\mathcal{N}(X_{ij};\omega_{kj},\tau_j)$, where $\mathcal{N}(x;\mu,\tau)$ denotes a Gaussian distribution with mean $\mu$ and precision $\tau$, $i$ indexes a particular observed data location, and $k$ denotes the $k$th knot for covariate $j$, $k \in \{1,2,\hdots,k_j \}$.
-We can then introduce additional regression coefficients $\boldsymbol{\alpha}_j$ to help model the smooth relationship, and incorporate these into the linear predictor.
-In the case of one smooth term, we then have  in Eq.
+\hspace{.5in}There are a number of ways smooth relationships can be modeled, depending upon the type of basis function employed (e.g.
+cubic smoothing or thin plate splines).
+Regardless of basis choice, it is often possible to write GAMs by substituting the following linear predictor into Eq.
 \ref{eq:basic_model}:
@@
   \label{eq:GAM}
-  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + {\bf K}_j \boldsymbol{\alpha}_j
+  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + {\bf K} \boldsymbol{\alpha} \end{equation}
+\end{linenomath*}
+Here, the component ${\bf X} \boldsymbol{\beta}$ correspond to fixed effects in the usual GLM sense (which may or may not be modeled), ${\bf K}$ represents a smooth basis matrix, and $\boldsymbol{\alpha}$ is a vector of additional parameters.
+Our main point in writing the GAM as in Eq.
+\ref{eq:GAM} is to emphasize the structural similarity between GLMs and certain classes
+of GAMs.  In particular,
+we can rewrite Eq. \ref{eq:GAM} as
+\begin{linenomath*}
+  \begin{equation}
+    \label{eq:GAM-aug}
+    \boldsymbol{\xi}={\bf X}_{aug} \boldsymbol{\beta}_{aug},
   \end{equation}
 \end{linenomath*}
-When fitting such a model, one needs to select the number and location of knots, as well as the precision parameter $\tau_{\omega,j}$ for the Gaussian basis kernels.
-This could be done using some notion of optimality or prediction error (as with cross validation).
-In practice, a simple recommendation is to set $\tau_{\omega,j}$ less than the range of the modeled covariate, and greater than the typical difference between observed values of the covariate.
-
-\hspace{.5in}Our main point in writing the GAM as in Eq. \ref{eq:GAM} is to emphasize the structural similarity between GLMs and certain classes
-of GAMs.  In particular,
-we can rewrite Eq. \ref{eq:GAM} as
-$\boldsymbol{\xi}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
-where ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K}_j \right]$ (i.e., concatenating ${\bf X}$ and ${\bf K}_j$ horizontally), and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha}_j \right] ^\prime$ (the subscript $aug$ denotes augmentation).
-We shall exploit this structure when examining
-extrapolation bias.  We note that similar formulations for GAMs (e.g. using thin plate splines) can be represented in a similar form \citep[cf.][]{CrainiceanuEtAl2005}.
+where ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K} \right]$ (i.e., concatenating ${\bf X}$ and ${\bf K}$ horizontally), and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha} \right] ^\prime$ (the subscript $aug$ denotes augmentation).
 
@@
 However, it is common for residuals from GLM and GAM model fits to include spatial autocorrelation, which violates their common assumption of independently distributed error \citep{Legendre1993,LichsteinEtAl2002}.
-Parameter estimates from GLMs and GAMs that display residual autocorrelation should be interpreted with caution, as they will tend to have overstated precision and may even be biased.
+GLMs and GAMs that display residual autocorrelation should be interpreted with caution, as they will tend to have overstated precision and may even be biased.
 In such situations, analysts often employ spatial regression models which explicitly account for spatial autocorrelation above and beyond that explained by modeled covariates.
@@
 
-We now describe methods for conducting Bayesian analysis for the models described above.
-We utilize Gibbs sampling, characterized by cyclically sampling groups of parameters according to their so-called full conditional distributions conditionally on data and other parameters \citep{GelmanEtAl2004}.
+We now describe methods for conducting Bayesian analysis for the GLM and STRM models described above (note that we used the \texttt{mgcv} R package of \citet{Wood2006} to conduct frequentist analysis of the GAM model, as described in the next section).
+We utilize Gibbs sampling, characterized by cyclically sampling groups of parameters according to their so-called full conditional distributions conditionally on data and other parameters \citep{Gelman2004}.
 In some cases, we are able to solve for full conditional distributions analytically and to sample from known families of probability distributions; in others, we resort to Metropolis-Hastings steps.
@@
 
-\subsection{Calculating the gIVH}
+\section{Frequentist analysis using \texttt{mgcv} package}
 
-In order to calculate prediction variance on the real scale for generalized models, one can use a procedure such as the delta method \citep{Dorfman1938}, or in the case of Bayesian inference, can simply calculate posterior predictive variance of $\boldsymbol{\lambda}_x$.
-For the delta method approach, we can write the variance of the Poisson expectation as
-\begin{linenomath*}
-\begin{eqnarray}
-  \label{eq:var-lambda}
-  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
-  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
-%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
-%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
-%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
-%      \vdots & \vdots & \ddots & \vdots \\
-%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
-%  \end{array} \right].
-\end{eqnarray}
-\end{linenomath*}
-where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).
-Under common univariate link functions (e.g.
-log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g.
-multinomial logit) $\boldsymbol{\Delta}$ will be dense.
-
-\hspace{.5in} Unlike Cook's original formulation, the gIVH will often involve unknown parameters.
-Further, the exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the underlying model structure and estimation procedure.
-We impose a probability distribution $[\boldsymbol{\theta}]$ for unknown parameters, and compute prediction variance as
-\begin{linenomath*}
-\begin{equation}
-  \label{eq:E_var_lambda}
-  {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}) = \int_{\boldsymbol{\theta}} {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}, \boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
-\end{equation}
-\end{linenomath*}
-
-\hspace{.5in} Our approach in this paper is to conduct a Bayesian analysis (using the Gibbs sampling algorithms outlined above) and to use posterior predictive inference to summarize ${\rm var}(\boldsymbol{\lambda}_x | {\bf Y})$.
-Although we use a Bayesian formulation for modeling convenience, one could also consider substituting parametric approximations for ${\rm Var}(\mu_x)$ and $\mu_x$ based on output from standard likelihood-based analyses.
+In this section, we describe analysis using the \texttt{mgcv} R package \citep{Wood2006}, and how the gIVH can be calculated conditioned on a fitted GAM model.
+In particular, we calculate that gIVH using prediction variance with the following steps:
+\begin{enumerate}
+  \item Fit a GAM model using the \texttt{gam} function.
+For this  presentation, we will assume this object is given the name ``gam.fit."  If the proportion of each sample unit (e.g., grid cell) that is surveyed is less than 1.0, an offset should be included in the formula object required by the \texttt{gam} function.
+For instance, a count model might specify a formula object such as \texttt{count \textasciitilde offset(log(Offset)) + s(my.cov,5)}.
+In this case \texttt{Offset} is a vector holding the proportion of each sample unit that is surveyed.
+  \item Extract the augmented design matrix ${\bf X}_{aug}$ from Eq.
+\ref{eq:GAM-aug} for observed data using the \texttt{predict.gam} function from \texttt{mgcv}.
+\citet{Wood2006} calls this the ``prediction matrix," and one can obtain it using the \texttt{type="lpmatrix"} option in \texttt{predict.gam}.
+  \item Calculate variance on the linear predictor scale as $V_\mu = {\bf X}_{aug} \hat{\textrm{var}}(\hat{\boldsymbol{\beta}}_{aug}){\bf X}_{aug}^\prime$, where $\hat{\textrm{var}}(\hat{\boldsymbol{\beta}}_{aug})$ is the variance-covariance matrix for estimated parameters (obtainable using \texttt{vcov(gam.fit)}).
+  \item Use the delta method \citep{Dorfman1938} to calculate the variance-covariance matrix on the response scale (call this $V_{obs}$).
+Set $v_{max}$ equal to the maximum diagonal element of $V_{obs}$.
+  \item Assemble a new \texttt{data.frame} (call this \texttt{Pred}), which includes covariate values for all sample units (grid cells) one wants to make inference to.
+  \item Perform steps 2-3 from above using this new dataset (using \texttt{newdata=Pred} when calling predict.gam).
+  \item Use the delta method \citep{Dorfman1938} to calculate the variance-covariance on the response scale for all prediction locations (call this $V_{pred}$).
+  \item All sample units $i$ for which $V_{pred}[i,i] \le v_{max}$ are included in the gIVH.
+\end{enumerate}
 

### 2015-07-30 manuscript/Revision/Text_S2.tex
Commit prior to revision of PLoS One ms. Switched Bayesian GAM out for frequentist one using mgcv, changed color scheme for plot_N_map

--- 2015-04-16 manuscript/Text_S2.tex
@@
   \item Simulate animal abundance across the landscape as a function of environmental covariates,
-  \item Simulate animal population surveys across the landscape, including the position of count quadrats and the resulting animal counts,
+  \item Simulate surveys across the landscape, including the position of count quadrats and the resulting animal counts,
   \item Estimate animal abundance as a function of two of the environmental covariates according three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).
 For the latter, only a spatial dimension was modeled.
-All such models were specified hierarchically, with MCMC used for posterior simulation.
+The GLM and STRM models were specified hierarchically, with MCMC used for posterior simulation.
+For the GAM, we conducted a frequentist analysis using the \texttt{mgcv} R package \citep{Wood2006}.
   \item Calculate the gIVH using realized posterior variance (i.e. after data were collected and analyzed).
@@
 
-For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models, corresponding to a generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a generalized additive model (GAM; smooth effects of covariates on the log scale), and a spatio-temporal regression model (with both linear, fixed effects and spatially autocorrelated random effects).
+For each set of count data, we attempted to estimate animal abundance over the landscape using three different models, corresponding to a hierarchical, Bayesian generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a frequentist generalized additive model (GAM; smooth effects of covariates on the log scale) fit using \texttt{mgcv}, and a hierarchical spatial regression model (STRM; with both linear, fixed effects and spatially autocorrelated random effects).
 Further details on these models, together with the procedure used for posterior simulation, are presented in Text S1.
@@
 Regression parameters ($\beta$) were given vague $\mathcal{N}(0,\tau=0.01)$ priors.
+For the GLM and STRM, we used the median posterior prediction of total abundance as a point estimator.
+For the GAM, we used the sum of predictions of total abundance on the response scale, obtained using the \texttt{predict.gam} function in \texttt{mgcv}.
 
@@
 
-As suggested in Text S1, we used Eq. S1.7 to calculate the gIVH, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.
+As suggested in Text S1, we used Eq.
+S1.7 to calculate the gIVH for the GLM and STRM models, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.
+For the GAM, we used the delta method \citep{Dorfman1938} to calculate prediction variance on the response scale, conditioning on estimated $\beta$ values.
 

### 2015-07-30 SpatPred/R/sim_data_abundance.R

@@ -3,6 +3,7 @@
 #' @param n.covs Number of habitat covariates (default 4)
 #' @param tau.epsilon Precision of exchangeable errors in log space
 #' @return A SpatialPolygonsDataFrame holding covariates, abundance, and lambda (expected abundance)
+#' @import mvtnorm RandomFields spatstat gridExtra
 #' @export
 #' @keywords spatio-temporal, simulation
 #' @author Paul B. Conn
@@ -73,7 +74,6 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
     Cov.plot=ggplot()+geom_density(data=Cor.df,aes(x=Cor))+xlab("Correlation coefficient")+ylab("Density")+
       theme(plot.title = element_text(hjust = 0),axis.text=element_text(size=rel(1.5)),title=element_text(size=rel(1.5)))+ggtitle("D.")
     pdf(file="Covs_coregion.pdf")
-    library(gridExtra)
     grid.arrange(arrangeGrob(Grid.plot1,Grid.plot2,Grid.plot3,Cov.plot,nrow=2))
     dev.off()
   }
@@ -128,6 +128,7 @@ sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
 #' @param prop.sampled A numeric value giving the proportion of each survey unit that is actually sampled (default 0.1)
 #' @param n.points Number of survey units that are sampled.  The default is to round (0.1*S)
 #' @return A data.frame with two columns, "Mapping" - indicating which survey units are sampled, and "Counts" - giving no. of animals counted
+#' @import spsurvey grid
 #' @export
 #' @keywords spatio-temporal, simulation
 #' @author Paul B. Conn

### 2015-07-30 SpatPred/R/util_funcs.R

@@ -120,12 +120,15 @@ d_biv_normal<-function(Tmp.vec,XY,Sigma){
 #' @param leg.title Title for legend of plot (if different from the default "Abundance")
 #' @param hcolor If highight provided, gives color for highlighting (default = "yellow")
 #' @param cur.max If provided, sets maximum value for plotted abundance
+#' @param rainbow If TRUE, use rainbow plot; otherwise use color blind safe palette
 #' @return An abundance map
+#' @import sp ggplot2 RColorBrewer rgeos
 #' @export 
 #' @keywords abundance map, plot
 #' @author Paul Conn \email{paul.conn@@noaa.gov}
-plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance",hcolor='yellow',cur.max=NULL){
-  myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))  
+plot_N_map<-function(cur.t,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance",hcolor='yellow',cur.max=NULL,rainbow=F){
+  if(rainbow)myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral"))) 
+  if(!rainbow)myPalette <- colorRampPalette(rev(brewer.pal(9, "YlOrRd")))
   Tmp=Grid[[1]]
   if(is.null(highlight)==FALSE){
     midpoints=data.frame(gCentroid(Tmp[highlight,],byid=TRUE))

### 2015-07-30 SpatPred/inst/analyze_sim_data.R

@@ -2,17 +2,13 @@
 ### note: this has some hardwiring for local directory structures
 S=900
 n.sims=1000
-Bias=matrix(0,n.sims,24)
-colnames(Bias)=c("GLM.sys.all","GAM.sys.all","RSR.sys.all","RSR2.sys.all",
-                 "GLM.sys.gIVH","GAM.sys.gIVH","RSR.sys.gIVH","RSR2.sys.gIVH",
-                 "GLM.sys.gIVH2","GAM.sys.gIVH2","RSR.sys.gIVH2","RSR2.sys.gIVH2",                 
-                 "GLM.clust.all","GAM.clust.all","RSR.clust.all","RSR2.clust.all",
-                 "GLM.clust.gIVH","GAM.clust.gIVH","RSR.clust.gIVH","RSR2.clust.gIVH",
-                 "GLM.clust.gIVH2","GAM.clust.gIVH2","RSR.clust.gIVH2","RSR2.clust.gIVH2")
-Ncell.gIVH=matrix(0,n.sims,8)
-colnames(Ncell.gIVH)=c("GLM.sys","GAM.sys","RSR.sys","RSR2.sys","GLM.clust","GAM.clust","RSR.clust","RSR2.clust")
-Ncell.gIVH2=matrix(0,n.sims,8)
-colnames(Ncell.gIVH2)=c("GLM.sys","GAM.sys","RSR.sys","RSR2.sys","GLM.clust","GAM.clust","RSR.clust","RSR2.clust")
+Bias=matrix(0,n.sims,20)
+colnames(Bias)=c("GLM.sys.all","GAM.sys.all","GAM2.sys.all","RSR.sys.all","RSR2.sys.all",
+                 "GLM.sys.gIVH","GAM.sys.gIVH","GAM2.sys.gIVH","RSR.sys.gIVH","RSR2.sys.gIVH",
+                 "GLM.clust.all","GAM.clust.all","GAM2.clust.all","RSR.clust.all","RSR2.clust.all",
+                 "GLM.clust.gIVH","GAM.clust.gIVH","GAM2.clust.gIVH","RSR.clust.gIVH","RSR2.clust.gIVH")
+Ncell.gIVH=matrix(0,n.sims,10)
[…]

### 2015-07-30 SpatPred/inst/compile_plot_ribbon_data.R

@@ -65,7 +65,7 @@ new.colnames[1:2]=c("Easting","Northing")
 colnames(tmp2)=new.colnames
 tmp.theme=theme(axis.ticks = element_blank(), axis.text = element_blank(),plot.title = element_text(hjust = 0))
 
-tiff(file="covariates_col.tiff",height=8,width=8,units='in',res=300)
+tiff(file="covariates_col.tiff",height=8,width=8,units='in',res=300,compression='lzw')
 pushViewport(viewport(layout=grid.layout(2,2)))
 p1=ggplot(tmp2)+aes(Easting,Northing,fill=dist_mainland)+geom_raster()+tmp.theme+scale_fill_gradientn(colours=greenPalette(100),name="Value")+ggtitle("A. Distance from mainland")
 print(p1,vp=viewport(layout.pos.row=1,layout.pos.col=1))

### 2015-07-30 SpatPred/inst/plot_IVH_simple.R

@@ -146,7 +146,7 @@ pdf(file="IVH_simple2.pdf")
 grid.arrange(arrangeGrob(linear.plot,quad.plot,plot.biv,plot.biv.2,nrow=2,widths=unit.c(unit(.5,"npc"))))
 dev.off()
 
-tiff(file="IVH_simple2.tiff",res=300,width=8,height=8,units='in')
+tiff(file="IVH_simple2.tiff",res=300,width=8,height=8,units='in',compression='lzw')
 grid.arrange(arrangeGrob(linear.plot,quad.plot,plot.biv,plot.biv.2,nrow=2,widths=unit.c(unit(.5,"npc"))))
 dev.off()
 

### 2015-07-30 SpatPred/inst/plot_sim_example.R

@@ -3,6 +3,7 @@ library(sp)
 library(rgeos)
 library(ggplot2)
 library(gridExtra)
+library(RColorBrewer)
 source("./spatpred/r/util_funcs.R")
 isim=51
 S=900
@@ -33,6 +34,11 @@ Cov1 = plot_N_map(1,matrix(Grid@data$cov1,S,1),Grid=Grid.list,leg.title="Value")
 Cov2 = plot_N_map(1,matrix(Grid@data$cov2,S,1),Grid=Grid.list,leg.title="Value")+ggtitle("B. Covariate 2")+theme(plot.title = element_text(hjust = 0,size = rel(4)),legend.text=element_text(size=rel(3)),legend.title=element_text(size=rel(3)),text=element_text(size=rel(3)))+xlab('')+ylab('')
 Cov3 = plot_N_map(1,matrix(Grid@data$cov3,S,1),Grid=Grid.list,leg.title="Value")+ggtitle("C. Covariate 3")+theme(plot.title = element_text(hjust = 0,size = rel(4)),legend.text=element_text(size=rel(3)),legend.title=element_text(size=rel(3)),text=element_text(size=rel(3)))+xlab('')+ylab('')
 
+cov.palette = colorRampPalette(rev(brewer.pal(9, "Blues")))
+Cov1=Cov1+scale_fill_gradientn(colours=cov.palette(100),name="Value")
+Cov2=Cov2+scale_fill_gradientn(colours=cov.palette(100),name="Value")
+Cov3=Cov3+scale_fill_gradientn(colours=cov.palette(100),name="Value")
+
 cur.max=max(c(Grid@data$N,apply(MCMC.sys$GLM$MCMC$Pred,1,'median'),apply(MCMC.clust$GLM$MCMC$Pred,1,'median')))
 
 N.plot=plot_N_map(1,matrix(Grid@data$N,S,1),Grid=Grid.list,leg.title="N",cur.max=cur.max)+ggtitle("D. True Abundance")+theme(plot.title = element_text(hjust = 0,size = rel(4)),legend.text=element_text(size=rel(3)),legend.title=element_text(size=rel(3)),text=element_text(size=rel(3)))+xlab('')+ylab('')
@@ -54,7 +60,8 @@ Est.clust=plot_N_map(1,N,Grid=Grid.list,leg.title="N",highlight=which(MCMC.clust
[…]

### 2015-07-30 SpatPred/inst/run_mgcv_sim_analysis.R

@@ -1,0 +1,109 @@
+# run_mgcv_sim_analysis.R
+# script to run mgcv analysis on generic spatio-temporal count data simulations
+# requires data already simulated with "run_generic_sims.R"
+# Some of the code in this script was provided to us by D. L. Miller
+library(mgcv)
+
+mainDir=getwd()
+subDir="Sim_data"
+dir.create(file.path(mainDir, subDir), showWarnings = FALSE)
+set.seed(123456)
+n.sims=1000 #number of simulations at each design point
+S=900
+prop.sampled=0.1
+n.transects=45
+
+dtmfn <- function(eta){sapply(eta, grad, f=tmfn)}
+
+#call estimation routines
+for(isim in 1:n.sims){ 
+  for(igen in 1:2){  #loop over generating model to generate data sets
+    Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
+    load(Cur.file) #load abundance,covariate grid
+    
+    if(igen==1){
+      Cur.file=paste("./Sim_data/Effort_sys",isim,".Rda",sep='')
+      load(Cur.file) #load Effort 
+    }
+    if(igen==2){
+      Cur.file=paste("./Sim_data/Effort_clust",isim,".Rda",sep='')
+      load(Cur.file) #load Effort 
+    }   
+    
+    #format dataset suitable for mgcv
+    Sim.data=Grid@data
+    Sim.data$Offset=Sim.data$count=NA
+    Sim.data$Offset[Effort$Mapping]=prop.sampled
+    Sim.data$count[Effort$Mapping]=Effort$Counts
+    
+    Area.adjust=rep(1,S)
+    
+    # run mgcv quasipoisson and tweedie GAMs
+    k1=5
+    b <- gam(count~offset(log(Offset))+s(cov1,k=k1)+s(cov2,k=k1),
+             data=Sim.data, family=quasipoisson(), method="REML",control=gam.control(scale.est="robust"))
+    b.tw <- gam(count ~offset(log(Offset))+s(cov1,k=k1)+s(cov2,k=k1),
+             data=Sim.data, family=tw(), method="REML")
+
+    Pred.df=Sim.data
+    Pred.df$Offset=1
+    
+    ###prediction stuff for quasipoisson
+    
+    lpmat.obs <- predict(b, type="lpmatrix")
+    tmfn <- b$family$linkinv
+     
[…]

### 2015-07-30 SpatPred/inst/run_ribbon_models.R

@@ -45,39 +45,6 @@ Grid.list[[1]]=Cur.grid
 cell.width=25067
 naive.glm.plot=plot_N_map(1,matrix(apply(glm.out.cov$MCMC$Pred,1,'mean'),S,1),Grid=Grid.list,leg.title="Abundance",highlight=which(glm.out.cov$gIVH==0),cell.width=25067,hcolor='black')
 
-#gam model
-formula=~1
-Data=Cur.grid
-spat.mod=0
-Offset=Area.photo
-Area.adjust=Area.hab
-Assoc=NULL
-Knots=NULL
-Prior.pars=NULL
-Names.gam=c("ice_conc","dist_mainland","dist_edge","dist_shelf")
-n.par=length(Names.gam)
-Knots.gam=vector("list",n.par) 
-Kern.gam.sd=rep(0,n.par)
-for(ipar in 1:n.par){
-  cur.min=min(Data[[Names.gam[ipar]]])
-  cur.max=max(Data[[Names.gam[ipar]]])
-  Range=cur.max-cur.min
-  incr=0.333*Range
-  Knots.gam[[ipar]]=c(cur.min-2*incr,cur.min-incr,cur.min,cur.min+incr,cur.min+2*incr,cur.max,cur.max+incr,cur.max+2*incr)
-  Kern.gam.sd[ipar]=incr
-}
-Control=list(iter=10100,burnin=100,thin=2,adapt=TRUE,fix.tau.epsilon=FALSE,srr.tol=NULL,predict=TRUE,Kern.gam.sd=Kern.gam.sd,MH.nu=rep(0.4,length(Mapping)))
-gam.out <- spat_pred(formula=formula,Data=Data,spat.mod=spat.mod,Offset=Offset,Effort=Effort,Area.adjust=Area.adjust,Control=Control,Assoc=Assoc,Names.gam=Names.gam,Knots.gam=Knots.gam,Prior.pars=Prior.pars)
-Control=gam.out$Control
-Control$adapt=FALSE
-Control$iter=60000
-Control$burnin=10000
-Control$thin=10
-gam.out <- spat_pred(formula=formula,Data=Data,spat.mod=spat.mod,Offset=Offset,Effort=Effort,Area.adjust=Area.adjust,Control=Control,Assoc=Assoc,Names.gam=Names.gam,Knots.gam=Knots.gam,Prior.pars=Prior.pars)
-
[…]

### 2015-07-30 make_ggplot_Alaska_Josh.R

@@ -1,0 +1,91 @@
+library(ggplot2)
+library(nPacMaps)
+library(sp)
+library(rgeos)
+library(rgdal)
+
+load('plotfiles/2012_Tracks_for_ST_analysis.Rdata')
+load('plotfiles/Data_for_ST_plot.Rdata')
+
+#reconstitute knots
+Coords=coordinates(Data$Grid[[1]])
+x.min=min(Coords[,1])-100000
+x.max=max(Coords[,1])+100000
+y.min=min(Coords[,2])-100000
+y.max=max(Coords[,2])+100000
+
+X=x.min+(x.max-x.min)/6*c(0:6)
+Y=y.min+(y.max-y.min)/6*c(6:0)
+XY=expand.grid(x=X,y=Y)
+
+Knots=SpatialPoints(coords=XY,proj4string=CRS(proj4string(Data$Grid[[1]])))
+
+Distances=gDistance(Knots,Data$Grid[[1]],byid=TRUE)
+Distances=apply(Distances,2,'min')
+my.buffer=150000
+Which.include=which(Distances<my.buffer)
+Knots=Knots[Which.include,]
+
+# coerce Knots to data frame for ggplot
+Knots <- as.data.frame(Knots)
+
+# use the nPackMaps ggExpansion function to 
+# expand the plot extent by 0.1 in x and y
+lims <- nPacMaps::ggExpansion(Knots,x="x","y",x_fac=0.1,y_fac=0.1)
+xlim <- lims$xlim
+ylim <- lims$ylim
+
+# fortifi is a modification of ggplot's fortify that
+# allows us to simplify complex polygon objects like
+# the alaska and russia dcw
+ak<-nPacMaps::fortifi(alaska_dcw,tol=2,minarea=1000)
+rus<-nPacMaps::fortifi(russia_dcw,tol=2,minarea=1000)
+
+# fortify Data$Grid[[1]] and Tracks for ggplot
+boss_grid <- fortify(Data$Grid[[1]])
+trks <- fortify(Tracks)
+
+# fortify requires a SpatialLinesDF, so we'll fake it
+shelf <- SpatialLinesDataFrame(gBoundary(Shelf_break[1,]),data.frame(id=1))
+eez <- SpatialLinesDataFrame(gBoundary(EEZ_Alaska),data.frame(id=1))
+
+# fortify shelf and eez
+shelf <- fortify(shelf)
+eez <- fortify(eez)
+
[…]
